# The Research Compendium {#sec-compendium}

## Learning objectives

By the end of this chapter the reader should be able
to:

- Explain what a research compendium is, and trace its
  lineage from Gentleman and Temple Lang through
  `rrtools`.
- Justify why the R-package structure is the ideal
  compendium container, and describe how `R CMD check`
  thereby becomes a universal integrity gate.
- Enumerate the Five Pillars of a compendium and name
  the single artifact that captures each.
- Distinguish the five compendium archetypes and state,
  for each, how the directory layout, the presence of a
  rendered report, and the applicable validation gates
  differ while the reproducibility machinery stays the
  same.
- Apply the principle of progressive disclosure to let a
  project evolve from one archetype to another without
  structural reorganisation.

## Orientation

Consider a first-year epidemiology student who has spent
three months on a cohort analysis: an incident-diabetes
study drawn from a registry, with a cleaning script, a
handful of models, four figures, and a draft of a
results section. The work exists, but it exists as a
scatter. The cleaning script sits in a downloads folder,
the models in a second file that reads a spreadsheet
whose path is hard-coded to one laptop, the figures as
loose `png` files pasted by hand into a manuscript, and
the record of which package versions produced any of it
nowhere at all. Nothing here is wrong as statistics. The
problem is that the analysis is not an object. It cannot
be handed to a colleague, cited, archived, or re-run,
because there is no single thing to hand over.

The remedy is to make the analysis one object. A
*research compendium* is that object: a single,
self-describing directory that bundles the code, the
data, the prose, and the record of the environment in
which the analysis was run, organised by convention so
that another person, or the same person a year later,
can find every part and regenerate the result. The
compendium is the third of the recurring ideas of this
book, after the levels ladder and the
capture-versus-validate distinction, and it is the one
that gives the other two something to act on. A
reproducibility level is a property *of* a compendium; a
validation gate is a check run *against* a compendium.

The compendium also sets the starting rung on the
ladder. Merely by being a version-controlled directory
whose parts can be located, a compendium reaches L0,
locatable. It is not yet L1, because nothing has pinned
the package versions; it is not yet L2, because nothing
has pinned the environment; and it is certainly not yet
L3, because nothing has re-run it and shown the numbers
to match. What the compendium provides is the container
into which the later chapters install those higher
guarantees. We build the box first, and the chapters
that follow fill it.

## The analyst's contribution

Three judgements at the centre of this chapter are the
analyst's alone. No tool scaffolds them, and getting
them wrong produces a tidy directory that is
nonetheless the wrong shape for the work.

1. **Naming the archetype.** The analyst must decide
   what kind of object the project is: a data analysis,
   a manuscript, a reusable package, a simulation study,
   or a blog. This choice governs the layout, whether a
   report is rendered, and which gates apply. It is a
   statement of intent, not a technical setting, and it
   is the organising decision of the chapter.

2. **Fixing the raw-and-derived boundary.** The analyst
   must decide what is an immutable input and what is a
   product of the analysis. A file placed in
   `raw_data/` is a promise that it will never be
   modified; a file in `derived_data/` is a promise that
   it can always be regenerated from the raw inputs and
   the code. No program can draw this line, because it
   depends on where the analyst judges the science to
   begin.

3. **Deciding what becomes a function.** The analyst
   must decide which code is a throwaway pipeline step,
   belonging in `analysis/scripts/`, and which is a
   reusable procedure worth documenting and testing,
   belonging in `R/`. This judgement determines what
   `R CMD check` and the test suite will actually
   protect, and it is revisited continuously as a
   project matures.

::: {.callout-note}
## The vocabulary of this chapter

- **Research compendium**: a single self-describing
  directory bundling the code, data, prose, and
  environment record of an analysis, organised by
  convention so that it can be located and re-run.
- **R-package skin**: the minimal set of files
  (`DESCRIPTION`, `NAMESPACE`, an `R/` directory) that
  makes a directory a valid R package, and therefore
  checkable by the R build tools.
- **`R CMD check`**: the standard R command that builds
  a package, runs its tests, and applies several hundred
  correctness checks; here it serves as a universal
  integrity gate on the whole compendium.
- **The Five Pillars**: the five artifacts a compendium
  manages, one per capture point: the `Dockerfile`,
  `renv.lock`, `.Rprofile`, the source code, and the
  data.
- **Archetype**: the kind of research object a
  compendium is (data analysis, manuscript, package,
  simulation study, or blog), which varies the layout
  and the gates but not the underlying machinery.
- **Render gate**: a validation check that a report or
  document rebuilds cleanly from source; it applies to
  archetypes that carry a report and not to those that
  do not.
- **Raw versus derived data**: immutable original
  inputs versus the processed datasets regenerated from
  them by code.
- **Progressive disclosure**: the property that one
  layout supports a project's evolution from one
  archetype to another with no reorganisation.
:::

## The compendium as the unit of reproducible work

The idea is not new, and its history is worth a
paragraph because it explains the shape of the modern
object. Gentleman and Temple Lang, writing in 2007,
proposed the compendium in its current sense: a document
that combines text, code, and data in a form a reader
can actually execute, rather than a paper that merely
describes an analysis performed elsewhere
[@gentleman2007]. The proposal drew on Knuth's earlier
notion of literate programming, in which prose and code
are woven into one source [@knuth1984]. For some years
the compendium remained an aspiration without a
conventional implementation, and Peng, arguing that
reproducibility should be treated as a minimum standard
for computational science, sharpened the motivation
without settling the container [@peng2011].

The decisive move came from Marwick, Boettiger, and
Mullen, who observed that a container of exactly the
required shape already existed and was in daily use by
thousands of R developers: the R package
[@marwick2018]. An R package, as standardised by Wickham
[@wickham2015], already provides a `DESCRIPTION` file
recording metadata and dependencies, an `R/` directory
for reusable functions, a `tests/` directory for unit
tests, a `vignettes/` directory for long-form prose, and
a rigorously specified build system. Marwick and
colleagues extended this layout with a single top-level
`analysis/` directory holding subdirectories for raw
data, derived data, scripts, figures, and the manuscript
itself, and packaged the whole as the `rrtools` tool.
The resulting compendium can be installed, tested, and
checked with the very same commands used for any R
package. Nust, Boettiger, and Marwick later wrote a
companion guide on how to *read* such a compendium, which
is a useful sign that the object had matured enough to
have a reading protocol at all [@nust2018].

## The R package as the compendium container

The choice of the R package as the container is the
single most consequential design decision in this
lineage, and it repays careful statement. It appears at
first to be a convenience, reuse of a familiar layout,
but its real value is that it makes an existing,
rigorous, and universal gate apply to the whole
compendium for free.

The gate is `R CMD check`. When R was designed as a
system for distributing statistical software, its
authors built a command that takes a package and subjects
it to several hundred checks: that the code parses and
loads, that every documented function matches its
documentation, that declared dependencies are actually
available, that the examples run, and that the tests
pass. The command was built to keep the tens of thousands
of packages on CRAN mutually consistent. By dressing a
compendium in the R-package skin, one inherits this
entire apparatus at no cost. A compendium that passes
`R CMD check` is one whose code loads, whose stated
dependencies resolve, and whose tests succeed, in a clean
process that carries none of the author's accumulated and
undeclared state.

This is why the R-package structure is not incidental to
the compendium but constitutive of it. Every archetype we
shall meet below, however different in outward shape,
carries the same minimal R-package skin, and therefore
`R CMD check` is the one gate that applies to all of
them. A manuscript compendium adds a render gate on top;
a simulation compendium adds a seeding discipline; a
package compendium leans on the check most heavily of
all; but the universal floor is the same. We return to
this point when we tabulate the archetypes, because it is
the thread that unifies them.

::: {.callout-note icon=false collapse=true}
## Check your understanding: why an R package?

A colleague proposes to organise a compendium as a plain
folder with sensible subdirectory names, arguing that the
R-package files (`DESCRIPTION`, `NAMESPACE`) are empty
ceremony for an analysis that is not a package anyone will
install. What is lost by this choice?

The plain folder can still be locatable and can still
hold all five artifacts, so it can in principle reach the
higher levels of the ladder. What is lost is the universal
gate. Without the R-package skin, `R CMD check` has
nothing to check, so there is no single, standard,
pre-built command that verifies the code loads, the
dependencies resolve, and the tests pass. The analyst
must then assemble an equivalent set of checks by hand for
every project. The ceremony is precisely the interface
that makes an existing verification apparatus apply.
:::

## The Five Pillars

A compendium is defined by convention, but it *manages* a
specific and small set of artifacts. The `zzcollab`
framework organises these as the Five Pillars: the five
things a compendium reads, writes, and version-controls,
one for each point at which reproducibility can be
captured [@marwick2018]. An analysis is reproducible only
to the extent that all five are captured, and the pillars
are layered rather than independent, each resting on the
ones beneath it.

1. **Computational environment**, captured by the
   `Dockerfile`. This pillar fixes the operating system,
   the R version itself, the system libraries, and the
   document toolchain, all the software beneath the R
   packages. It is the subject of @sec-containers.

2. **Package versions**, captured by `renv.lock`. This
   pillar records the exact version of every R package the
   analysis loads, so that the same set can be restored on
   another machine. It is the subject of
   @sec-dependencies.

3. **Session configuration**, captured by the project
   `.Rprofile`. This pillar fixes the R options, the
   random-number-generator kind, the locale and timezone,
   and the thread counts that shape a session before any
   user code runs.

4. **Source code**, captured by the version-control
   history. This pillar holds the scripts, the functions,
   the manuscript source, and the tests that implement the
   analysis.

5. **Data**, captured either by inclusion in the
   compendium, for small redistributable files, or by
   reference to an external repository with a resolvable
   identifier, for large or restricted data.

Two of the pillars are not coordinate with the other
three. Source code and data, pillars four and five, are
the inputs whose re-execution is the goal; the
environment, packages, and session, pillars one through
three, are the conditions under which that re-execution
occurs. This chapter is about the container that holds all
five and the shape that container takes. The individual
pillars are developed in the chapters named against them.

::: {.callout-important}
## The honest level

Assembling the five artifacts into a tidy compendium is a
capture activity, and it is easy to mistake the tidiness
for a reproducibility guarantee. A compendium that
contains a `Dockerfile`, a `renv.lock`, and a clean
`analysis/` tree looks reproducible, but until the image
has been rebuilt from the recipe, the lockfile restored,
and the outputs regenerated and compared, the compendium
sits at L0 or L1, not at L2 or L3. Match the claim to the
rung actually reached. A well-organised directory is a
precondition for reproducibility, not evidence of it.
:::

## The directory layout, concretely

It helps to see the whole layout at once. The following
is the top-level structure of a `zzcollab` compendium.
The block is illustrative and is not executed.

```
penguin-analysis/
├── DESCRIPTION           # package metadata and dependencies
├── NAMESPACE             # generated; exported functions
├── LICENSE
├── R/                    # reusable, documented functions
├── man/                  # generated function documentation
├── tests/                # tinytest runner
├── inst/tinytest/        # unit tests (test-*.R)
├── vignettes/            # long-form prose
├── analysis/
│   ├── data/
│   │   ├── raw_data/       # immutable inputs
│   │   ├── derived_data/   # regenerable script outputs
│   │   └── metadata/       # data dictionaries, provenance
│   ├── scripts/           # numbered pipeline steps
│   ├── report/            # report.Rmd / index.qmd manuscript
│   ├── figures/           # generated figures
│   └── tables/            # generated tables
├── Dockerfile            # Pillar 1: environment
├── renv.lock             # Pillar 2: package versions
├── .Rprofile             # Pillar 3: session configuration
├── Makefile              # workflow entry points
├── zzcollab.yaml         # configuration: base image, archetype
├── .zzcollab-state       # generated state record; never hand-edited
└── .github/workflows/    # continuous-integration definitions
```

The reader will note the division. Above the `analysis/`
directory sits the standard R-package machinery: this is
the skin that makes `R CMD check` apply. Within
`analysis/` sits the research proper, in the Marwick
convention, with the raw-and-derived data separation that
is the analyst's second judgement made physical. At the
root sit the three environment artifacts, one per pillar,
alongside the `Makefile` that provides the commands and
the `zzcollab.yaml` that records, among a small set of
parameters, the archetype. That last file is our bridge
to the centrepiece of the chapter, because the archetype
is a first-class parameter recorded there, elicited at
creation time alongside the choice of base image.

## The five archetypes

We come now to the heart of the matter, and to the point
this book returns to in every chapter: a research
compendium is not only a manuscript. It is tempting, once
one has seen the `analysis/report/` directory, to imagine
that the purpose of a compendium is always to render a
paper. This is the common case, but it is one case among
five, and treating it as the whole is the error the
present section exists to prevent.

`zzcollab` treats the research archetype as a first-class
initialisation parameter, orthogonal to the choice of
computational environment [@marwick2018]. Where the base
image fixes the environment, the archetype determines the
starting scaffolding: the directory emphasis, whether a
primary report exists, and which validation gates apply on
top of the universal one. Five archetypes are advertised,
and because every one of them carries the minimal
R-package skin, `R CMD check` remains the universal gate
across all five, as the framework's own whitepaper stresses
in its account of the archetype axis. What varies is not
the reproducibility machinery, which is identical, but the
shape of the work and the gates it warrants. We take each
in turn, with a health-sciences example, and then tabulate
the differences.

### 1. Data analysis (a cohort study)

The first archetype is the exploratory and confirmatory
data analysis with no manuscript as its primary organising
document. This is the closest to the original `rrtools`
use case, and it is the archetype of the epidemiology
student of our orientation. The work lives in
`analysis/scripts/`, as numbered pipeline steps, and in
`analysis/figures/`, with the derived datasets in
`analysis/data/derived_data/`. Consider an
incident-diabetes cohort study: a cleaning script that
reads the registry extract from `raw_data/` and writes an
analysis-ready table to `derived_data/`, a modelling
script that fits a Cox proportional-hazards model and
writes the coefficient table to `tables/`, and a plotting
script that writes the Kaplan-Meier curves to `figures/`.
There may be a short results document, but it is not the
point; the point is the pipeline and its outputs.

The applicable gates are the universal `R CMD check` and
the unit tests, the latter guarding the cleaning logic and
the data invariants (that the cohort has the expected
number of subjects, that no derived rate is negative). A
render gate is optional here, because there may be no
report to render. The reproducibility machinery,
environment and packages and session, is nonetheless
present in full.

### 2. Scholarly manuscript (a comorbidity paper)

The second archetype makes the manuscript the organising
principle. A `report.Rmd` or `index.qmd` file in
`analysis/report/` is created on day one and written in
parallel with the analysis, so that the prose and the
numbers never drift apart, and every figure in the paper is
regenerated on each render rather than pasted in by hand. A
representative deployment is a study of the comorbidity of
PTSD, depression, and diabetes among Black women drawn from
a trauma-study cohort, using variable selection, multiple
imputation, and causal mediation analysis. The compendium
organises the manuscript in `analysis/report/report.Rmd`,
the statistical functions in `R/` with `roxygen2`
documentation, the analysis pipeline in
`analysis/scripts/`, and the imputed datasets in
`analysis/data/derived_data/`.

Here the render gate is central, not optional: a
validation run rebuilds the report from source, and a
failure to render is a failure of the compendium. The gates
are therefore `R CMD check`, the unit tests, and the render
of `report.Rmd` to PDF. This is the archetype most readers
imagine when they hear the word compendium, and the burden
of the present section is that it is only the second of
five.

### 3. R package (a trial-design function)

The third archetype arises when the reusable code matures
into something of general interest, at which point `R/`,
`man/`, and `tests/` become the first-class content and the
compendium is shareable as a CRAN-style package. The
distinguishing feature, and the reason this archetype
matters for the argument, is that *there is no report to
render at all*. A package that computes the required sample
size for a group-sequential clinical trial, or the
operating characteristics of a Bayesian dose-finding
design, has no manuscript as its deliverable; its
deliverable is the exported, documented, tested function. A
representative deployment is a package for generating
publication-quality clinical-trial summary tables, hosting
its source in `R/`, its examples in `vignettes/`, and its
documentation as a generated website.

The gates for this archetype are `R CMD check`, run at its
full strength, and the unit tests. There is no render gate,
because there is nothing to render. This case is the
clearest refutation of the manuscript-only view: an
entirely valid, fully reproducible compendium can produce
no report whatsoever, and its integrity rests on the
universal check and the tests alone.

### 4. Simulation study (a power analysis)

The fourth archetype is the simulation study: a power
calculation, or an assessment of the operating
characteristics of a proposed trial design, in which the
compendium enumerates a grid of parameter configurations,
runs Monte Carlo replicates at each, and stores the
results. The layout leans on `analysis/scripts/` for the
simulation driver and on `analysis/data/derived_data/` for
the stored replicate results, which are treated as derived
data because they can be regenerated from the code and the
seeds.

The distinguishing discipline of this archetype is
*seeding*. A simulation whose random draws are not pinned
by an explicit `set.seed()` is not reproducible even in
principle, because a second run produces different numbers
by construction. The framework's simulation scaffolding
therefore seeds a parallel-ready starter script, and a
recommended practice is to store a seed column alongside
the parameter grid so that each replicate is individually
reproducible. Consider a power simulation for a
longitudinal trial: the driver builds a grid over sample
size and effect size, assigns each row an integer seed,
simulates a dataset for each row, fits the planned model,
and records whether the null was rejected; the empirical
power is then the rejection rate within each cell. The
gates are `R CMD check`, the unit tests, and, as an
emphasis particular to this archetype, a test that the
seeded simulation returns identical output on a second
call. A render gate is optional, since the deliverable is
often the stored results rather than a paper.

### 5. Blog (a teaching notebook)

The fifth archetype is the blog or lab notebook, in which
the compendium renders not one report but many documents. A
teaching blog carrying a series of methods posts, or a
running lab notebook of analyses, is a single compendium
whose `analysis/` tree holds many `index.qmd` files, each
rendered to a static site by the same machinery that renders
a manuscript. Integration with Quarto [@allaire2024]
enables publication to the site from the same compendium
that produced the underlying analyses, so that the full
five-pillar reproducibility stack applies to web-published
teaching content without any change to the workflow.

The gate here is again the universal `R CMD check`, joined
by a render gate that must succeed for *each* post rather
than for a single report. The multiplicity is the
distinguishing feature: a blog compendium is a
render-many, where the manuscript is a render-one and the
package is a render-none. The machinery beneath all three
is, once more, identical.

The five archetypes and their differences are summarised
in @tbl-archetypes, and @fig-gates shows which validation
gates apply to each. The reader should read both across the
rows, to see how the layout and the gates vary, and down
the two universal columns, to see how much stays fixed.

```{r}
#| label: tbl-archetypes
#| tbl-cap: 'The five research-compendium archetypes. The reproducibility machinery is identical across all five; only the directory emphasis, the presence of a rendered report, and the gates that apply on top of the universal R CMD check differ.'

library(tibble)
library(knitr)

archetypes <- tibble::tibble(
  archetype = c(
    'Data analysis', 'Scholarly manuscript', 'R package',
    'Simulation study', 'Blog'
  ),
  directory = c(
    'analysis/scripts/, analysis/figures/',
    'analysis/report/',
    'R/, man/, tests/',
    'analysis/scripts/, derived_data/',
    'many analysis/ index.qmd posts'
  ),
  report = c(
    'Optional', 'Yes, to PDF', 'No',
    'Optional', 'Yes, many'
  ),
  gate = c(
    'R CMD check and tests',
    'R CMD check plus render',
    'R CMD check and unit tests',
    'R CMD check plus seed determinism',
    'R CMD check plus render of each post'
  ),
  example = c(
    'Incident-diabetes cohort analysis',
    'PTSD and diabetes comorbidity paper',
    'Group-sequential trial-design function',
    'Power and operating-characteristics simulation',
    'Teaching blog or lab notebook'
  )
)

knitr::kable(
  archetypes,
  col.names = c(
    'Archetype', 'Directory emphasis', 'Renders a report?',
    'Primary validation gate', 'Health-sciences example'
  )
)
```

```{r}
#| label: fig-gates
#| fig-cap: 'Validation gates by archetype. The R-package check and the unit tests apply universally; the render and seeding gates apply selectively. The underlying reproducibility machinery, the Five Pillars, is identical across every row.'

library(ggplot2)

archs <- c(
  'Data analysis', 'Scholarly manuscript', 'R package',
  'Simulation study', 'Blog'
)
gates <- c(
  'R CMD check', 'Unit tests', 'Render report',
  'Seed determinism'
)
grid <- expand.grid(
  archetype = archs, gate = gates,
  stringsAsFactors = FALSE
)

renders <- c('Data analysis', 'Scholarly manuscript', 'Blog')
seeded <- c('Simulation study')

grid$status <- 'not emphasised'
grid$status[grid$gate %in% c('R CMD check', 'Unit tests')] <-
  'universal'
grid$status[grid$gate == 'Render report' &
              grid$archetype %in% renders] <- 'applies'
grid$status[grid$gate == 'Seed determinism' &
              grid$archetype %in% seeded] <- 'applies'

grid$archetype <- factor(grid$archetype, levels = rev(archs))
grid$gate <- factor(grid$gate, levels = gates)
grid$status <- factor(
  grid$status,
  levels = c('universal', 'applies', 'not emphasised')
)

ggplot(grid, aes(x = gate, y = archetype, fill = status)) +
  geom_tile(colour = 'white', linewidth = 1.2) +
  scale_fill_manual(
    values = c(
      'universal' = '#2c7fb8',
      'applies' = '#7fcdbb',
      'not emphasised' = '#edf2f4'
    )
  ) +
  labs(x = NULL, y = NULL, fill = NULL) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid = element_blank(),
    legend.position = 'bottom'
  )
```

::: {.callout-note icon=false collapse=true}
## Check your understanding: the render-none case

A reviewer objects that a compendium producing no rendered
document cannot be a reproducible research object, since
there is no result to reproduce. Which archetype answers
this objection, and how?

The R-package archetype answers it. A package that exports
a trial-design function has no report to render, yet it is
fully reproducible: its result is the behaviour of the
function, and that behaviour is protected by `R CMD check`
and by the unit tests, which regenerate and verify the
function's outputs on every check. The absence of a render
gate does not weaken reproducibility; it simply reflects
that the deliverable is a tested procedure rather than a
document. Reproducibility is a property of re-execution,
not of rendering.
:::

## Progressive disclosure

The five archetypes are not walls between separate kinds
of project. They are, rather, stages a single project may
pass through, and the framework's most useful structural
property is that the same layout supports the whole
trajectory without reorganisation. This is what the
framework calls *progressive disclosure*: the same
directory layout, the same `Makefile`, and the same Five
Pillars carry a project from one archetype to the next, so
that no transition demands that the analyst tear the
project apart and rebuild it in a new shape.

The trajectory is easy to picture, and it is the one our
epidemiology student is likely to walk. The project begins
as a data analysis: scripts and figures, no manuscript. As
the findings firm up, a `report.Rmd` is added to
`analysis/report/` and the project becomes a manuscript
compendium, with the render gate now active, but nothing
already written needs to move. In the course of the
analysis a cleaning routine and a modelling helper prove
useful enough to document and test, so they migrate from
`analysis/scripts/` into `R/`, and the package archetype's
gates begin to bite on them without any change to the
layout. Should one of those functions reach general
interest, the compendium can be shared as a package, its
`R/` and `man/` and `tests/` now first-class, while the
`analysis/` directory recedes to a development notebook.
The same compendium has been, in turn, a data analysis, a
manuscript, and a package, and at no point was it
reorganised.

The lesson we draw, and it is the organising lesson of the
chapter, is that the archetype is a description of a
project's current emphasis rather than a permanent
category. The reproducibility machinery is invariant across
the archetypes and across the transitions between them, and
it is precisely this invariance that lets a project evolve
without ever leaving the reproducible state it started in.

## Worked example

To make the progressive-disclosure trajectory concrete, we
follow a single compendium through its first two
archetypes, staying at the level of the directory and the
gates rather than the analysis. Suppose our student begins
the incident-diabetes study. The compendium is created and
its archetype recorded as a data analysis. The shell
session below is illustrative and is not executed.

```bash
$ mkdir diabetes-cohort && cd diabetes-cohort
$ zzc analysis        # scaffold; archetype recorded in zzcollab.yaml
```

The student places the registry extract under `raw_data/`,
writes `analysis/scripts/01_clean.R` to produce a derived
cohort table, and writes `analysis/scripts/02_model.R` to
fit the Cox model. A unit test in `inst/tinytest/` asserts
the cohort size and that no fitted hazard ratio is
negative. At this point the only gates that apply are the
universal check and the tests, run in a clean environment:

```bash
$ make check-renv     # dependencies declared and locked
$ make docker-test    # R CMD check plus tinytest
```

Some weeks later the results are firm enough to write up.
The student adds `analysis/report/report.Rmd`, moving the
figure code out of `02_model.R` and into report chunks that
regenerate the figures on render. The compendium is now a
manuscript, and the archetype line in `zzcollab.yaml` is
updated to reflect it, but not a single existing file has
moved. A render gate now joins the others:

```bash
$ make docker-render  # rebuild report.Rmd to PDF
```

The cleaning routine, meanwhile, has grown into a general
helper worth documenting, so it migrates from
`analysis/scripts/01_clean.R` into `R/prepare_cohort.R`,
gains a `roxygen2` block, and picks up its own unit tests.
The package archetype's discipline now protects it, and
`R CMD check` verifies that the documentation matches the
function signature. The same compendium is now doing the
work of three archetypes at once, on one unchanged layout,
with each gate applying to the part of the project that
warrants it. This is progressive disclosure in operation,
and it is the payoff of having chosen a single, principled
container at the outset.

## Collaborating with an LLM

A language model is a capable assistant for scaffolding and
organising a compendium, and an unreliable one for the
judgements that define it. The three triples below mark the
division.

**Prompt.** 'I have a folder of R scripts and CSV files for
a cohort analysis. Propose a research-compendium directory
structure and tell me which files go where.'

*Watch for*: a plausible layout that silently guesses the
raw-and-derived boundary, placing a file the model cannot
distinguish as input or output into the wrong directory,
and an omission of the R-package skin (`DESCRIPTION`,
`NAMESPACE`) that quietly forfeits the `R CMD check` gate.

*Verification*: confirm by hand that every file in
`raw_data/` is a genuine immutable input and every file in
`derived_data/` is regenerable from code, because this is
the analyst's judgement and not the model's. Confirm the
R-package skin is present, and run `R CMD check` to
establish that the proposed structure is in fact checkable.

**Prompt.** 'Which archetype should this project be, and
which validation gates apply?'

*Watch for*: a default to the manuscript archetype, since
that is the most common case in the model's training data,
even when the project has no report and is really a package
or a simulation study; and a confident enumeration of gates
that includes a render gate for a project that renders
nothing.

*Verification*: decide the archetype yourself from the
project's actual deliverable, a document, a function, or a
set of stored simulation results, and then check that the
gates the model lists match @tbl-archetypes for that
archetype. A render gate on a render-none package is the
tell that the model has guessed the archetype wrongly.

**Prompt.** 'Write the unit tests that should gate this
compendium.'

*Watch for*: tests that assert only that functions exist or
that a script runs without error, which are structural
checks rather than correctness tests, and, for a simulation
compendium, an absence of any test that the seeded output
is reproducible.

*Verification*: read each generated test and confirm it
checks a value the analysis actually depends on, not merely
that code executes. For a simulation study, add or verify a
test that a fixed seed yields identical output on a second
call, since that is the property the archetype exists to
protect and the one a model is most likely to omit.

## Exercises

1. Take an existing analysis of your own, currently a
   scatter of scripts and data files, and classify it as
   one of the five archetypes. State the deliverable that
   determines the classification, and list the validation
   gates that should apply on top of `R CMD check`.

2. For that same analysis, draw the raw-and-derived
   boundary explicitly: list every data file and assign it
   to `raw_data/` or `derived_data/`, justifying each
   assignment by whether the file is an immutable input or a
   regenerable product. Identify any file whose status is
   genuinely ambiguous and explain the ambiguity.

3. Explain, in a paragraph, why `R CMD check` is described
   in this chapter as a universal gate. What does dressing a
   compendium in the R-package skin buy that a plain,
   well-named folder does not?

4. Sketch a progressive-disclosure trajectory for a project
   that begins as a simulation study and ends as an R
   package. At each transition, state which gate becomes
   active or inactive, and confirm that no directory needs
   to be reorganised.

5. Consider a simulation compendium whose driver script
   contains no `set.seed()` call. At which rung of the
   reproducibility ladder can such a compendium sit at best,
   and why can it not, even in principle, reach L3? Write
   the one-line test you would add to gate the seeding.

6. The chapter claims that the manuscript is only one of
   five archetypes. Construct the strongest argument you can
   for the opposing view, that every serious research
   compendium should ultimately produce a rendered document,
   and then state where the argument fails.

## Further reading

The foundational proposal is Gentleman and Temple Lang's
account of the compendium as an executable document
combining text, code, and data [@gentleman2007]; it reads
well alongside Knuth's original statement of literate
programming, from which the idea descends [@knuth1984]. The
operationalisation that this chapter builds on is Marwick,
Boettiger, and Mullen's proposal that an R package is the
natural compendium container, together with their `rrtools`
tool [@marwick2018]; the reader who wants the R-package
mechanics in full should consult Wickham's book on R
packages, which is the standard reference for the skin that
makes `R CMD check` apply [@wickham2015]. For the
complementary skill of *reading* a compendium one has
received rather than writing one, Nust, Boettiger, and
Marwick offer a short and practical guide [@nust2018].
Finally, Peng's argument that reproducibility is a minimum
standard for computational science supplies the motivation
that the compendium is built to serve [@peng2011], and the
individual pillars named in this chapter are developed in
@sec-dependencies, @sec-containers, and the validation
chapters that follow.
