# The Reproducibility Crisis {#sec-crisis}

::: {.bookepigraph}
```{=latex}
\begin{flushright}\small
```
*More than 70% of researchers have tried and failed to
reproduce another scientist's experiments, and more than
half have failed to reproduce their own experiments.*

[Monya Baker]{.smallcaps}, *1,500 Scientists Lift the Lid
on Reproducibility* (2016)
```{=latex}
\end{flushright}
```
:::

## Learning objectives

By the end of this chapter the reader should be able
to:

- Summarize the empirical evidence that a large
  fraction of published health-sciences research
  cannot be regenerated, and distinguish the
  studies that measure replication from those that
  measure computational re-execution.
- State the distinction the National Academies draw
  between reproducibility and replicability, note that
  the wider literature adds robustness as a third term,
  and explain why the terminology is contested.
- Explain why many re-execution failures are
  computational rather than statistical, using the
  silent-default changes in R as concrete cases.
- Describe, in first outline, the stack of
  determinants beneath a computational result, and
  locate where in that stack a published script
  leaves the outcome unpinned.
- Argue why reproducibility is a floor rather than a
  ceiling, and why a reproducible analysis can
  nonetheless be wrong.

## Orientation

Consider a situation that every quantitative health
scientist eventually meets. A biostatistician on a
clinical-trial team runs the primary analysis,
obtains the treatment effect and its confidence
interval, exports the table that will anchor the
regulatory submission, and files the script. Two
years later, during an audit, an inspector asks that
the analysis be regenerated from the archived code
and data. On a fresh machine the script errors on
the first `library()` call. Once the missing package
is located and installed, the analysis runs. The
point estimate has moved in the third decimal place,
however, and one confidence interval no longer excludes
the null. Nothing in the statistical reasoning was wrong.
What failed was everything around the statistics: a
package that changed beneath the code, a default that
shifted between releases of R, a system library
present on the first machine and absent on the
second.

This book exists to prevent that failure, and this
chapter establishes that the failure is real, common,
and specifically computational. The stakes in a
health science are not abstract. A result that cannot
be regenerated cannot be audited. An analysis that a
regulator cannot reproduce is an analysis a regulator
may decline to trust. And a clinical recommendation
built on a number that no one can recompute is a
recommendation resting on faith rather than evidence.
Reproducibility is, in the end, a precondition for
the trust that clinical and public-health decisions
require.

Throughout the book we place an analysis on a
four-rung ladder, and we shall begin to use its
vocabulary at once. An analysis is L0, locatable,
when it is under version control and its source can
be found, but nothing computational is pinned. It is
L1, pinned packages, when the exact version of every
package is fixed by a lockfile. It is L2, pinned
environment, when the operating system, the language
runtime, and the system libraries are fixed as well,
usually inside a container. It is L3, verified, when
the recorded computational outputs have been
regenerated and shown to match. The auditor's failure
above is the failure of an analysis that reached, at
best, L0: the code could be found, and nothing else
about the computation had been fixed. The ladder is
developed in full in @sec-levels; here we use it only
to name what was missing.

## The analyst's contribution

Three judgments at the center of this chapter belong
to the analyst and are automated by no tool.

1. **Naming the failure mode.** When a result will
   not regenerate, or regenerates differently, the
   analyst must decide whether the discrepancy is
   statistical (the method was inappropriate, the
   inference was misdrawn) or computational (the code
   is sound but its environment has shifted). The
   remedy differs entirely by case, and the diagnosis
   is a human act of judgment, not a mechanical one.

2. **Matching the claim to the level reached.** A
   reproducibility claim is a scientific claim and
   must be calibrated to evidence. The analyst decides,
   and states honestly, which rung of the ladder an
   analysis has actually reached, and resists the
   temptation to assert a level higher than the one
   secured.

3. **Deciding what must be pinned.** Not every
   determinant matters equally for every analysis. A
   simulation study lives or dies by its random-number
   configuration; a geospatial pipeline depends on
   system libraries that a survey analysis never
   touches. Judging which layers of the stack are
   load-bearing for the analysis at hand is a decision
   the analyst makes before any tool is invoked.

::: {.callout-note}
## The vocabulary of this chapter

New to these terms? Each is defined again on first
use.

- **Computational reproducibility.** Obtaining the
  same numerical results from the same data and the
  same code, on another machine or at a later time.
- **Replicability.** Obtaining consistent findings
  from a new study, with new data, aimed at the same
  question.
- **Robustness.** Reaching the same qualitative
  conclusion from the same data under different but
  defensible analytical choices.
- **Determinant stack.** The layered set of things a
  computational result depends on, from the hardware
  and operating system, through the system and
  numerical libraries, the language runtime, and the
  package versions, up to the session configuration
  and the code itself.
- **Silent default.** A behavior that software
  applies unless told otherwise, and that can change
  between software versions without any error or
  warning to the analyst.
- **Seed and RNG kind.** The seed (`set.seed`) is the
  fixed starting point that makes a random draw
  repeatable; the RNG kind is the generator algorithm
  that turns that seed into numbers, and its default
  has changed across R versions.
- **Package churn (code rot).** The gradual breakage
  of once-working code as the packages beneath it are
  updated over time.
- **Research compendium.** The self-describing bundle
  of code, data, and text that travels as the unit of
  a reproducible analysis, developed in
  @sec-compendium.
:::

## The evidence that re-execution fails

The case for taking reproducibility seriously is
empirical, and it is not encouraging. We begin with
the large replication studies, because they set the
scale of the problem, and then turn to the narrower
and, for our purposes, more damning studies that
isolate the computational failures.

In psychology, the Open Science Collaboration
re-ran one hundred published studies and could
replicate the original finding in only thirty-nine of
them [@opensciencecollaboration2015]. In preclinical
oncology, Begley and Ellis, working from within
industry, reported that only six of fifty-three
landmark findings could be reproduced, a success rate
close to one in ten [@begley2012]. These figures were
not, at first, universally believed.
But the suspicion that the published literature was less
solid than it appeared already had a theoretical footing.
Ioannidis had argued, given the prevailing rates of
low statistical power, analytical flexibility, and
publication bias, that most published research findings
are false [@ioannidis2005]. When Baker later surveyed
more than fifteen hundred researchers for Nature, a
majority agreed that science was facing a
reproducibility crisis. More than seventy percent
reported that they had tried and failed to
reproduce another scientist's experiments
[@baker2016].

These are, for the most part, failures of
replication, at the scientific end of a spectrum we
shall shortly define, and their causes are many. But
a growing body of work isolates the specifically
computational failures. Here the question is not
whether a finding holds up in a new sample, but merely
whether the original analysis can be made to run a
second time. Here the evidence is starker, because the
failures are more avoidable. Trisovic and colleagues
re-ran more than nine thousand R files deposited
alongside published articles in the Harvard Dataverse.
Roughly one in four executed without
error, before any question of whether the output
matched the published result [@trisovic2022]. Saju and
colleagues examined R code supplements archived on
the Open Science Framework. They found that only about a
quarter of the projects ran to completion, and that
the great majority, close to ninety-nine percent,
carried no formal description of their dependencies at
all [@saju2025].

The distinction between the two bodies of evidence
matters. The replication studies tell us that
scientific findings are fragile. The re-execution
studies tell us something more specific and more
remediable, namely that a large fraction of published
analyses cannot be regenerated even in principle. The
reason is that the computational scaffolding needed to
run them was never captured. It is the second failure,
and not the first, that the tooling in this book
addresses.

::: {.callout-note icon=false collapse=true}
## Check your understanding: two kinds of failure

A colleague re-runs the archived code from a
published cohort study, obtains a number that differs
from the printed one, and concludes that the paper's
finding does not replicate. Why is that conclusion
premature, and which of the studies cited above speaks
most directly to the colleague's situation?

::: {.callout-note icon=false collapse=true}
## Answer

The colleague has demonstrated, at most, a failure of
computational reproducibility: the same code and data
did not regenerate the same number, most likely
because the software environment shifted between the
original run and this one. That is a statement about
re-execution, not about whether the scientific
finding holds in a new sample, which is what
replication means. The situation is precisely the one
that Trisovic and colleagues [@trisovic2022] measured
at scale. Until the environment is reconstructed and
the code is shown either to regenerate the original
number or to contain a genuine error, no claim about
replication is warranted.
:::
:::

## Three words that are not synonyms

Before proceeding we must fix our terms, because the
vocabulary of this field is genuinely contested and a
reader who does not fix it will be confused by the
literature. Goodman and colleagues surveyed the
usage and found that the same words carry opposite
meanings in different research communities, so that
one field's 'reproducibility' is another field's
'replication' [@goodman2016]. We adopt the
distinctions of the National Academies, whose 2019
consensus report offers the settlement most widely
followed in the health sciences [@nas2019].

*Reproducibility* is obtaining consistent
computational results using the same input data, the
same computational steps, methods, and code, and the
same conditions of analysis. It is a statement about
re-execution, and it is the subject of this book.
*Replicability* is obtaining consistent results across
studies aimed at the same question, each collecting
its own data. It is a statement about the durability
of a scientific finding. To these two the report and
the wider literature add a third, *robustness*,
obtaining consistent conclusions from the same data
when the analytical choices are varied within the
range a competent analyst would defend. The three are
independent. An analysis can be perfectly
reproducible and entirely wrong, faithfully
regenerating the same mistaken number on every
machine. A finding can replicate across laboratories
while no single analysis in the set is computationally
reproducible. And a result can be reproducible yet
fragile to a defensible change in model specification.

The three terms sort cleanly once one notices that they
answer the same question about two different things: what
was held fixed, the data or the analysis. @fig-three-words
lays the possibilities out as a grid, a presentation the
Turing Way has made the common one [@turingway2022]. The
fourth cell, in which both the data and the analysis
change, is generalizability, which this book names here for
completeness and does not treat.

```{r}
#| label: fig-three-words
#| fig-cap: 'Reproducibility, replicability, and robustness distinguished by what is held fixed. The columns vary the data and the rows vary the analysis, so each cell names the claim supported when a consistent result survives that particular change. Only the shaded cell, in which both the data and the analysis are held fixed, is the subject of this book. The grid presentation follows the Turing Way [@turingway2022]; the definitions are those of the National Academies [@nas2019] adopted above.'
#| fig-alt: 'A two-by-two grid. The columns are same data and new data; the rows are same analysis and varied analysis. The cell where the data and the analysis are both the same is shaded and labeled reproducibility, re-execution, which is described as the subject of this book. Same data with a varied analysis is labeled robustness. New data with the same analysis is labeled replicability. New data with a varied analysis is labeled generalizability and marked as not treated in this book.'
#| fig-width: 6.4
#| fig-height: 3.6

library(ggplot2)
library(tibble)

grid <- tibble(
  data     = rep(c('Same data', 'New data'), each = 2),
  analysis = rep(c('Same analysis', 'Varied analysis'), 2),
  term = c('Reproducibility',
           'Robustness',
           'Replicability',
           'Generalizability'),
  gloss = c('the same code re-run',
            'defensible analytic\nchoices varied',
            'a new study of the\nsame question',
            'not treated in\nthis book'),
  focus = c(TRUE, FALSE, FALSE, FALSE)
)

grid$data <- factor(grid$data, levels = c('Same data', 'New data'))
grid$analysis <- factor(grid$analysis,
                        levels = c('Varied analysis', 'Same analysis'))

ggplot(grid, aes(x = data, y = analysis)) +
  geom_tile(aes(fill = focus), color = 'white', linewidth = 2) +
  geom_text(aes(label = term), vjust = -0.6,
            fontface = 'bold', size = 3.9) +
  geom_text(aes(label = gloss), vjust = 1.25, size = 3,
            lineheight = 0.95, color = 'grey25') +
  scale_fill_manual(values = c('TRUE' = '#cfe6e3',
                               'FALSE' = '#f2f2f2'),
                    guide = 'none') +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 10) +
  theme(
    panel.grid = element_blank(),
    axis.text = element_text(face = 'bold', size = 9.5)
  )
```

The idea that computational work should be
reproducible in the narrow, re-execution sense is
older than the current crisis. It is usually traced
to Claerbout and his colleagues in geophysics. In the
early 1990s they built their published papers so
that every figure could be regenerated from the
archived code and data by a single command
[@claerbout1992]. Their standard was demanding. The
phrasing that has become the field's motto is Buckheit
and Donoho's paraphrase of it [@buckheit1995;
@claerbout1992]: the published article is merely the
advertisement of the scholarship, and the scholarship
itself is the full computational apparatus that produced
the figures.

Reproducibility so understood is not a single
threshold but a spectrum. Peng arranged computational
work along a scale, from the merely published, through
code available, code and data available, and linked
and executable, to fully replicable. He argued that
reproducibility is a minimum standard for judging a
scientific claim when full independent replication is
impractical [@peng2011; @peng2006]. The lesson we take
from Peng, and develop into the levels ladder of
@sec-levels, is that reproducibility comes in degrees.
Honesty about which degree one has reached is
itself a scientific virtue.

## Why the failures are computational, not statistical

It is worth dwelling on why re-execution fails,
because the reasons point directly at the remedy.
Consider two changes to R itself, each silent, each an
improvement, and each capable of altering a published
result without any error or warning to announce it.

In R 3.6.0, released in 2019, the default algorithm
that turns a seed into a random sample was corrected.
The old sampler introduced a slight bias in
`sample()`; the new one removed it. The consequence
for reproducibility is direct. Code that set a seed and
drew a sample obtained one sequence of draws before
the release and a different sequence after it, with
identical code and an identical seed [@rcoreteam2019].
Every bootstrap, every cross-validation fold, every
permutation test, and every simulation written under
the old default draws differently under the new one.
Then, in R 4.0.0, released in 2020, the
`stringsAsFactors` default flipped from `TRUE` to
`FALSE`. Reading a data frame now silently produced
character columns where it once produced factors. This
changed the behavior of any code that depended on
factor-specific behavior, such as the ordering of
levels or a function that requires a factor
[@rcoreteam2020]. Neither
change was a bug; both were deliberate improvements.
Both would nonetheless break the exact reproduction of
an analysis written under the old default, and neither
would say a word.

To these language-level shifts we must add the
steadier erosion of package churn, the phenomenon by
which working code rots as the packages beneath it are
updated. A modern analysis may depend on dozens or
hundreds of packages, each with its own release history
and its own breaking changes. It is this churn that
Trisovic and colleagues measured when they found that
most deposited R scripts ceased to run within a few
years [@trisovic2022]. The code did not change; its
dependencies did.

These are instances of a general phenomenon that we
shall develop fully in @sec-levels. A computational
result depends not only on the code and the data but
on a stack of determinants beneath them: the operating
system and its system libraries, the numerical
libraries that implement linear algebra, the language
runtime, the exact versions of every package, the
session configuration (the random-number kind, the
contrasts, the number of printed digits, the locale
and time zone), and, underneath all of it, the
hardware. A published script pins almost none of
these. It records the analysis logic and, if the
reader is fortunate, the data, and leaves the
remaining layers to whatever the next machine happens
to provide. Reproducibility fails in the gap between
what the script fixes and what the result depends on.
The organizing question of this book is how far
down that stack a given effort has actually reached.

This diagnosis, that the failures are computational
rather than statistical, is what makes the problem
tractable. A statistical error requires a scientific
judgment to correct; a computational failure requires
only that the determinant stack be captured, layer by
layer, so that the same computation can be run again.
The crisis is not, at bottom, a shortage of
statistical competence. It is a shortage of
computational hygiene, and hygiene can be taught,
tooled, and enforced.

The point applies to every kind of computational
output, not only to the polished manuscript. A
research compendium comes in several archetypes,
developed in full in @sec-compendium: a data-analysis
project, a scholarly manuscript, an R package, a
simulation study, and a blog of many rendered posts.
Each varies the directory layout, whether a rendered
report exists at all, and which validation gates
apply. But all share the same determinant stack, and
so are exposed to the same silent shifts. An R package
with no report to render can still be broken by a
changed `sample.kind`. A simulation study, whose whole
output is a set of random draws, is more exposed to
the RNG change than any manuscript. And a teaching
blog that renders forty posts inherits the churn of
every package any post loads. The discipline that
follows applies across all of them.

::: {.callout-note icon=false collapse=true}
## Check your understanding: locating the failure

An analyst reports that a script producing a
principal-components plot gives visibly different
loadings on a colleague's laptop, although both
installed the same package versions from the same
lockfile. Setting a seed changes nothing. Which layer
of the determinant stack is the most likely culprit,
and why does the lockfile not protect against it?

::: {.callout-note icon=false collapse=true}
## Answer

The most likely culprit is the numerical-library
layer, specifically the BLAS and LAPACK
implementation that performs the eigendecomposition
underlying `prcomp()`. Different implementations
(reference BLAS, OpenBLAS, Intel MKL, Apple's
Accelerate) return results that are numerically
equivalent but not bit-identical, and a sign flip or
reordering of components can make the difference
visible in a plot. A lockfile pins the R packages but
says nothing about the system libraries the packages
link against. Those live below the package layer, at
L2 rather than L1, and are fixed only by controlling
the environment itself, the subject of @sec-containers.
Setting a seed does not help because the difference is
not stochastic.
:::
:::

## A reproducible analysis can still be wrong

We must guard against a misreading that the foregoing
invites. Nothing in this book claims that a
reproducible analysis is a correct one. The point is
pressed by Leek and Peng, who observe that
reproducibility guarantees only that a computation can
be regenerated, not that the computation was the right
one to perform [@leek2015]. A script that
faithfully regenerates a mistaken p-value on every
machine, forever, is perfectly reproducible and
perfectly wrong. Reproducibility disciplines the
computation; it does not discipline the science.

It follows that reproducibility is a floor and not a
ceiling. It is the minimum standard beneath which a
computational claim cannot be assessed at all, because
a result no one can regenerate is a result no one can
check. Above that floor sit all the harder questions,
of study design, of confounding, of model choice, of
inference, that reproducibility does not touch. We
insist on the floor precisely so that the harder
questions become answerable: an analysis that can be
re-run is an analysis whose statistical choices can be
inspected, criticized, and, where necessary, corrected.

::: {.callout-important}
## The honest level

A reproducibility claim is a scientific claim and must
be matched to the level actually reached, never
asserted above it. To say that an analysis is
reproducible when the code merely exists in a
repository (L0) is to claim L3 while having secured
L0 only. The gap will be discovered at exactly the
moment, an audit, a revision, a replication attempt,
when it is most costly. State the rung you have
reached, and no higher.
:::

## Worked example

To make the RNG change concrete rather than anecdotal,
we reproduce it here in a few lines of base R. The
generator that turns a seed into a sample is
selectable through `RNGkind()`, and R still ships both
the pre-3.6.0 sampler, named `Rounding`, and the
current one, named `Rejection`. We can therefore hold
the seed fixed at a single value, vary only the
sampler, and observe directly that the same seed does
not give the same draws when the RNG kind differs. The
function below sets the sampler, sets the seed, and
shuffles the integers one through ten. We call it under
each sampler and restore the session's original
generator afterward, so that nothing downstream is
disturbed.

```{r}
#| label: tbl-rng-kind
#| tbl-cap: 'The same seed yields the same draws only under the same RNG kind. Each row shows the first six values of a shuffled sequence of the integers one through ten, drawn under seed 123.'

draw_sample <- function(sample_kind) {
  suppressWarnings({
    RNGkind(sample.kind = sample_kind)
    set.seed(123)
    sample(1:10)
  })
}

original_rng <- RNGkind()

rounding_draw  <- draw_sample('Rounding')
rejection_draw <- draw_sample('Rejection')

suppressWarnings(RNGkind(
  kind        = original_rng[1],
  normal.kind = original_rng[2],
  sample.kind = original_rng[3]
))

comparison <- data.frame(
  sampler = c('Rounding (R < 3.6.0)',
              'Rejection (R >= 3.6.0)'),
  draws = c(
    paste(head(rounding_draw),  collapse = ', '),
    paste(head(rejection_draw), collapse = ', ')
  )
)

knitr::kable(
  comparison,
  col.names = c('RNG sample kind', 'First six draws'),
  align = 'l'
)
```

The two rows of @tbl-rng-kind share a seed and share a
line of code, `set.seed(123); sample(1:10)`, and yet
they diverge almost immediately, coinciding only
sporadically thereafter. An analyst who set a
seed under R 3.5, confident that the seed made the
analysis reproducible, would find the draws silently
changed on upgrading to R 3.6 or later. The seed was
never the whole story; the seed together with the RNG
kind is. This is the determinant stack in miniature:
the code fixed one layer, the seed, and left another,
the sampler, to whatever the running version of R
supplied. Pinning the R version, which we reach at L2
with a container, is what closes this particular gap.
Recording the RNG kind in the session
configuration, meanwhile, is what makes the gap
visible in the interim.

## Collaborating with an LLM

A large language model is a capable assistant for the
material of this chapter, provided the analyst treats
its output as a draft to be verified rather than an
authority to be trusted. Three uses recur, and each
carries a specific verification burden.

**Prompt.** 'Explain the difference between
reproducibility, replicability, and robustness, and
give a health-sciences example of each.'

*Watch for.* Models frequently conflate the first two
terms, or adopt the reversed convention that some
fields use, because both usages are well represented
in their training data. The example offered may be
plausible-sounding but mislabelled.

*Verification.* Check the definitions against the
National Academies framing [@nas2019] adopted in this
chapter, and confirm that the example labeled
'reproducibility' concerns re-executing the same code
on the same data, not re-collecting data. Do not
accept the term assignment without this check.

**Prompt.** 'My results changed after I upgraded R.
What could have caused this and how do I diagnose it?'

*Watch for.* The model may propose confident but
untested causes, or invent version numbers and release
dates for the change it blames. It has no access to
your session and cannot know which layer actually
moved.

*Verification.* Treat every proposed cause as a
hypothesis to test against your own `sessionInfo()`
and the actual release notes, not as a diagnosis.
Confirm the specific mechanism (for the RNG change, by
the demonstration above; for `stringsAsFactors`, by
inspecting a column's class) before acting on it.

**Prompt.** 'Write a paragraph for my methods section
stating that my analysis is fully reproducible.'

*Watch for.* The model will happily assert a level of
reproducibility that your project has not reached,
because it is completing a pattern, not auditing your
compendium. A confident claim of full reproducibility
is exactly the overstatement the honest-level caution
warns against.

*Verification.* Establish which rung of the ladder the
analysis has genuinely reached, and edit the paragraph
to claim that rung and no higher. The claim is your
responsibility, not the model's.

## Exercises

1. Locate a published article in your own field that
   provides code and data. Attempt to run the code on
   your machine without modification. Record what
   happened at each step, and classify any failure you
   meet as computational or statistical. On which rung
   of the ladder (L0 to L3) would you place the
   deposited analysis, and why?

2. The chapter distinguishes reproducibility,
   replicability, and robustness. For a study you know
   well, write one sentence describing what it would
   mean for that study to fail on each of the three
   dimensions separately. Which failure would be most
   damaging to the study's clinical conclusion?

3. Modify the worked-example chunk to draw a bootstrap
   resample of a small synthetic vector rather than a
   shuffle, again comparing the two RNG kinds under a
   fixed seed. Does the choice of sampler still change
   the result? Explain what this implies for a
   bootstrap confidence interval computed under the two
   R versions.

4. Consider the archetypes named in the chapter (a
   data-analysis project, a manuscript, an R package, a
   simulation study, and a blog). For each, state which
   single determinant you would expect to be most
   load-bearing for its reproducibility, and justify
   the choice in one sentence.

5. Reproducibility is described as a floor rather than
   a ceiling. Construct a short, concrete example of an
   analysis that is fully reproducible (L3) and yet
   scientifically wrong, and explain what a reader
   would need beyond reproducibility to detect the
   error.

6. Read the abstract of the Open Science Collaboration
   report [@opensciencecollaboration2015] and of the
   Trisovic study [@trisovic2022]. In two or three
   sentences, explain why the two papers measure
   different things, and which one is the more relevant
   evidence for the argument of this book.

## Solutions

Attempt an exercise before expanding its solution. Several
exercises in this chapter ask the reader to work with an
article or a study of their own choosing, so they have no
single correct answer. For those the box says what a good
answer establishes, and names the failure modes worth
looking for.

::: {.callout-note icon=false collapse=true}
## Exercise 1

No single correct answer, since it depends on the article
chosen. What a good answer establishes is a classification
and a placement, both defended.

On the classification, the discipline is to ask what kind
of thing went wrong. A missing package, a package that
installs but at a newer version, a hard-coded path, a file
that was never deposited, a script that assumes a working
directory: these are computational failures, and they are
the ones this book addresses. A number that regenerates
exactly but that you believe was the wrong number to
compute is a statistical disagreement, and no amount of
environment capture touches it. Most readers find that
their failures are overwhelmingly of the first kind, which
is the point of the exercise and matches what Trisovic and
colleagues measured at scale [@trisovic2022].

On the placement, the honest default for a deposited
analysis is L0. Code and data that can be located and
downloaded are locatable, and that is what most deposits
supply. A deposit that also carries a lockfile is L1; one
carrying a container recipe is L2; one carrying evidence
that the outputs were regenerated is L3. Note that the
rung is set by what the authors deposited, not by whether
the code happened to run on your machine. An analysis can
be L0 and still run, because you were lucky, and it can be
L1 and still fail, because the lockfile pinned packages
and something below them moved.
:::

::: {.callout-note icon=false collapse=true}
## Exercise 2

No single correct answer, since it depends on the study.
What a good answer establishes is that the three
dimensions are independent, which is easiest to see by
writing three sentences that could each be true while the
other two are false.

A serviceable template. *Reproducibility fails* if a
reader with the authors' own data and code cannot
regenerate the published numbers. *Replicability fails*
if a new cohort, collecting its own data to answer the
same question, does not find the effect. *Robustness
fails* if the effect disappears when a competent analyst
makes a different but defensible modeling choice, a
different adjustment set, a different handling of missing
data, a different cut point.

On which failure is most damaging to the clinical
conclusion, the defensible answer is replication, and the
reasoning matters more than the answer. A reproducibility
failure impugns the record but leaves open whether the
finding is true. A robustness failure shows the finding
depends on a choice, which weakens it without refuting
it. A replication failure is evidence that the effect is
not there. A reader who argues instead that
reproducibility is most damaging, on the grounds that it
undermines the ability to evaluate the other two at all,
has made a reasonable case and should say so explicitly.
:::

::: {.callout-note icon=false collapse=true}
## Exercise 3

Yes, the choice of sampler still changes the result, and
for the same reason as in the worked example. Both
`sample(1:10)` and a bootstrap resample,
`sample(x, replace = TRUE)`, obtain their randomness by
drawing from the same underlying generator, so a change
of `sample.kind` changes the index sequence and therefore
the resample. Substituting `sample(x, length(x),
replace = TRUE)` for the shuffle in `draw_sample()` and
calling it under `Rounding` and `Rejection` at seed 123
produces two different resamples.

The implication for a bootstrap confidence interval is
the substantive part of the exercise. The interval is a
function of the resamples, so two R versions straddling
3.6.0 produce different intervals from the same seed, the
same data, and the same code. The difference is small
when the number of replicates is large, since both
intervals estimate the same quantity and the Monte Carlo
error shrinks, but it is not zero, and it is enough to
move a reported bound in its last printed digits. If a
published interval is quoted as excluding the null by a
narrow margin, a rerun under the other sampler can move
it across that boundary.

The lesson to draw is that a seed is not a sufficient
record of a stochastic analysis. The seed pins the
starting point, but the RNG kind determines what the
generator does with it, and the RNG kind is a session
configuration that neither a lockfile nor a container
necessarily fixes. An analysis that depends on a
bootstrap should record `RNGkind()` alongside the seed,
or set it explicitly.
:::

::: {.callout-note icon=false collapse=true}
## Exercise 4

Answers will vary in their justification, and more than
one choice is defensible for each archetype. What a good
answer establishes is that the archetypes stress
different layers, so a single reproducibility recipe
serves them unevenly.

A defensible set. For a **data-analysis project**, the
data itself, since the analysis is worthless without the
extract it ran on and that extract is often the artifact
least likely to travel. For a **manuscript**, the
rendering toolchain, since the deliverable is a document
and the numbers in it must be regenerated rather than
retyped. For an **R package**, the declared dependency
set, since the package is consumed by other code and its
correctness is a claim about behavior under a stated
environment. For a **simulation study**, the random-number
discipline, since the entire output is a function of the
seed and the generator, as this chapter's worked example
shows. For a **blog**, the pinned package versions, since
posts are written once and read for years while the
packages beneath them move.

A reader who answers 'the data' for every archetype has
not made an error so much as missed the exercise, which
is to notice that the binding constraint moves.
:::

::: {.callout-note icon=false collapse=true}
## Exercise 5

Answers will vary. What a good answer establishes is that
reproducibility is a property of re-execution and
carries no claim about correctness.

A clean construction: a cohort analysis computes
thirty-day readmission but the cohort was defined with an
off-by-one error in the date window, so patients
readmitted on day thirty are excluded. The analysis is
containerized, its packages pinned, its outputs verified
against recorded hashes, and it regenerates the same
number on every machine and every rerun. It is L3. It is
also wrong, and it will remain wrong, identically, for as
long as anyone runs it. Other serviceable examples are a
sign error in a contrast, a covariate adjusted for that
sits on the causal pathway, and a survival analysis that
treats censoring as an event.

What a reader needs beyond reproducibility to detect the
error is subject-matter review of the code and the design:
someone who knows what thirty-day readmission means
reading the cohort definition, or a unit test asserting
the boundary case against a hand-computed answer. This is
the point at which @sec-testing takes over from
environment capture, and it is why the book treats
validation as a separate register from capture. Note the
uncomfortable corollary: reproducibility makes an error
easier to find once someone suspects it, and does nothing
to make anyone suspect it.
:::

::: {.callout-note icon=false collapse=true}
## Exercise 6

The two papers measure different things because they
address two of the three words this chapter separates.

The Open Science Collaboration ran new studies collecting
new data, and asked whether the original findings held
[@opensciencecollaboration2015]. That is *replication*: a
statement about the durability of the science. Trisovic
and colleagues took deposited code and data and asked
whether it would execute at all [@trisovic2022]. That is
*reproducibility* in the narrow sense: a statement about
re-execution, in which the science is not in question
because it never got as far as being evaluated.

For the argument of this book, Trisovic is the more
relevant evidence, and it is worth being explicit about
why, since the replication literature is the more famous.
This book is about the computational floor. It has
nothing to say about whether an effect is real, and its
whole claim is that a large fraction of deposited
analyses cannot be re-executed, which is a problem
soluble by engineering rather than by more studies. The
Open Science Collaboration measures a problem this book
does not address and cannot fix.

The stronger version of the answer notes that the two are
not independent in practice: a field that cannot
re-execute its own analyses cannot easily diagnose why a
replication failed, because it cannot establish what the
original analysis actually did.
:::

## Further reading

- The National Academies consensus report
  [@nas2019] is the authoritative source for the
  reproducibility, replicability, and robustness
  distinctions, and repays reading in full for the
  care with which it separates them.

- Goodman and colleagues [@goodman2016] document the
  terminological confusion directly and propose a
  cleaner lexicon; read it to understand why the same
  words mean opposite things across fields.

- Peng's short article [@peng2011], together with his
  earlier treatment [@peng2006], is the origin of the
  spectrum idea that becomes the levels ladder of
  @sec-levels, and states the case for reproducibility
  as a minimum standard.

- For the empirical scale of the problem, the Open
  Science Collaboration report
  [@opensciencecollaboration2015] and Begley and Ellis
  [@begley2012] establish the replication failures,
  while Trisovic and colleagues [@trisovic2022] and
  Saju and colleagues [@saju2025] isolate the
  computational ones. Baker's Nature survey
  [@baker2016] captures how researchers themselves
  perceive the crisis, and Ioannidis [@ioannidis2005]
  supplies the theoretical argument that anticipated
  it.

- On the load-bearing caveat that reproducibility is a
  floor and not a ceiling, Leek and Peng
  [@leek2015] are brief, pointed, and worth
  reading before one is tempted to equate a
  reproducible result with a correct one.

- For the historical origin of the ideal, Claerbout's
  work in reproducible geophysics [@claerbout1992] is
  the standard reference for where the modern practice
  begins.
