# Levels of Reproducibility {#sec-levels}

::: {.bookepigraph}
```{=latex}
\begin{flushright}\small
```
*Reproducibility has the potential to serve as a minimum
standard for judging scientific claims when full
independent replication of a study is not possible.*

[Roger D. Peng]{.smallcaps}, *Reproducible Research in
Computational Science* (2011)
```{=latex}
\end{flushright}
```
:::

## Learning objectives

By the end of this chapter the reader should be able
to:

- Enumerate the layers of the determinant stack on
  which a computational result stands, from the
  hardware at its base to the data at its top, and say
  which layers common practice pins and which it
  leaves to chance.
- Place an analysis on the four-rung ladder of this
  book (L0 locatable, L1 pinned packages, L2 pinned
  environment, L3 verified) and name the artifact that
  marks each transition.
- Distinguish capturing reproducibility from
  validating it, and explain why continuous
  integration adds no new pinned layer.
- Distinguish a result that is reproduced from one
  that is replicated, and navigate the shifting use of
  those two words across the literature.
- Recognize that the ladder grades every archetype of
  research compendium, a manuscript, an analysis
  project, an R package, a simulation study, and a
  blog, by the same means.
- State plainly the level an analysis has reached, and
  no higher, including the candid admission that no
  present framework pins the whole stack.

## Orientation

Consider a registry study of thirty-day readmission
after heart failure. The analyst fits a mixed-effects
model, draws a forest plot, and reports an adjusted
odds ratio of 1.24. The script runs, the manuscript
renders, the paper is accepted. Two years later a
reader requests the code, downloads it, and runs it on
a fresh laptop. The model will not fit: a package that
the analysis relied upon has changed its default
optimizer, and the estimate now returned is 1.19. Which
number is the finding? The question is not statistical.
The statistics were the same on both machines. What
differed was everything underneath the statistics, and
the analyst never recorded it.

This chapter is about naming that 'everything
underneath', and about grading how much of it a given
analysis has secured. We introduced in @sec-intro the
observation that a computational result depends on a
stack of determinants beneath the code and the data,
and that a published script pins almost none of them.
Here we make that observation precise. We shall lay out
the stack layer by layer. We then define the ladder
that this book uses to say how far down the stack a
particular effort has reached: L0 locatable, L1 pinned
packages, L2 pinned environment, and L3 verified.

The ladder is the conceptual backbone of everything
that follows. Every later chapter is, in effect, an
account of how to climb one rung of it. Pinning
packages with a lockfile (@sec-dependencies) is the
L0-to-L1 step. Pinning the environment with a container
(@sec-containers) is the L1-to-L2 step. And verifying
the outputs (@sec-verification) is the L2-to-L3 step.
Before we can teach the climbing we need the ladder
itself, together with the discipline of naming, without
flinching, the rung one has actually reached.

## The analyst's contribution

Three judgments at the center of this chapter belong
to the analyst and are automated by no tool. It is
worth stating them at the outset, because the machinery
of later chapters serves them rather than replaces
them.

1. **Choosing, and declaring, the honest level.** A
   framework can compute which rung a repository has
   reached and print it; it cannot decide what the
   author will claim. The judgment that a paper's
   reproducibility statement will match the level
   actually secured, and never assert above it, is the
   analyst's alone. It is the single habit this book
   most wishes to instill.

2. **Matching the investment to the stakes.** Climbing
   the ladder costs effort, and not every analysis
   warrants the climb. A two-day exploratory
   calculation may honestly rest at L0; a registry study
   destined for a guideline warrants L2 or L3. Deciding
   where on the ladder a given project should sit, given
   its lifespan, its audience, and the cost of being
   wrong, is a design decision, not a default.

3. **Naming what remains exposed.** No rung of the
   ladder pins the whole stack. The analyst who reaches
   L2 must still ask whether the layers left unpinned,
   the numerical libraries, the random-number
   discipline, the integrity of an external data source,
   matter for this particular analysis. Those that do
   must be closed by hand. Recognizing a live exposure
   from an inert one is a matter of subject-matter
   judgment.

::: {.callout-note}
## The vocabulary of this chapter

New terms are defined again on first use; they are
collected here for reference.

- **Determinant stack.** The layered set of things a
  computational result depends upon, from the hardware
  at the base through the environment and the packages
  to the analysis code and the data at the top. A layer
  is 'pinned' when it has been fixed to a recorded
  value.
- **Capture.** Fixing an input so that it cannot drift:
  writing a lockfile, building a container image.
  Capture sets the reproducibility level.
- **Validation.** Confirming that the captured inputs
  actually suffice, by rebuilding and re-running.
  Validation does not pin anything new; it checks.
- **The level ladder (L0 to L3).** The four-rung scale,
  L0 locatable, L1 pinned packages, L2 pinned
  environment, L3 verified, used throughout this book to
  grade how far down the stack an analysis has been
  secured.
- **Reproduced.** Obtaining the same computational
  result by re-executing the same code on the same data.
  Contrast **replicated**, obtaining a consistent
  finding from a new study with new data.
- **Lockfile.** A file, such as `renv.lock`, recording
  the exact version of every package an analysis used.
- **Container image.** A packaged environment, built
  from a recipe, that carries the operating system, the
  system libraries, and the language runtime, and runs
  identically wherever a container runtime runs.
- **RNG kind.** The random-number algorithm in force,
  set by `RNGkind()`. Two runs with the same seed but a
  different kind draw different numbers.
:::

## The stack a result stands on

We begin with the determinants, because the ladder is
only intelligible once one sees what it grades. A
computational result is the output of running some code
against some data. But the code and the data are only the
two topmost of a tall stack of dependencies, and the
result is a function of the whole stack, not of its top
two layers alone. Arranging the determinants from the
hardware upward to the analysis logic yields, depending
on how finely one splits them, nine or ten layers.

At the base sits the **hardware**: the processor
architecture, its instruction-set extensions, and the
behavior of its floating-point unit. Above it lie the
**numerical libraries**, the BLAS and LAPACK
implementations that perform the linear algebra beneath
every regression and every matrix decomposition. These
matter more than their obscurity suggests: a different
BLAS, or the same BLAS run with a different thread
count, reorders the floating-point summations and can
change a result in its low-order digits. Above these
sit the **operating system and its system libraries**,
the kernel, the C library, the locale data, the fonts,
and the document toolchain of `pandoc` and LaTeX. Then
comes the **R runtime** itself, whose default behaviors
shift between releases. The random-number change of
R 3.6.0 and the `stringsAsFactors` change of R 4.0.0
both illustrate the point.

Above the runtime sit the layers most analysts think of
first. The **package versions**, the exact release of
every R package the analysis loads, are the layer that
churns fastest and breaks analyses most often. The
**session configuration** is a cluster of settings that
shape the session before any user code runs: the
random-number kind set by `RNGkind()`, the contrasts
used for factor variables, the number of printed
`digits`, the locale that governs sorting and decimal
marks, and the time zone that governs date arithmetic.
Then come the two layers the analyst actually writes,
the **analysis code**, including every `set.seed()` call
and the order of execution, and the **input data**, its
content and its integrity. Finally, threaded through the
whole stack, sits **intrinsic non-determinism**: the use
of the wall-clock, race conditions in parallel code, and
iteration over unordered structures, each capable of
producing a different answer on two runs of identical
inputs.

Now, which of these layers does common practice pin? The
answer is sobering, and it motivates the entire book. A
bare published script pins exactly one layer with any
reliability, the analysis code, since the code is the
file itself. If the reader is fortunate, it pins a
second, the data, when it has been deposited alongside.
Every other layer is left to whatever the next machine
happens to provide. @fig-determinant-stack maps the stack
against the rung of the ladder that first secures each
layer. The reader will note that three layers, the
hardware, the numerical libraries, and intrinsic
non-determinism, are secured by no rung of the capture
ladder at all.

```{r}
#| label: fig-determinant-stack
#| fig-cap: 'The determinant stack, from the hardware at the base to the data at the top, colored by the rung of the ladder that first secures each layer. Package versions are secured at L1, the operating system and R runtime at L2, and the analysis code and data are merely locatable at L0. The session-configuration layer is only partially pinned at L2: the locale and the time zone are fixed, but the RNG kind, the contrasts, and the number of printed digits are not. Three further layers, the hardware, the numerical libraries, and intrinsic non-determinism, are pinned by no rung of the capture ladder.'
#| fig-alt: 'A horizontal bar chart of nine stacked layers, hardware at the bottom and intrinsic non-determinism at the top, each bar colored by the rung that secures it. Hardware, numerical libraries, and intrinsic non-determinism are colored as unpinned by any rung; the operating system, system libraries, and R runtime are colored L2; package versions are colored L1; analysis code and input data are colored L0, that is, locatable only. Session configuration is colored L2 but is noted as only partially pinned.'
#| fig-width: 7
#| fig-height: 4.4
library(ggplot2)
library(tibble)

stack <- tibble::tibble(
  layer = c('Hardware',
            'Numerical libraries (BLAS, LAPACK)',
            'Operating system, system libraries',
            'R runtime',
            'Package versions',
            'Session configuration',
            'Analysis code',
            'Input data',
            'Intrinsic non-determinism'),
  height = 1:9,
  secured = c('Unpinned by any rung',
              'Unpinned by any rung',
              'L2 pinned environment',
              'L2 pinned environment',
              'L1 pinned packages',
              'L2 partially pinned',
              'L0 locatable',
              'L0 locatable',
              'Unpinned by any rung')
)

stack$secured <- factor(
  stack$secured,
  levels = c('L0 locatable', 'L1 pinned packages',
             'L2 pinned environment', 'L2 partially pinned',
             'Unpinned by any rung'))

ggplot(stack, aes(x = reorder(layer, height),
                  y = 1, fill = secured)) +
  geom_col(width = 0.9) +
  coord_flip() +
  scale_fill_manual(
    values = c('L0 locatable' = '#6baed6',
               'L1 pinned packages' = '#3182bd',
               'L2 pinned environment' = '#08519c',
               'L2 partially pinned' = '#fd8d3c',
               'Unpinned by any rung' = '#737373'),
    name = NULL) +
  labs(x = NULL, y = NULL,
       title = 'Which rung first secures each layer') +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_blank(),
        panel.grid = element_blank(),
        legend.position = 'bottom')
```

Two features of the session-configuration layer deserve
a word, because they are the layers most often forgotten
by an analyst who has already adopted a lockfile and a
container. The random-number kind is the more dangerous.
An analysis that seeds its generator and draws a sample
is reproducible only if the algorithm behind the seed is
also fixed. The change of the default `sample.kind` in
R 3.6.0 broke exactly this, silently, for every analysis
that had set a seed under the old default. Given that
simulation studies, multiple imputation, and bootstrap
confidence intervals are staples of health research, the
RNG kind is a first-order concern, not a footnote. The
locale is the second. `sort()` and `order()` on a
character vector depend on `LC_COLLATE`, and the parsing
of a decimal mark depends on `LC_NUMERIC`. An analysis
developed under one locale and re-run under another can
therefore reorder a table or misread a number without
any error.

::: {.callout-note icon=false collapse=true}
## Check your understanding: locating a failure on the stack

An epidemiologist reports that a colleague, re-running
her cohort analysis, obtained an incidence rate identical
to five decimal places but a different value in the sixth.
The code, the data, the package versions, and the
operating system were all identical. Which layer of the
determinant stack is the most likely culprit, and is the
discrepancy a bug?

::: {.callout-note icon=false collapse=true}
## Answer

Neither the code nor the data nor the pinned layers can
be at fault, since all were identical. A discrepancy
confined to the low-order digits is the signature of the
numerical-library layer. The two machines used a
different BLAS implementation, or the same one with a
different thread count, and so the floating-point
summations were reordered. It is not a bug in the
analysis. It is the reason that byte-for-byte identity
across different hardware is a stronger claim than
number-for-number agreement on a fixed architecture.
And it is a layer that no rung of the capture ladder
pins.
:::
:::

## The ladder: capture sets the level, validation confirms it

With the stack in view we can define the ladder. The
central design idea, which we take from the framework
this book uses and develop as a general principle, is
that the features acting on reproducibility divide into
two kinds that are constantly confused. Separating them
is what makes an honest grade possible.

The **capture axes** fix inputs, and therefore set the
level. There are two of them: the package backend, which
pins the package layer, and the computational
environment, which pins the operating system, the
runtime, and the system libraries. Adding a capture
feature raises the level; removing one lowers it. The
**validation layer** confirms the captured level without
capturing anything new. Its members are continuous
integration, the dependency-consistency check, and the
output-verification command. Removing a validation
feature leaves a repository exactly as reconstructable as
before; it only stops the automated checking. This is why
continuous integration, which we develop in
@sec-verification, is best understood as the harness that
runs the verification rather than as a layer of
reproducibility in its own right. Pinning is capture;
re-running to confirm is validation.

The four rungs follow from this distinction. Levels L0
through L2 are reached by the capture axes, and L3 is
reached by the validation layer. @tbl-ladder states each
rung, what it pins, the artifact that marks it, and what
the rung means for a reader of health research.

```{r}
#| label: tbl-ladder
#| tbl-cap: 'The reproducibility ladder used throughout this book. Each rung names what it pins, the primary artifact that marks it, and what reaching it means for a reader of health research.'
library(tibble)
library(knitr)

ladder <- tibble::tibble(
  Level = c('L0 locatable',
            'L1 pinned packages',
            'L2 pinned environment',
            'L3 verified'),
  `What is pinned` = c(
    'Nothing computational; the source is under version control',
    'The exact version of every R package',
    'The operating system, the R runtime, and the system libraries as well',
    'Nothing new; the recorded outputs have been regenerated and checked'),
  `Primary artifact` = c(
    'A version-controlled repository',
    'A lockfile such as renv.lock',
    'A container image or a Nix environment',
    'A passing verification run comparing regenerated outputs against a recorded baseline, re-run automatically by continuous integration'),
  `Health-research reading` = c(
    'A reviewer can find the code but cannot yet rebuild it',
    'A colleague installs the same packages the analysis used',
    'The analysis yields the same numbers on a different machine',
    'The reported table has been shown to regenerate, not merely asserted to'))

knitr::kable(ladder)
```

The rungs are cumulative. An L2 analysis is also
locatable and package-pinned; the label names the
highest rung reached. The transition that matters most
for a first-year student to internalize is the one
between L1 and L2. A lockfile alone, the L1 artifact,
pins the packages but not the R version, the operating
system, or the system libraries a package links against.
A lockfile restored under a different R version may fail
to install or, more insidiously, may install and behave
differently. The container image, the L2 artifact, fixes
what the lockfile cannot. The two are complements, not
substitutes, and a genuinely pinned-environment analysis
carries both.

Where does the ladder come from, and is it arbitrary?
It is not. It consolidates several graded views of
reproducibility that the literature has offered in
different vocabularies. @peng2011 arranged computational
work along a spectrum, from the merely published, through
code available, code and data available, and linked and
executable, to full replication. Our L0, L1, and L2 track
the lower and middle reaches of that spectrum. The
artifact-review badges used by the ACM
and several computing venues express a related ordering,
from artifacts that are available, through artifacts
evaluated as functional and then reusable, to results
independently reproduced [@acm2020]. Our L3 is the
analogue of the 'Results Reproduced' badge. The
correspondence is close but not exact. The 'functional'
and 'reusable' badges grade how well an artifact is
documented and exercised rather than the
package-versus-environment pinning depth that separates
L1 from L2, so the middle rows are analogies rather than
alignments. The contribution of the L-scheme is not a new
definition of reproducibility but a mapping of these
graded views onto artifacts a tool can detect. The point
is that the level of a repository can then be determined
mechanically and reported honestly, rather than asserted
by its author.

::: {.callout-note icon=false collapse=true}
## Check your understanding: capture or validation?

A laboratory adds a continuous-integration workflow that
rebuilds its compendium and runs the tests on every push.
A colleague objects that this cannot improve
reproducibility, because it pins no new layer of the
determinant stack. Is the colleague correct, and has the
laboratory wasted its effort?

::: {.callout-note icon=false collapse=true}
## Answer

The colleague is correct on the narrow point and wrong on
the conclusion. Continuous integration is a validation
feature, not a capture feature: it pins no new layer, and
removing it would leave the compendium exactly as
reconstructable as before. What it adds is confirmation.
It re-runs the captured analysis in a clean environment
that carries none of the author's undeclared state. It
therefore catches the silent decay that a
captured-but-never-retested analysis suffers, and the
'works on my machine' failure at its source. The
laboratory has moved from L2, a pinned environment
asserted to reproduce, toward L3, a pinned environment
shown to reproduce. That is the whole purpose of the
capture-versus-validation distinction.
:::
:::

## Reproduced, replicated, and a warning about the words

The ladder grades re-execution, and re-execution is only
one of the things the word 'reproducibility' is made to
carry. A first-year student must fix the terms before
reading the literature, because the same two words are
attached to opposite meanings across the sources a reader
will meet.

The National Academies, in their 2019 consensus report,
adopt the distinction this book follows [@nas2019].
*Reproducibility* is obtaining consistent computational
results using the same input data, the same code, and the
same conditions of analysis: it is a statement about
re-execution, and it is what the ladder grades.
*Replicability* is obtaining consistent results across
studies aimed at the same question, each with its own
data: it is a statement about the robustness of a
scientific finding. The two are independent. A result can
be perfectly reproducible and entirely wrong, if the code
faithfully regenerates a mistaken number every time. And a
finding can replicate across laboratories while no single
analysis in the set is computationally reproducible. Leek
and Peng press exactly this point, that a reproducible
analysis can still be wrong, and that reproducibility is
therefore a floor and not a ceiling [@leek2015]. Patil,
Peng, and Leek go further and offer a statistical framing
of the distinction, casting replication as a question
about the agreement of estimates across studies rather
than about the re-running of code [@patil2016].

Here is the warning. The current ACM artifact-badge
vocabulary, in its 2020 revision, agrees with the National
Academies usage rather than inverting it [@acm2020]. Under
both conventions, 'Reproduced' denotes obtaining a prior
result using the author's own artifacts and the same
setup, and 'Replicated' denotes obtaining it
independently, with a different team and a new setup. The
genuine word-swap that trips students belongs to two
other places. One is the pre-2020 ACM badging, which
attached the two words in the opposite direction along
the re-execution axis. The other is some biomedical and
field usages, which a reader may still encounter, where
'reproduce' names an independent study rather than a
re-run. A reader who
carries the sense of the words from one of those older
documents into a current one will be misled, silently and
without any error to warn them, in exactly the manner of a
shifted package default. This book fixes the meaning and
states it once: throughout, L3 denotes the ACM 'Results
Reproduced' sense, which now coincides with the National
Academies reproducibility sense. That sense is that the
captured analysis, rebuilt from its pinned definition and
re-run, regenerates its own recorded outputs. It is
deliberately not replication in the sense of an independent
study on new data, which no software framework can supply.
When the reader meets 'reproduced' or 'replicated' in a
journal's author instructions or a venue's badging policy,
the first task is to determine which convention that
document uses. The word alone does not tell them.

::: {.callout-important}
## The honest level

A reproducibility claim must be matched to the level
actually reached, and never asserted above it. An
analysis that carries a lockfile but no container is at
L1, and a manuscript that describes it as 'fully
reproducible' overstates by two rungs. The remedy is not
modesty for its own sake but accuracy: state the rung,
name the artifact that marks it, and name the layers left
exposed. 'This analysis is pinned at L2, with the RNG kind
fixed and the input data checksummed; it has not been
verified under `zzc verify` and so is not claimed at L3'
is a sentence that a reviewer can act upon. 'Fully
reproducible' is not.
:::

## Two numbering schemes, and which this book uses

A reader who consults the documentation of the framework
this book uses will encounter a second numbering scheme,
and the collision is a genuine source of confusion that we
disarm here rather than leave to chance.

The framework's teaching materials include a vignette that
presents a *five-level* ramp: Level 1, a basic R project
with manual package management; Level 2, the addition of
`renv`; Level 3, the addition of Docker; Level 4, the
addition of unit testing; and Level 5, the addition of
continuous integration. That scale counts *which tools a
project has adopted*, and it is a pedagogical on-ramp,
meant to be walked from the bottom as a project matures.
It is a useful thing, but it is not the scale of this book.

The scale of this book is the *architectural* ladder of the
preceding sections, L0 through L3, and it grades something
different: *how far down the determinant stack a result has
been locked, and whether the lock has been verified*. The
two axes are related but not identical. Adopting `renv`
(the tool scheme's Level 2) is what lifts a project to L1.
Adopting a container (the tool scheme's Level 3) is what
lifts it to L2. The tool scheme's testing and
continuous-integration levels belong to the validation
layer that reaches L3. But one can imagine a project that
has adopted every tool and still verified nothing, and the
architectural ladder, not the tool count, is what tells the
reader whether the result has been shown to reproduce.

To avoid ambiguity we adopt a single convention. This book
uses the L0 to L3 architectural ladder throughout, in every
chapter and every worked example. Where the framework's own
documentation speaks of 'Level 1' through 'Level 5', it is
using the tool-adoption ramp, and the reader should read it
as such and not conflate it with our L0 to L3. When we
write a bare 'L2', we always mean the architectural rung.

## The ladder grades every archetype

It would be a mistake to read the ladder as a scale for
manuscripts alone. A research compendium, as
@sec-compendium develops in full, comes in several
archetypes, and the ladder grades each of them by the same
means and to the same rungs.

A **data-analysis project** and a **scholarly manuscript**
are the familiar cases, and for them the reading is the
one given above: the lockfile lifts to L1, and the container
lifts to L2. The verification run, for a manuscript,
re-renders the report and compares its figures and tables
against a recorded baseline, and this is what lifts it to
L3. An **R package**, by contrast, has no report to render.
Its L3 verification is instead the passing of `R CMD check`
and its test suite in a clean environment, rather than the
regeneration of a manuscript, though the pinning of its
packages and its environment is otherwise identical. A
**simulation study** reaches the same rungs by the same
artifacts, but its exposure on the session-configuration
layer is acute, because its outputs are functions of the
random-number stream. An unseeded or wrongly seeded
simulation can sit at L2, with a perfectly pinned
environment, and still fail to reproduce. For this
archetype, the discipline of fixing the RNG kind and the
seed is the difference between a nominal L2 and a real
one. A **blog** renders many posts rather than one report.
Its L3 verification is the successful rebuild of the whole
site from its pinned environment, and the ladder applies
post by post exactly as it applies to a single manuscript.

The lesson is that the ladder is a property of the
determinant stack, which every archetype shares, and not of
any particular directory layout. A package or a blog reaches
L2 by the same container that an analysis project does, and
reaches L3 by a verification appropriate to what it produces.

## No framework pins the whole stack

We end the exposition where honesty requires, with the
boundary of the guarantee. It would be incoherent to spend
a chapter teaching that a reproducibility claim must be
matched to the level actually reached, and then to imply
that reaching L2 or L3 pins the entire determinant stack.
It does not, and no present framework does.

By its own published accounting, the framework this book
uses robustly pins something like four of the ten
determinants: the operating system and system libraries
(as strongly as its base image is pinned), the R package
versions (nominally, through the lockfile), the locale and
time zone, and the source code. It leaves others to
convention or to the base image: the random-number kind is
not pinned by default, the BLAS and LAPACK threading regime
is inherited from the base image and neither pinned nor
documented, the integrity of external data is recorded by
provenance but not by checksum, and the fonts and the
graphics device are pinned only for some profiles. These
are not defects peculiar to one tool. They are the layers
that the whole class of integrative frameworks currently
leaves exposed, and an honest L2 or L3 claim names them
rather than papering over them. The reader will therefore
find, at each rung of the ladder in the chapters ahead, an
explicit account of what is secured and what remains
exposed. Reach plus honesty about the boundary of that
reach is the standard this book holds every effort to,
including the framework it teaches.

## Worked example

To make the ladder concrete, consider a single analysis
walked up all four rungs. A biostatistician has a folder
containing `analysis.R`, which reads `cohort.csv`, fits a
Cox model with the `survival` package, and writes a hazard
ratio to `results.csv`. We shall place it on the ladder and
name each transition.

At the start the folder is on the analyst's laptop and
nowhere else. It is at no rung of the ladder at all, because
it cannot even be found by a second person. The first act,
placing the folder under version control and pushing it to
a repository, lifts it to **L0, locatable**. A reviewer can
now find `analysis.R` and read it. Nothing computational is
pinned; a reviewer who runs it will install whatever version
of `survival` happens to be current, which may not be the
version the analyst used.

The analyst now runs `renv::init()` and `renv::snapshot()`,
which we develop in @sec-dependencies. This writes a
`renv.lock` recording that the analysis used `survival`
version 3.5-7, together with the exact version of every
other package in the closure. The folder is now at **L1,
pinned packages**. A colleague who restores from the
lockfile obtains the same `survival` the analyst used. But
the colleague's laptop runs a newer R and a different
operating system, and one of the packages links against a
system library that is absent, so the restore fails to
build.

To close that gap the analyst adds a container, developed in
@sec-containers, built from a dated `rocker` base image
pinned by its content digest. The image carries a fixed
operating system, a fixed R, and the system libraries the
packages need, and the lockfile restores inside it. The
folder is now at **L2, pinned environment**. It will build
and run identically on the colleague's laptop, on a
reviewer's, and on a server. The analyst also sets the RNG
kind and a seed in the project `.Rprofile`, because the
analyst computes bootstrap confidence intervals for the
hazard ratio, closing a
session-configuration exposure that L2 does not close by
default.

Finally the analyst adds a verification step, developed in
@sec-verification, that rebuilds the image, re-runs
`analysis.R`, and compares the regenerated `results.csv`
against a recorded baseline. The first time it runs it
passes, and a continuous-integration service re-runs it on
every change. The folder is now at **L3, verified**: the
reported hazard ratio has been shown to regenerate, not
merely asserted to. The honest reproducibility statement for
the paper now writes itself, naming the rung, the artifacts,
and the one layer, the cross-architecture numerical
determinism of the bootstrap, that remains outside the
guarantee.

## Collaborating with an LLM

A large language model is a capable assistant for reasoning
about levels. But it is confident in a register that this
chapter has just taught the reader to distrust, and its
confidence is not calibrated to the honest level. The
following triples illustrate productive prompts and the
verification each demands.

**Prompt.** 'Here is my project directory listing. What
reproducibility level has it reached on the L0 to L3 ladder,
and what would it take to reach the next rung?'

*Watch for.* The model will read the presence of a
`renv.lock` as L1 and a `Dockerfile` as L2, which is the
right heuristic. But it cannot tell from a listing whether
the lockfile is current, whether the image is pinned by a
mutable tag or an immutable digest, or whether a
verification run has ever passed. It may report L3 on the
strength of a continuous-integration file that has never
been run green.

*Verification.* Confirm each rung against the artifact
itself, not its filename. For L1, check that the lockfile is
in step with the code. For L2, check that the base image
carries a digest and that the library is not shadowed by a
bind mount. For L3, find an actual passing verification run.
The presence of a file is capture asserted; a green run is
validation observed.

**Prompt.** 'Draft the reproducibility statement for my
methods section.'

*Watch for.* The model gravitates to the phrase 'fully
reproducible', which this chapter has argued is almost never
an honest claim. It will also tend to conflate 'reproducible'
and 'replicable' in the National Academies sense, or to use
the ACM sense without saying so.

*Verification.* Rewrite the draft to name the specific rung,
the artifacts that mark it, and the layers left exposed, and
to state which convention of 'reproduced' is meant. Reject
any sentence you could not defend to a reviewer who ran the
code.

**Prompt.** 'My simulation gives slightly different numbers
each time even inside the container. Why, and is it a bug?'

*Watch for.* The model may propose plausible but wrong
culprits, a package version or an operating-system
difference, that the container has already pinned. It may
also not distinguish an unseeded RNG (a real defect) from
low-order-digit variation due to BLAS threading (an inherent
limit at L2).

*Verification.* Trace the discrepancy to a layer of the
determinant stack. If the numbers differ in all digits, look
to the RNG kind and the seed, which the analyst controls. If
they differ only in the low-order digits, suspect the
numerical-library layer, which no rung of the capture ladder
pins, and decide whether that residual matters for the
inference at hand.

## Exercises

1. For an analysis of your own, list the nine or ten layers
   of the determinant stack and, for each, state whether your
   current setup pins it, and by what artifact. Identify the
   lowest layer you have left unpinned.

2. Take a published paper in your field that provides code.
   Without running it, place it on the L0 to L3 ladder from
   the materials the authors supply, and write the
   one-sentence reproducibility statement the paper should
   have carried.

3. Explain, to a colleague who believes that a Docker image
   makes an analysis 'fully reproducible', two determinants
   that the image does not pin and one circumstance in which
   each would cause two runs to disagree.

4. The framework's documentation grades a project 'Level 4'
   on its five-level tool-adoption scale. State what this does
   and does not tell you about the project's rung on the L0 to
   L3 architectural ladder, and name one artifact you would
   inspect to settle the question.

5. A colleague reports that a finding 'replicated' in a second
   cohort. A reviewer reads this as an ACM artifact badge and
   asks for the container. Explain the miscommunication in
   terms of the two conventions for the word, and state which
   sense the colleague almost certainly intended.

6. Construct a small simulation in R that draws a sample under
   a fixed seed, and show that changing the RNG kind with
   `RNGkind()` changes the draw while leaving the seed
   unchanged. Explain which layer of the determinant stack
   this demonstrates and why a container alone would not
   protect against it.

## Solutions

Attempt an exercise before expanding its solution. Two of
these ask the reader to work with an analysis or a paper of
their own, so they have no single correct answer; for those
the box says what a good answer establishes.

::: {.callout-note icon=false collapse=true}
## Exercise 1

No single correct answer. What a good answer establishes is
that the reader can name the layers in order and attach an
artifact, or an admission, to each.

Working up from the base: **hardware**, pinned by nothing
short of running on the same machine; **numerical
libraries**, pinned by the container image if it fixes the
BLAS, and otherwise not at all; **operating system and
system libraries**, pinned by the container image;
**R runtime**, pinned by the container image, and merely
*recorded* by the lockfile's `R` block; **package
versions**, pinned by `renv.lock`; **session
configuration**, partly pinned by the image through the
locale and the time zone, and not pinned at all for the RNG
kind, the contrasts, and the printed digits; **analysis
code**, pinned by version control; **input data**, pinned
by a checksum or an archived deposit, and often only
locatable; and **intrinsic non-determinism**, such as
parallel reduction order, pinned by nothing.

Most readers find their lowest unpinned layer is either the
numerical libraries or the session configuration, and both
answers are common and honest. A reader who finds their
lowest unpinned layer is the input data has discovered
something more urgent than anything this chapter teaches,
because the analysis is not reconstructable at all.
:::

::: {.callout-note icon=false collapse=true}
## Exercise 2

No single correct answer. What a good answer establishes is
that the rung is read off the *deposited artifacts*, not off
the quality of the paper or the reputation of the authors.

The procedure is mechanical. Is there code and data that
can be located? That is L0. Is there a lockfile, or an
equivalent exact version record? That is L1. Is there a
container recipe, or another artifact pinning the runtime
and the system libraries? That is L2. Is there evidence
that the outputs were regenerated from the pinned
definition, a CI log, a verification report, recorded output
hashes? That is L3. Stop at the highest rung for which an
artifact is actually present.

The reproducibility statement should then claim that rung
and no more. For the common L0 case: 'The analysis code and
the derived dataset are deposited at [location]. The
software environment was not recorded, so re-execution may
require package versions that differ from those used here.'
The second clause is the part most papers omit, and it is
the part that makes the statement honest rather than
merely modest.
:::

::: {.callout-note icon=false collapse=true}
## Exercise 3

Two determinants an image does not pin, with a
circumstance for each.

**The numerical libraries, when the image does not fix
them.** An image built on a base that links against a
BLAS chosen at run time, or one run on hosts that expose
different thread counts, will reorder floating-point
summations. Two runs of the same image on two machines can
then return an incidence rate agreeing to five decimal
places and differing in the sixth. The image pinned the
software and not the arithmetic beneath it.

**The session configuration, specifically the RNG kind.**
An image fixes the R version, and the R version determines
the *default* sampler, so this exposure is narrower than
it first appears. It bites when the analysis sets
`RNGkind()` itself, or inherits it from a project
`.Rprofile` that is mounted rather than baked in, or when
a package changes the generator as a side effect. Two runs
then draw different samples from the same seed, and a
bootstrap interval moves.

A third acceptable answer is the input data: an image
mounting a host directory reproduces the environment and
not the extract, so a changed file underneath produces a
different result with the image unchanged. A fourth is
intrinsic non-determinism from parallelism, where the
reduction order of a parallel sum varies run to run on the
same machine.

The general form of the answer is that a container pins
the environment and the environment is not the whole
stack. It is the L1-to-L2 step, not the whole ladder.
:::

::: {.callout-note icon=false collapse=true}
## Exercise 4

'Level 4' on the framework's scale means the project has
adopted a particular set of *tools*: an R project, `renv`,
Docker, and unit testing. It is a statement about tool
adoption, and it is read from the bottom up, so Level 4
implies the first three as well.

What it tells you about the architectural ladder is
therefore quite a lot, but indirectly and with one gap. The
presence of `renv` implies the machinery for L1; the
presence of Docker implies the machinery for L2. What it
does not tell you is whether any of it has been
*verified*, because the tool scale counts adoption and the
architectural ladder grades locking and verification. A
project can hold every tool in the list and have never
rebuilt from its pinned definition, in which case it sits
at L2 and is asserted rather than shown to reproduce. L3
is the tool scale's Level 5 territory, and Level 4 is
below it.

The artifact to inspect is the continuous-integration
record: a log showing that the container was rebuilt from
the recipe and the analysis re-run, and its outputs checked
against recorded values. Absent that, the honest reading of
'Level 4' is L2. Inspecting `renv.lock` and the Dockerfile
tells you the capture artifacts exist; only the CI record
tells you the capture works.
:::

::: {.callout-note icon=false collapse=true}
## Exercise 5

The colleague almost certainly meant the National
Academies sense: a second cohort, collecting its own data,
found a consistent result. That is *replication*, a
statement about the durability of the finding, and it is
the sense in which the word is ordinarily used in health
research.

The reviewer read it in an artifact-badging frame, where
the pair of words attaches to re-execution rather than to
new data. Under the current 2020 ACM vocabulary the two
conventions actually agree, so the reviewer's request is
not simply a misreading of the current standard. The
miscommunication traces to one of two places: the
pre-2020 ACM badging, which attached the two words in the
opposite direction, or a field usage in which 'reproduce'
names an independent study rather than a re-run. A reviewer
carrying the sense of the words from an older document into
a current one is misled silently, with no error to warn
them.

The practical resolution is that the container request is a
category error given what the colleague did. A container
would let the reviewer re-execute the *original* analysis,
which is not what the colleague claimed and not what a
second cohort demonstrates. The colleague should answer by
naming the convention, stating that the second cohort
collected its own data, and offering the artifact that
actually matches the claim, which is the second cohort's
own analysis code.
:::

::: {.callout-note icon=false collapse=true}
## Exercise 6

The demonstration is the worked example of @sec-crisis in
miniature.

```r
draw <- function(kind) {
  suppressWarnings({
    RNGkind(sample.kind = kind)
    set.seed(42)
    sample(1:10, 5)
  })
}
original <- RNGkind()
draw('Rounding')
draw('Rejection')
suppressWarnings(RNGkind(
  kind = original[1], normal.kind = original[2],
  sample.kind = original[3]
))
```

The two calls share a seed and return different draws.
Restoring the original generator afterward matters, since
`RNGkind()` sets session state that would otherwise persist
and quietly affect everything downstream.

The layer demonstrated is the **session configuration**
layer, which sits above the package versions and below the
analysis code. It is the layer this chapter singles out as
only partially pinned at L2: an image fixes the locale and
the time zone, and the R version it carries fixes the
*default* sampler, but the RNG kind, the contrasts, and the
number of printed digits are all settable from inside the
session and are not properties of the image.

A container alone would not protect against this because
the change is made by the analysis, inside the container,
after the environment is fixed. Two runs of the identical
image produce different draws if the code sets the sampler
differently, or if a mounted `.Rprofile` does. The
protection is not a heavier container but an explicit
declaration: set `RNGkind()` in the analysis, record it
alongside the seed, and treat it as part of the analysis
code rather than part of the environment.
:::

## Further reading

- @nas2019 is the National Academies consensus report that
  fixes the reproducible-versus-replicable terminology this
  book follows; read its opening chapter for the definitions
  and the reasoning behind them.
- @peng2011 introduces the reproducibility spectrum that our
  ladder consolidates, and argues for reproducibility as a
  minimum standard when full replication is impractical; the
  earlier @peng2009 applies the idea to epidemiology directly.
- @leek2015 makes the case that reproducibility is a floor and
  not a ceiling, and @patil2016 offers a statistical framing of
  the distinction between reproducing a computation and
  replicating a finding.
- @acm2020 documents the artifact-review badges whose
  'available', 'functional', 'reusable', and 'reproduced'
  ordering our ladder maps onto; read it alongside @nas2019 to
  see how the 2020 revision aligns the two vocabularies.
- @turingway2022 crosses the reproducibility axis with a
  what-varies axis, same or different data against same or
  different analysis, and is the most accessible community
  handbook for a first-year student.
- For the determinant stack and the capture-versus-validation
  distinction as a framework applies them, @sandve2013 and
  @wilson2017 give the practices, the former stressing the
  recording of random-number seeds and exact versions that the
  ladder's session-configuration layer names.
