# Pinning Packages with renv {#sec-dependencies}

::: {.bookepigraph}
```{=latex}
\begin{flushright}\small
```
*In order to exactly reproduce a given result, it may be
necessary to use programs in the exact versions used
originally.*

[Sandve, Nekrutenko, Taylor, and Hovig]{.smallcaps}, *Ten
Simple Rules for Reproducible Computational Research*,
Rule 3 (2013)
```{=latex}
\end{flushright}
```
:::

## Learning objectives

By the end of this chapter the reader should be able
to:

- Explain the package-churn failure mode and why
  pinning package versions is the step that lifts an
  analysis from locatable (L0) to package-pinned (L1).
- Describe what an `renv.lock` file records, and carry
  out the init, snapshot, and restore workflow that
  produces and consumes it.
- State the dependency-triad invariant relating the
  code, the `DESCRIPTION`, and the lockfile, and run
  the automated check that keeps the three in
  agreement.
- Configure a dated snapshot of a package repository
  so that a restore draws fixed, pre-compiled binaries.
- Delimit what `renv` does and does not pin, and thus
  explain why a lockfile alone cannot reach the
  pinned-environment level.
- Compare `renv` with the lighter, date-based
  alternatives `rang`, `groundhog`, and `capsule`.

## Orientation

Consider a familiar situation. A doctoral student
completes the analysis for a registry study of
hospital readmissions, renders a clean report, and
deposits the code alongside the manuscript. The work
is under version control, so a reader can locate every
script; on the ladder of @sec-levels the analysis has
reached L0, locatable. Eighteen months later a
reviewer asks for a minor revision. The student pulls
the repository onto a new laptop, installs the
packages afresh, and reruns the analysis. The model
now reports a different coefficient. Nothing in the
code changed. What changed is that `dplyr` had moved
from version 1.1.0 to 1.2.1, a grouping default had
been refined, and one intermediate summary was
computed over a slightly different partition. The
number in the published table can no longer be
regenerated, and the student cannot say with certainty
which number was correct.

This is the package-churn failure mode, and it is the
specifically computational failure that the empirical
studies of @sec-crisis isolate. The R package
ecosystem is healthy precisely because it changes:
maintainers fix bugs, sharpen defaults, and deprecate
awkward interfaces. Every such improvement is,
however, a potential threat to the exact reproduction
of an analysis written against the older behavior. A
script names the packages it loads but almost never
the versions it loaded them at, and so it inherits
whatever versions the next machine happens to provide.
Reproducibility fails in the gap between the packages
the analysis was written against and the packages a
later installation supplies.

The remedy for this one layer is a lockfile: a file
that records the exact version of every package the
analysis used, so that the same set can be reinstalled
later. Recording that file is the transition from L0 to
L1, from locatable to package-pinned, and it is the
whole business of this chapter. We shall be equally
careful about the transition's limit. A lockfile pins
the packages and nothing beneath them, so an analysis
at L1 is pinned at the package layer and exposed at
every layer below, a boundary that motivates the
container of @sec-containers.

## The analyst's contribution

The tooling in this chapter automates a great deal, but
three judgments remain irreducibly the analyst's.

1. **Deciding when to snapshot, and what a snapshot
   claims.** A lockfile is a dated assertion that these
   versions produced this result. The analyst chooses
   the moment to make that assertion, typically once an
   analysis is stable, and must understand that a
   snapshot taken mid-exploration pins whatever
   half-installed state the library happens to be in.
   The tool records; the analyst decides what is worth
   recording.

2. **Distinguishing a result dependency from a
   development tool.** No scanner can know that
   `languageserver` powers the editor while `dplyr`
   powers the analysis. The analyst must keep the
   lockfile an honest, minimal record of what
   determines the result, and keep mere tooling out of
   it, a distinction we develop below and that the
   accompanying whitepaper treats at length.

3. **Matching the reproducibility claim to the layer
   actually pinned.** A lockfile secures the package
   layer and no other. Presenting a lockfile-equipped
   analysis as fully reproducible, when the R version,
   the operating system, and the system libraries
   remain unpinned, overstates what has been achieved.
   The honesty is the analyst's responsibility, not the
   tool's.

::: {.callout-note}
## The vocabulary of this chapter

- **Lockfile.** A machine-generated file, `renv.lock`,
  recording the exact version, source repository, and
  content hash of every package in a project's library.
- **Package library.** The directory into which R
  installs packages. A project-local library belongs to
  one project rather than being shared across all of a
  user's work.
- **Snapshot.** The act of scanning a project, resolving
  the versions currently installed, and writing them
  into the lockfile.
- **Restore.** The reverse act of reading the lockfile
  and reinstalling exactly the packages it records.
- **`DESCRIPTION`.** The human-authored file that
  declares a project's direct dependencies by role
  (`Imports`, `Suggests`, `Depends`), with optional
  version floors.
- **Content hash.** A short fingerprint computed from a
  package's contents, so that two installations claiming
  the same version can be confirmed byte-identical.
- **Package repository.** A server, such as CRAN or the
  Posit Package Manager, from which packages are
  installed.
- **Dated snapshot.** A view of a repository frozen as
  it stood on a particular calendar day, so that an
  install run today and one run next year fetch the same
  versions.
:::

## From packrat to renv

The idea of a project-local library governed by a
lockfile is not new to R. The `packrat` package,
introduced in 2013, was the first widely adopted
attempt to give each project its own isolated set of
packages recorded in a manifest. Its mechanics were
sound but its ergonomics were not, and it acquired a
reputation for surprising behavior that discouraged
routine use. The `renv` package, introduced by
@ushey2020 and now the community standard, superseded
it in 2019. It kept the essential idea, a
project-local library plus a lockfile. But it rebuilt the
surface around a small, predictable set of verbs, a
human-readable JSON lockfile, and a shared package
cache that lets many projects share one physical copy
of a given package version.

The design rests on a separation the analyst should
hold clearly in mind. The **library** is where
packages physically live; the **lockfile** is the
written record of what should live there. The two are
kept in agreement by two complementary operations. A
snapshot reads the library and updates the lockfile to
match it; a restore reads the lockfile and updates the
library to match it. Everything `renv` does is a
variation on keeping these two representations in step.

### What the lockfile records

The lockfile is the artifact that does the pinning, so
it repays close reading. For each package it records
four things that together make a later reinstallation
unambiguous: the exact version, the source repository
the package came from, a content hash, and the
package's own declared dependencies. The top of the
file additionally records the R version and the
repositories in force. A lightly abbreviated excerpt,
shown here as a static block because `renv` does not
run while this book renders, conveys the structure.

```json
{
  "R": {
    "Version": "4.6.0",
    "Repositories": [
      {
        "Name": "CRAN",
        "URL": "https://packagemanager.posit.co/cran/2026-06-29"
      }
    ]
  },
  "Packages": {
    "dplyr": {
      "Package": "dplyr",
      "Version": "1.2.1",
      "Source": "Repository",
      "Repository": "CRAN",
      "Hash": "5d1c2e3a...",
      "Requirements": [
        "R6", "cli", "generics", "glue",
        "lifecycle", "magrittr", "pillar",
        "rlang", "tibble", "tidyselect", "vctrs"
      ]
    }
  }
}
```

Three features deserve comment. First, the version is
recorded exactly, `1.2.1` and not `>= 1.2`, so there is
no latitude for a later install to drift upward.
Second, the source is named. A package that came from
CRAN, from a Posit Package Manager mirror, from
Bioconductor, or from a specific GitHub commit is
restored from that same source. This matters because
the same version number can denote different code on
different repositories. The `Repository` value is the
registered name of the repository the package was drawn
from, matching one of the names in the `Repositories`
block above. This is why the excerpt records `CRAN`
even though that entry points at a Posit mirror. Third, the content
hash lets `renv` confirm that a restored package is
byte-identical to the one recorded, catching the rare
but real case of a repository serving altered contents
under an unchanged version string. The `Requirements`
field lists each package's own dependencies, so the
lockfile captures the full transitive closure, not
merely the packages the analyst named.

### The init, snapshot, restore workflow

Three commands drive the ordinary lifecycle. Because
none of them can execute while this book renders, we
show them as static R blocks; the reader runs them at
an interactive prompt inside the project.

Initialization creates the project-local library and a
first lockfile.

```r
renv::init()
```

From this point, packages installed into the project go
into its private library rather than the shared
system-wide one, so one project's `dplyr` upgrade
cannot disturb another's. As the analysis develops and
new packages are installed, a snapshot writes the
current state into the lockfile.

```r
renv::snapshot()
```

The snapshot does not blindly record every installed
package. It first scans the project's source for the
packages the code actually references, through
`library()`, `require()`, and `pkg::fun()` calls, and
records those together with their transitive
dependencies. A package sitting in the library but
called nowhere in the code is, by default, left out, a
behavior that keeps the lockfile a record of what the
analysis uses rather than of everything that happens to
be installed.

On another machine, or the same machine a year later, a
restore reads the lockfile and rebuilds the library to
match.

```r
renv::restore()
```

This is the operation that redeems the pin. A
collaborator who clones the repository and calls
`renv::restore()` obtains the exact package set the
analysis was written against, irrespective of what
versions are current. Snapshot is capture; restore is
the later consumption of what was captured.

In the framework this book uses, this lifecycle is
partly automated so that an analyst does not forget the
snapshot. The project's generated `.Rprofile` installs
an exit hook that snapshots the library when an
interactive session ends, and restores missing packages
when a session begins. A package installed during
analysis is thus captured rather than lost. The automation
does not change the model; it merely removes the two
most common ways the library and the lockfile drift
apart.

::: {.callout-note icon=false collapse=true}
## Check your understanding: snapshot versus restore

A colleague clones your project, opens R, and is
puzzled that the packages are not yet installed. They
ask whether they should run `renv::snapshot()` to get
them. What do you advise, and why?

::: {.callout-note icon=false collapse=true}
## Answer

They should run `renv::restore()`, not
`renv::snapshot()`. Restore reads the existing lockfile
and installs the recorded versions into the local
library, which is exactly what a fresh clone needs.
Snapshot runs in the opposite direction: it would
overwrite the lockfile with whatever is currently
installed, which on a fresh clone is nothing relevant,
and would thereby destroy the very record the colleague
needs. The mnemonic is that snapshot writes the
lockfile and restore reads it.
:::
:::

## The dependency-triad invariant

A lockfile solves the pinning problem, but it
introduces a bookkeeping problem in its place: the
lockfile can fall out of step with the code it is
supposed to describe. A contributor installs a package
for an exploratory notebook and forgets to snapshot, so
the code now uses a package the lockfile does not
record. Or a package is dropped from the analysis but
left in the project's `DESCRIPTION`, so the declared
dependencies name something the code no longer touches.
Either drift silently degrades the pin.

The framework this book uses formalizes the correct
relationship as the dependency triad. Three sources of
information describe the packages an analysis uses, and
in a healthy compendium they satisfy a nesting relation.

- The **code** references some set of packages, through
  its `library()`, `require()`, and `pkg::fun()` calls.
- The **`DESCRIPTION`** declares a set of direct
  dependencies.
- The **`renv.lock`** records a set of pinned packages.

The invariant is the composition of two atomic
inclusions, each set contained in the next:

$$
\mathrm{Code} \subseteq \mathrm{DESCRIPTION}
\subseteq \mathrm{renv.lock}.
$$

Read left to right, it says that every package the code
calls must be declared in `DESCRIPTION`, and every
declared dependency must be pinned in the lockfile. The
lockfile is the largest set because it additionally
records the transitive dependencies that the analyst
never named directly. Base and recommended packages,
which every R installation provides, are excluded from
the code set, so the first inclusion concerns only
contributed packages.

One qualification keeps the invariant honest. Its strict
form, with `DESCRIPTION` as the binding middle term,
holds cleanly for the R-package archetype, where
`DESCRIPTION` is the package's declared dependency
interface and every package the code calls belongs there
by definition. For a research or analysis compendium the
middle term is looser: `DESCRIPTION` declares only the
API of the reusable code in `R/`, while packages used
solely by analysis scripts or reports legitimately live
in `renv.lock` without appearing in `DESCRIPTION`. The
operative invariant for a compendium is therefore the
weaker $\mathrm{Code} \subseteq \mathrm{renv.lock}$, so
the strict `DESCRIPTION` inclusion is exact for the
package archetype but archetype-conditional in general.
As a matter of placement an analysis-script dependency
may still be listed in `DESCRIPTION` under `Suggests`. But
the framework does not enforce the strict `DESCRIPTION`
inclusion as a gate on a compendium, precisely because
that inclusion does not hold for one. The gate it enforces
on a compendium is the single inclusion $\mathrm{Code}
\subseteq \mathrm{renv.lock}$, that nothing the code uses
goes unpinned. The strict double inclusion is applied only
in the package archetype, where `DESCRIPTION` is the
authoritative dependency interface.

@fig-triad draws the nesting, with the archetype
qualification marked rather than hidden. The regions are
what matter: a package may sit legitimately in any of the
three bands, and the only genuinely broken position is
outside the lockfile altogether.

```{r}
#| label: fig-triad
#| fig-cap: 'The dependency triad as nested sets, with example packages placed in each region. Everything the code calls must be pinned, so a package outside the lockfile is the one broken position and is marked with a cross. The band between `DESCRIPTION` and the lockfile holds the transitive dependencies no analyst names directly, together with, in a compendium, the packages used only by analysis scripts. That second case is the archetype qualification discussed above: it is a violation in an R package and legitimate in a compendium.'
#| fig-alt: 'Three nested rectangles. The outermost is labeled renv.lock, everything pinned; inside it a rectangle labeled DESCRIPTION, declared dependencies; innermost a rectangle labeled code, packages actually called. Example packages are placed in each band: dplyr inside the code rectangle, an unused declared package between DESCRIPTION and code, and transitive dependencies such as rlang and vctrs plus an analysis-script-only package in the outer band. A package named survival sits outside all three rectangles, marked with a cross and labeled used but unpinned, the one broken position.'
#| fig-width: 6.6
#| fig-height: 4.2

library(ggplot2)
library(tibble)

bands <- tibble(
  label = c('renv.lock: everything pinned',
            'DESCRIPTION: declared',
            'Code: actually called'),
  xmin = c(0.4, 1.5, 2.5), xmax = c(9.6, 8.5, 7.5),
  ymin = c(0.5, 1.3, 2.1), ymax = c(6.2, 5.4, 4.6),
  fill = c('#eef4f4', '#dcebea', '#c3ddda')
)

items <- tibble(
  x = c(5.0, 5.0, 2.6, 7.4),
  y = c(3.3, 1.75, 0.92, 0.92),
  lab = c('dplyr\n(called, declared, pinned)',
          'an unused declaration',
          'rlang, vctrs\n(transitive)',
          'ggplot2, used only by\nanalysis scripts')
)

ggplot() +
  geom_rect(data = bands,
            aes(xmin = xmin, xmax = xmax,
                ymin = ymin, ymax = ymax, fill = fill),
            color = 'grey40', linewidth = 0.4) +
  scale_fill_identity() +
  geom_text(data = bands, aes(x = xmin + 0.15, y = ymax - 0.25,
                              label = label),
            hjust = 0, fontface = 'bold', size = 3.1) +
  geom_text(data = items, aes(x = x, y = y, label = lab),
            size = 2.9, lineheight = 0.95, color = 'grey20') +
  annotate('point', x = 3.55, y = 0.02, shape = 4,
           size = 2.6, stroke = 1.1, color = '#a33') +
  annotate('text', x = 3.75, y = 0.02, hjust = 0,
           label = 'survival: used but unpinned',
           size = 3.1, fontface = 'bold', color = '#a33') +
  coord_cartesian(xlim = c(0.2, 9.8), ylim = c(-0.35, 6.4)) +
  theme_void()
```

The invariant matters because it is checkable by
machine. The framework's `check-renv` gate, backed by
the `zzrenvcheck` package, scans all three sources,
tests the two inclusions, and reports any violation: a
package used in code but undeclared, or declared but
unpinned. Crucially it can repair many violations
automatically. It writes a missing declaration into
`DESCRIPTION` through the `desc` package, and for a
package absent from the lockfile it queries the CRAN API
for the current version and injects a resolved entry
into `renv.lock` directly. It does not call
`renv::snapshot()` or `install.packages()`; the repair
edits the two manifests as text rather than installing
anything or re-deriving the lockfile from the library.
The analyst's ordinary workflow is thus to write the
analysis and let the check reconcile the bookkeeping.

```bash
$ make check-renv
```

This runs the check with automatic repair before a
commit. The same check runs again under continuous
integration, but there in a report-only mode: it runs
strict and with repair disabled. A drift that escaped
the developer's machine therefore fails the build rather
than being silently repaired. This forces the contributor
to commit the fix, instead of letting the clean build
paper over it. It is worth being
precise about what the gate does and does not verify.
It is a presence check: it guarantees that the three
sets nest, that nothing used is unpinned. It does not
verify that the version floors in `DESCRIPTION` agree
with the exact pins in the lockfile, which is a separate
concern.

### Scanning the documentation, not only the code

A dependency check is only as good as its reach, and the
most common way for one to fail is subtle. It scans the
analysis code, pronounces the project consistent, and
overlooks a package that is used only in a vignette or in
the rendered manuscript. `zzrenvcheck` is deliberately
more expansive than a plain code scan on this point, and
the reason is worth understanding, because it is exactly
where a documentation build quietly breaks on a fresh
machine.

The scan does not look at `.R` scripts alone. It parses
the literate and dynamic documents as well, R Markdown
(`.Rmd`), Quarto (`.qmd`), and Sweave (`.Rnw`), so that a
package loaded inside a report chunk or a vignette is seen
just as one loaded in `R/` is. Within every such file it
recognizes each of the forms by which R code reaches for a
package: the explicit `library(pkg)` and `require(pkg)`,
the namespaced `pkg::function()`, and the roxygen2
directives `@import` and `@importFrom` from which a
package's `.Rd` help files are generated. In its strict
mode it widens the search further, into the `tests/`,
`vignettes/`, and `inst/` directories rather than the
analysis directories alone. It is correspondingly careful
about what it ignores: `README` files and the example
directories (`man/examples`, `inst/examples`) are skipped,
since the install-and-demonstrate code they usually carry
would otherwise raise false alarms. A separate pass reads
the reproducibility files themselves, the `Dockerfile`,
the `Makefile`, `install.sh`, and `.Rprofile`, for
version-pinned installs such as `pak('survival@3.5-8')`,
catching a version fixed in the container recipe that has
drifted from the lockfile.

The payoff is precisely the failure a code-only audit
misses. Consider a manuscript compendium whose analysis in
`R/` restores cleanly from the lockfile, but whose report
draws a forest plot with a plotting package that appears
nowhere except the `.qmd` itself. A scanner that read only
the code would certify the project as consistent, and the
report would then fail to render on the reviewer's machine
for want of an unpinned package. Because `zzrenvcheck`
reads the document, the missing dependency is caught and
pinned before it can break the build. This reach matters
most for the archetypes whose dependencies live in their
documents rather than their code, the scholarly manuscript
and the blog. Their rendered outputs are the point, and
their package needs are therefore concentrated where a
naive check does not look.

::: {.callout-important}
## The honest level

Passing `check-renv` confirms that the package layer is
internally consistent, not that the analysis is
reproducible in any fuller sense. The triad says the
packages the code uses are all pinned; it says nothing
about the R version, the operating system, or the
system libraries beneath them. An analysis that passes
the dependency check has reached L1 and no higher, and
should be described as package-pinned, not as
reproducible without qualification.
:::

### The triad across archetypes

The lockfile discipline is identical across the
archetypes of @sec-compendium, because every archetype
records its packages in the same `renv.lock` by the
same snapshot and restore. What varies slightly is how
natural the declared-dependency middle term feels. In
the R-package archetype, `DESCRIPTION` is not an
optional extra but the package's defining interface:
`R CMD check` reads it, an installer reads it, and CRAN
reads it. The triad is therefore especially natural
there, since the declared set already exists for
reasons independent of reproducibility. A data-analysis
project, a manuscript compendium, a simulation study,
and a blog also carry a `DESCRIPTION`, but there it
declares only the API of the reusable code in `R/`.
Their binding invariant is instead the single
$\mathrm{Code} \subseteq \mathrm{renv.lock}$. The
R-package case is the one where the declared middle
term would exist even if reproducibility were not a
concern. A subtlety follows for the R-package
archetype: a package used only by an analysis script or
report belongs in `Suggests` rather than `Imports`, so
that `R CMD check` does not hard-require a package the
package's own exported functions never call.

::: {.callout-note icon=false collapse=true}
## Check your understanding: locating a drift

Continuous integration on an R-package archetype fails
with the message that `survival` is used in `R/model.R`
but is not declared in `DESCRIPTION`. Which inclusion of
the triad has been violated, and what single command
would most likely repair it?

::: {.callout-note icon=false collapse=true}
## Answer

The invariant is the composition of two atomic
inclusions, and the one violated here is the first,
$\mathrm{Code} \subseteq \mathrm{DESCRIPTION}$: the code
references `survival` while `DESCRIPTION` does not
declare it, which in turn leaves it unpinned in the
lockfile. The transitive composition
$\mathrm{Code} \subseteq \mathrm{renv.lock}$ therefore
fails as a consequence, but the atomic breach is at the
code-to-`DESCRIPTION` step. The likely cause is that
`survival` was installed interactively but never
declared. Running `make check-renv`, which invokes
`zzrenvcheck::check_packages()` with automatic repair,
would write `survival` into `DESCRIPTION` through the
`desc` package and inject a CRAN-resolved entry for it
directly into `renv.lock`, restoring both inclusions.
This edits the two manifests as text and is distinct
from an `renv::snapshot()`, which would instead re-derive
the lockfile from the installed library. The fix must
then be committed so the clean CI build sees it.
:::
:::

## Dated snapshots of a package repository

A lockfile records which versions to install. But a
plain CRAN mirror will happily serve only whatever is
current, and older source packages must often be
recompiled, which is slow and can fail when a system
library has since moved on. The Posit Package Manager
resolves both problems by offering dated snapshots of
CRAN. A snapshot URL freezes the repository as it stood
on one calendar day, so that a request for a package
returns the version that was current on that date, and
returns it as a pre-compiled binary rather than as
source to be built.

The mechanism is a URL that embeds a date.

```r
options(repos = c(
  CRAN = "https://packagemanager.posit.co/cran/2026-06-29"
))
```

With repositories pointed at a dated snapshot, an
install run today and the same install run next year
draw from an identical, frozen catalog. The
lockfile above records exactly this URL in its
`Repositories` block, so that a restore reproduces not
only the versions but the source they came from. The
gain is twofold: fixity, because the catalog cannot
drift, and speed, because binaries install in seconds
where source builds take minutes. On Linux the binary
path is more specific: pre-compiled binaries are served
only from the distribution-codename form of the URL,
such as
`https://packagemanager.posit.co/cran/__linux__/noble/2026-06-29`,
whereas the plain dated path shown above delivers source
to be built. In the framework this
book uses, the project's generated configuration points
R at a dated snapshot automatically, so that the two
benefits accrue without the analyst assembling the URL
by hand. The date of the snapshot becomes, in effect,
part of the environment's identity, and a topic we take
up again when the container fixes a snapshot date at
image-build time in @sec-containers.

## The limits of renv

It is essential to be precise about what a lockfile does
not do, because the temptation to overclaim is strong
and the gap it leaves is exactly the subject of the next
chapter. `renv` pins the R packages. It does not pin the
R interpreter itself, so a lockfile produced under R
4.6.0 may restore under R 4.2.1 with different results,
or fail to restore at all when a recorded version
requires a newer R. It does not pin the operating
system, so a lockfile assembled on macOS carries no
guarantee on Linux. And it does not pin the system
libraries that many packages link against, the GDAL
behind a spatial package or the `libxml2` behind an XML
parser. A restore on a machine lacking a required
system library may therefore fail to compile a package
or, more insidiously, may install a differently
behaving one.

The point is not a criticism of `renv`, which does its
one job well. It is that the package layer is one layer
of a taller stack, and pinning it leaves every layer
beneath it as the next machine happens to provide.
@ushey2020 designed `renv` to solve the package problem
and no more; the environment problem is out of its scope
by design. An analysis equipped with a faithful lockfile
has reached L1, pinned packages, and is exposed at L2 and
below. Closing that remaining gap is the work of the
container, and it is where @sec-containers begins.

## Lighter and alternative approaches

The lockfile-and-restore idiom is not the only way to
pin the package layer, and a reader should know the
principal alternatives, because some suit particular
settings better and because the tradeoffs illuminate
what `renv` is doing.

The most important shift of idiom is from preserving a
lockfile to reconstructing an environment from a date.
The `rang` package of @chan2023 takes a set of packages
and a target date, and resolves the versions that were
current then. It reaches back through CRAN's archive to
reconstruct a historical environment, even for a project
that never recorded a lockfile at all. Where `renv`
requires foresight, a committed lockfile written while
the analysis was live, `rang` offers hindsight,
recovering an approximate environment after the fact.
It is the natural instrument for a legacy analysis that
one wishes to resurrect.

The `groundhog` package of @simonsohn2026 pursues the
same date-based idea but at a different point in the
workflow. Rather than restoring a library before the
script runs, it loads version-specific packages at run
time: a `groundhog.library()` call carries a date, and
the packages current on that date are loaded for that
session. It needs no project library and no snapshot
step, trading `renv`'s explicit committed lockfile for a
date-stamped call evaluated against the repository's
history. The convenience is real, and so is the cost:
the pin now lives inside the script's control flow
rather than in a separate, inspectable manifest, which a
reviewer must read the code to discover.

Finally, `capsule` [@capsule] is best understood as a
lazier discipline layered over `renv` itself. It defers
building the project library until it is actually needed,
letting the analyst develop against their ordinary
user library and materialize the pinned library only
when reproducing or checking the work. For analysts who
find `renv`'s always-on project library intrusive during
exploratory work, it lowers the friction of adoption
while preserving the lockfile as the eventual source of
truth.

Across all of these the invariant lesson holds. Each
pins the package layer and only the package layer;
whether the pin is a committed lockfile, a reconstructed
date, or a run-time call, none reaches down to the R
version, the operating system, or the system libraries.
The choice among them is a choice of ergonomics and of
foresight versus hindsight, not a choice of how far down
the determinant stack one has pinned.

## Worked example

To make the package-churn failure mode concrete, and to
generate a computed artifact at build time, we
construct a small synthetic lockfile diff: a comparison
of two snapshots of the same project taken some months
apart. This is the kind of table a version-control diff
of `renv.lock` would summarize, and reading it is the
routine skill of noticing which pins moved. The data
are synthetic and no packages are installed; the
computation uses only `tibble`, `dplyr`, and `knitr`.

```{r}
#| label: tbl-lockdiff
#| tbl-cap: 'A synthetic lockfile diff between two snapshots of one project. The changed column flags every package whose pinned version moved between the earlier and later snapshot.'

library(tibble)
library(dplyr)

lock_diff <- tibble(
  package    = c('dplyr', 'ggplot2', 'tidyr',
                 'survival', 'lubridate', 'rlang'),
  version_a  = c('1.1.0', '3.4.4', '1.3.0',
                 '3.5-7', '1.9.3', '1.1.2'),
  version_b  = c('1.2.1', '3.4.4', '1.3.1',
                 '3.5-8', '1.9.3', '1.1.4')
) |>
  mutate(changed = if_else(
    version_a == version_b, '', 'changed'
  ))

knitr::kable(lock_diff)
```

Four of the six pins moved between the two snapshots.
Three of the moves are patch-level (`tidyr`, `survival`,
and `rlang`), the kind that ordinarily preserve
behavior but occasionally do not. One is a minor
version bump (`dplyr`, from 1.1.0 to 1.2.1) of the sort
that can refine a default. The `rlang` move, from 1.1.2
to 1.1.4, is likewise patch-level. It is singled out
here not for the distance it traveled but for its
transitive position: a package almost nothing calls
directly, yet almost everything depends on. It thus moved
beneath the whole stack. The lesson is that a diff
of the lockfile is a reviewable record of exactly this
churn. Without a committed lockfile, none of these
movements would be visible, and the analyst confronted
with a changed result would have no way to see which
package had shifted. The lockfile does not prevent
packages from changing; it makes the change legible, and
lets the analyst restore the earlier column deliberately
rather than discover the drift by accident.

## Collaborating with an LLM

A language model is a capable assistant for the
mechanics of `renv`, but its confident fluency is
precisely the hazard, because a plausible-looking
lockfile instruction can pin the wrong thing or nothing
at all. Each of the following triples pairs a useful
prompt with what the analyst must watch for and verify.

**Prompt.** 'My collaborator cloned our project and the
packages are missing. Write the R command they should
run.'

*Watch for.* A model may suggest reinstalling packages
by hand with `install.packages()`, or may propose
`renv::snapshot()`, either of which is wrong for a fresh
clone. The correct answer is `renv::restore()`.

*Verification.* Confirm that the suggested command reads
the lockfile rather than writing it, and that after
running it the installed versions match the lockfile.
Run the check gate and confirm the triad holds.

**Prompt.** 'Explain why my `renv.lock` includes
`roxygen2` and `testthat` when my analysis never calls
them, and whether I should remove them.'

*Watch for.* A model may assert flatly that any unused
package should be purged, missing the deliberate
convention by which test infrastructure is pinned while
pure development tooling is not. The distinction is a
judgment, not a rule the model can read off the file.

*Verification.* Decide the placement yourself against
the two-test procedure: does the package's version
affect the result, and does the code reference it?
Confirm that development-only tools such as
`languageserver` are in the Dockerfile and absent from
the lockfile, and that test packages are handled by your
stated convention.

**Prompt.** 'Set my repository to a dated Posit Package
Manager snapshot for 2026-06-29.'

*Watch for.* A model may produce a URL with the wrong
path structure, an implausible date, or a mirror that
does not offer binaries for your platform. It may also
silently assume a Linux distribution codename that does
not match your system.

*Verification.* Check the constructed URL against the
Posit Package Manager documentation, confirm the date is
the one you intend, and confirm that an install actually
draws binaries rather than falling back to source. The
recorded URL should then appear in the lockfile's
`Repositories` block.

## Exercises

1. Initialize `renv` in a throwaway project, install two
   packages, and snapshot. Open `renv.lock` in a text
   editor and locate, for one package, its version,
   source, and hash. Explain in a sentence what each of
   the three would let a later restore verify.

2. With the project of Exercise 1, upgrade one package
   to a newer version, snapshot again, and inspect the
   diff of `renv.lock` under version control. Which
   fields changed, and which stayed fixed? Relate the
   diff to the synthetic table of @tbl-lockdiff.

3. Deliberately break the dependency triad: add a
   `library()` call for a package you have not declared
   or pinned. Run the project's dependency check and
   record which inclusion it reports as violated and how
   it repairs the violation. Then remove the call and
   confirm the check passes.

4. Point your repository at a dated Posit Package Manager
   snapshot and reinstall one package. Time the install,
   then compare it with an install of the same package
   from source. Report the difference and explain, in
   terms of binaries versus source, why it arises.

5. Take a small analysis that has no lockfile and use
   `rang` to reconstruct the package environment as it
   stood on a chosen past date. Compare the resulting
   version set with what a current `install.packages()`
   would provide. Which packages differ, and by how much?

6. Argue, in a short paragraph, why an analysis that
   passes its dependency check has reached L1 and not
   L2. Name one concrete determinant that a lockfile
   leaves unpinned and describe a plausible way it could
   change a result across two machines.

## Solutions

Attempt an exercise before expanding its solution. Two of
the six have no single correct answer, because what they
report depends on the machine and the date; for those the
box says what a good answer establishes rather than
supplying a number.

::: {.callout-note icon=false collapse=true}
## Exercise 1

The three fields answer three different questions a later
restore has to settle.

The **version** is recorded exactly, `1.2.1` and not
`>= 1.2`, so a restore installs the release the analysis
actually used and has no latitude to drift upward. This
is the field that defends against the ordinary failure,
in which a package acquires a new default between the
original run and the rerun.

The **source** names where the package came from: CRAN, a
Posit Package Manager mirror, Bioconductor, or a specific
GitHub commit. It matters because the same version number
can denote different code on different repositories, so
the version alone does not identify the package. Note
that the `Repository` value is the registered name of the
repository rather than its URL, which is why a lockfile
can record `CRAN` while the corresponding entry in the
`Repositories` block points at a Posit mirror.

The **hash** lets `renv` confirm that what it restored is
byte-identical to what was recorded. It is the only one
of the three that detects a repository serving altered
contents under an unchanged version string. That case is
rare, but it is the one the other two fields cannot see.

A useful way to hold the three together: the version says
which release, the source says whose copy of it, and the
hash says whether the bytes actually match.
:::

::: {.callout-note icon=false collapse=true}
## Exercise 2

For the upgraded package, `Version` and `Hash` both
change, and they change together. `Package`, `Source`,
and `Repository` stay fixed, since the package is the
same package drawn from the same place. The
`Requirements` field may or may not change, depending on
whether the new release altered its own dependencies.

Two further effects are worth noticing in the diff, and
are easy to miss on a first reading. If the upgraded
package gained a dependency, new package entries appear
in the lockfile that you never installed by name, because
the lockfile captures the full transitive closure rather
than the packages you asked for. And the `R` block at the
top of the file is untouched, which is the point of
@tbl-lockdiff: the changed column flags packages, and the
lockfile pins the R version separately from them.

The relation to @tbl-lockdiff is that the table shows the
same structure in synthetic form. Each row is a package,
the earlier and later columns hold the pinned versions,
and the flag marks where they differ. A real
`git diff renv.lock` is that table expressed as JSON, with
the additional detail that the hash moves whenever the
version does.
:::

::: {.callout-note icon=false collapse=true}
## Exercise 3

The violated inclusion is the first atomic one,
$\mathrm{Code} \subseteq \mathrm{DESCRIPTION}$. The code
now references a package that `DESCRIPTION` does not
declare, which in turn leaves it unpinned in the
lockfile, so the transitive composition
$\mathrm{Code} \subseteq \mathrm{renv.lock}$ fails as a
consequence. The breach to report is the atomic one, not
the consequence.

Running `make check-renv` invokes
`zzrenvcheck::check_packages()` with automatic repair. It
writes the package into `DESCRIPTION` through the `desc`
package and injects a CRAN-resolved entry for it directly
into `renv.lock`, restoring both inclusions. This edits
the two manifests as text and is deliberately distinct
from an `renv::snapshot()`, which would instead re-derive
the whole lockfile from whatever is currently installed.

One observation the exercise is designed to provoke, and
which is worth stating because it surprises people.
Removing the `library()` call and re-running the check
does make the check pass, but it does not return the
manifests to their prior state: the repair already wrote
the package into `DESCRIPTION` and the lockfile, and
nothing removes it. A passing check means the inclusions
hold, not that the project is unchanged. Restoring the
prior state requires deleting the entries by hand or
discarding the changes under version control, which is
itself a small argument for making the repair on a branch.
:::

::: {.callout-note icon=false collapse=true}
## Exercise 4

This exercise has no single correct answer, because the
times depend on the package, the machine, and the network.
What a good answer establishes is the mechanism, and it
should report both timings and name the package.

The expected result is that the snapshot install is
substantially faster, and the explanation is that Posit
Package Manager serves a prebuilt binary for the declared
platform. Installing a binary is unpacking an archive;
installing from source is compiling it. For a package
carrying compiled code, typically anything depending on
`Rcpp`, the difference is commonly an order of magnitude
or more.

The instructive case is the one that does not show a large
difference. A pure-R package has little to compile, so the
two paths do similar work and the timings converge. A
reader who happens to pick such a package and reports a
small difference has the right finding, provided they
identify why. Two other conditions can flatten the
difference and are worth checking before concluding
anything: the `renv` cache may already hold the package,
in which case neither install does real work, and PPM does
not serve binaries for every platform, so a machine whose
platform is unserved falls back to source in both arms.
:::

::: {.callout-note icon=false collapse=true}
## Exercise 5

This exercise also has no single correct answer, since
what differs depends on the date chosen and on the size of
the dependency tree. A good answer reports the count of
differing packages, gives one or two concrete version
pairs, and draws the general conclusion.

The general conclusion is that `rang` reconstructs what
was *current on a date*, which is not the same thing as
what an analysis *actually used*. The two coincide only if
the original analyst installed everything on that date and
never updated anything afterward, which is not how most
projects proceed. Reconstruction by date is therefore an
approximation, and a lockfile is a record; the
approximation is what one falls back on when the record
does not exist.

The number of differing packages will generally grow with
the age of the date and with the depth of the dependency
tree, since a difference anywhere in the transitive
closure counts. It is worth noting that a package whose
version happens to be unchanged is not evidence that the
reconstruction succeeded, only that that particular
package was quiet over the interval.
:::

::: {.callout-note icon=false collapse=true}
## Exercise 6

A passing dependency check establishes that the triad
holds: every package the code uses is declared, and every
declared package is pinned in the lockfile. That is
exactly the L1 property, pinned packages, and the check is
the evidence for it.

It is not L2, because the lockfile pins packages and
nothing beneath them. The R version is recorded in the
lockfile but not supplied by it; the operating system, the
system libraries, the locale, the time zone, and the
numerical libraries are all outside its scope. L2 requires
an artifact that carries those, which is what
@sec-containers is about.

For a concrete unpinned determinant, the numerical-library
layer is the cleanest example. Two machines can hold
identical pinned package sets and still link against
different BLAS implementations, reference BLAS on one and
Apple's Accelerate or Intel MKL on the other. Because
these reorder the floating-point summations inside a
matrix decomposition, a principal-components analysis or a
mixed-model fit can return results that agree to several
decimal places and differ in the last, and a sign flip on
a component can make the difference visible in a plot. The
lockfile is silent on all of it, and setting a seed does
not help, because the difference is not stochastic.

A second acceptable answer names the locale. Two machines
under different collation orders sort character vectors
differently, which changes which row is taken first in a
tie and can therefore change a reported result through a
`slice()` or a `first()`, with no error raised anywhere.
:::

## Further reading

The authoritative reference for the tool is @ushey2020,
the package documentation for `renv`, whose introductory
and collaboration vignettes are the fastest route to
fluent daily use and whose treatment of the cache and of
custom repositories repays study once the basics are in
hand. The relationship of `renv` to its predecessor
`packrat`, and its place in the wider stack, is set out
compactly in the survey portions of @sec-intro.

For the date-based alternatives, @chan2023 presents
`rang` and its strategy of reconstructing a historical
environment from CRAN's archive, with a clear account of
the cases where hindsight succeeds and where it cannot;
@simonsohn2026 documents `groundhog` and its run-time,
date-stamped loading model; and @capsule describes the
lazy discipline that `capsule` layers over `renv`.
Reading the three together clarifies that pinning the
package layer admits several idioms with different
demands on foresight.

On the discipline of where a package belongs, the
placement whitepaper accompanying the framework develops
the two-test decision procedure and the two-layer model
that this chapter summarizes, and is the place to turn
when a package sits awkwardly between the lockfile and
the Dockerfile. Finally, the boundary this chapter
insists on, that the lockfile pins the packages and
nothing beneath them, is taken up directly in
@sec-containers, where the container fixes the R version,
the operating system, and the system libraries that
`renv` by design does not.
