# Pinning the Environment with Docker {#sec-containers}

::: {.bookepigraph}
```{=latex}
\begin{flushright}\small
```
*Though computational reproducibility seems more straight
forward than replicating physical experiments, the complex
and rapidly changing nature of computer environments makes
being able to reproduce and extend such work a serious
challenge.*

[Carl Boettiger]{.smallcaps}, *An Introduction to Docker
for Reproducible Research* (2015)
```{=latex}
\end{flushright}
```
:::

## Learning objectives

By the end of this chapter the reader should be able
to:

- Explain which determinants of a computational
  result a package lockfile leaves unpinned, and why
  a container is required to fix them.
- Distinguish a Docker image from a container, and
  explain image layers and the precise sense in which
  an image 'runs the same everywhere it runs'.
- Pin a base image by its immutable content digest and
  point its in-image installer at a dated repository
  snapshot, and say what each step secures.
- Set the environment variables that govern locale,
  time zone, and thread determinism, and predict a
  concrete failure that each one prevents.
- Select a Rocker profile appropriate to a given
  research archetype, from a minimal base to a
  publishing base carrying a LaTeX toolchain.
- State honestly which rung of the reproducibility
  ladder a given image reaches, and which it does not.

## Orientation

Consider an epidemiologist who has done everything
the previous chapter asked. Her cohort analysis is
under version control, and every R package it uses is
fixed to an exact version in a lockfile. She has
reached L1, pinned packages, and she is entitled to
some confidence. She sends the project to a
collaborator, who runs `renv::restore()` and watches
it fail. One of her packages links against the
geospatial library GDAL, which is present on her
machine and absent on his, and the package will not
install without it. On a third machine the restore
succeeds, but the analysis returns a subtly different
figure, because that machine runs an older release of
R in which a numerical default differs. The lockfile
did its job perfectly. What defeated the collaborator
was everything the lockfile does not describe.

The lockfile pins the R packages and nothing beneath
them. It says nothing about the operating system, the
system libraries such as GDAL or `libxml2` that many
packages link against, or the R interpreter itself.
As Ushey observed in documenting `renv`, the tool
records the packages an analysis used, not the
platform it ran on [@ushey2020]. A lockfile restored
under a different R version, or on an operating system
missing a required system library, may fail to
install or, more insidiously, may install and behave
differently. This is the gap the present chapter
closes. Its subject is the container, the technology
that fixes the operating system, the system
libraries, and the runtime, and so carries an
analysis from L1 to L2, pinned environment.

We shall build the container from first principles,
assuming no prior contact with it, because the health
scientist who most needs reproducibility is rarely the
one with a systems-administration background. We begin
with what a container is, proceed to the Rocker
project that supplies R-specific containers, and then
to the two refinements that make a container itself
reproducible rather than merely convenient. We close
with the profiles, a short menu of base environments,
and a first look at how a team shares one.

## The analyst's contribution

A container is a tool, and like every tool in this
book it automates the mechanical part of the work and
leaves the judgment to the analyst. Three decisions
belong to the analyst and to no script.

1. **The choice of base environment must match the
   archetype, not the maximum.** A container can carry
   anything, and it is tempting to reach for the
   largest base image that surely contains whatever
   one might need. That instinct is wrong. A manuscript
   archetype needs a base carrying a LaTeX toolchain;
   an R-package archetype needs only R and the check
   tools; a simulation archetype may want more cores
   than it wants software. Provisioning far above the
   need inflates the image, slows every rebuild, and
   obscures what the analysis actually depends on. The
   analyst chooses the smallest base that carries the
   real requirement.

2. **The decision of how far to pin is a decision
   about honesty, not convenience.** A container built
   on a mutable tag such as `latest`, installing
   whatever the package repository serves on the day
   of the build, looks reproducible and is not. Two
   builds from the same recipe on different days
   produce different images. Fixing this costs a little
   effort now, in pinning the base by digest and the
   installer by date, and the analyst must decide to
   spend it. The chapter will argue that the effort is
   the difference between claiming L2 and reaching it.

3. **The environment variables that govern
   determinism are the analyst's to set.** A container
   inherits a locale, a time zone, and a default
   thread count, and these silently shape sorting,
   date arithmetic, and the last bits of a floating
   point sum. No tool infers what the analysis needs;
   the analyst declares it, in a few lines that a
   later reader can inspect and trust.

::: {.callout-note}
## The vocabulary of this chapter

- **Image.** A frozen, read-only filesystem together
  with metadata, bundling an operating system, R, a
  set of packages, and configuration. It is analogous
  to a snapshot of a fully installed computer, packaged
  so it can be shipped.
- **Container.** A running instance of an image. One
  image can start many containers; each begins from the
  same frozen filesystem.
- **Layer.** Each instruction in an image's recipe
  produces one filesystem layer. An image is the stack
  of its layers, and identical layers are shared and
  cached across images.
- **Base image.** The starting image a recipe builds
  on, named on the `FROM` line. Here, a Rocker image
  providing a fixed R version on a fixed operating
  system.
- **Tag and digest.** A tag, such as
  `rocker/tidyverse:4.4.0`, is a human-readable pointer
  that may later be repointed. A digest,
  `@sha256:...`, is an immutable hash of the image's
  exact content. The same digest always denotes
  byte-identical bytes.
- **Rocker.** The community project that publishes the
  standard R Docker images.
- **Dated snapshot.** A frozen view of a package
  repository on a particular day, so that installing
  'the current version' resolves to the same version
  on every build.
- **Profile.** The base-image bundle a project is built
  from, such as `minimal`, `tidyverse`, `rstudio`, or
  `publishing`. It selects the `FROM` image.
:::

## What the lockfile leaves standing

It is worth stating precisely what a lockfile fixes,
because the boundary is the whole motivation for the
container. A lockfile records, for every R package an
analysis loaded, the exact version and the repository
it came from, and `renv::restore()` reinstalls that
precise set [@ushey2020]. That is the L0-to-L1
transition, and it is genuine progress. Beneath the
packages, however, sits a stack of determinants the
lockfile never touches.

Three of these matter most. The first is the operating
system, which supplies the C library and the
conventions for locale and sorting against which every
package was compiled. The second is the layer of
system libraries, the compiled non-R software that
packages link against: GDAL for spatial data, `libxml2`
for XML parsing, the numerical libraries that implement
linear algebra. A package that needs GDAL will not
install on a machine that lacks it, and the lockfile
cannot supply it, because GDAL is not an R package. The
third is the R interpreter itself. A lockfile pins the
packages but not the R version, and, as we saw in the
introduction, a change between R releases can alter a
result silently: the random-number default that changed
in R 3.6.0, the `stringsAsFactors` default that flipped
in R 4.0.0. Restore the same lockfile under a different
R, and the packages match while their behavior does
not.

The lesson is that `renv` is necessary and, by itself,
insufficient. It closes the package layer cleanly and
leaves the three layers beneath it exposed. To fix
those, we need a tool that packages not just the
packages but the whole computer they ran on. That tool
is the container.

## Images and containers

The idea of the container is simple, and it is easiest
to grasp through the distinction between an image and a
container, which are as different as a recipe from a
meal, or a class from an object.

An **image** is a frozen filesystem plus metadata. It
is, quite literally, a snapshot of a fully installed
computer: an operating system, the system libraries, a
particular R, a particular set of R packages, and the
configuration that ties them together, all captured as
a single read-only artifact. The image does not run.
It is inert, like a disk image or a virtual-machine
template, waiting to be started.

A **container** is what one gets by starting an image.
It is a running instance: the frozen filesystem is
made live, a process begins inside it, and that process
sees exactly the operating system and libraries the
image froze, regardless of what the host machine
outside the container happens to have. One image can
start many containers, each an independent running copy
that begins from the same frozen state. When a
container exits, any change it made to its own
filesystem is ordinarily discarded, and the next
container starts again from the pristine image. The
image is the durable, shippable thing; the container is
the ephemeral, running thing.

Images are built in **layers**, and the layering is
worth understanding because it explains both the speed
of the tool and the way we shall pin it. An image is
built from a recipe, the Dockerfile, whose instructions
execute in order, and each instruction adds one layer
to a stack. A layer that installs the operating system
sits at the bottom; a layer that installs R sits on top
of it; a layer that installs packages sits on top of
that. Because layers are content-addressed and cached,
an unchanged lower layer is never rebuilt, and two
images that share a base share its layers on disk. This
is why building a second project from the same base is
fast, and why, later, we shall be careful about which
layer a given change invalidates.

The property that earns the container its place in this
book is often summarized as 'it runs the same
everywhere it runs'. The claim is precise, and the
repetition of 'runs' is deliberate. Wherever a
container runtime is present, the same image starts a
container with the same operating system, the same
system libraries, and the same R, so a result computed
inside it does not depend on the host. The
qualification is that the container must actually run:
the guarantee holds across every machine with a
compatible runtime, and it is silent about machines
that have none. This is a real boundary, and we return
to it, and to the runtimes that widen it, in
@sec-backends. For present purposes the container gives
us what the lockfile could not: a fixed operating
system, fixed system libraries, and a fixed R,
traveling together as one artifact.

::: {.callout-note icon=false collapse=true}
## Check your understanding: image or container

A colleague says, 'I deleted a file inside the
container by accident, so I restarted it and the file
was back. Docker must have a recovery feature.' What is
actually happening, in the vocabulary of this section?

::: {.callout-note icon=false collapse=true}
## Answer

There is no recovery feature. A container is a running
instance of a read-only image, and changes a container
makes to its filesystem are ordinarily discarded when
it stops. Restarting produces a fresh container that
begins again from the pristine image, in which the file
was never deleted. The image is durable and the
container is ephemeral. This is the same property that
makes a container reproducible: every run starts from
the identical frozen state. It also means that anything
one wants to keep, such as edited source or output,
must live outside the container, on a host directory
made visible inside it.
:::
:::

## The Rocker project

A Dockerfile could, in principle, begin from a bare
operating system and install R by hand. Almost no one
does, because the R community maintains a set of ready
base images that have become the standard starting
point. This is the Rocker project. Boettiger introduced
Docker to the reproducible-research audience and argued
that a container is the natural unit for shipping a
computational environment [@boettiger2015]. Boettiger
and Eddelbuettel then built and documented the Rocker
images themselves, a family of R-specific base images
maintained in step with R releases [@boettiger2017].
The fuller account, including the tooling that has grown
around the images, is given by Nüst and colleagues in
their survey of what they call the Rockerverse
[@nust2020rockerverse].

The practical value of Rocker is that it supplies, for
each R version, a tested base image with R already
installed on a known operating system. The analyst's
recipe can then begin from a solid and shared
foundation, rather than reconstructing one. The images
form a small hierarchy. At the bottom sits
`rocker/r-ver`, R on Ubuntu and little else. Above it,
`rocker/rstudio` adds the RStudio Server development
environment. `rocker/tidyverse` builds on
`rocker/rstudio`, adding the tidyverse packages and the
`pandoc` document converter, so it carries RStudio
Server as well and is the larger of the two. And
`rocker/verse` adds a LaTeX toolchain for rendering
manuscripts to PDF. Each higher image is built on the
one below, so they share layers, and each fixes a
larger environment than the last. We shall meet these
again as the profiles.

## Making the image itself reproducible

Here we must be careful, because a container is
reproducible only if the recipe that builds it is, and
the naive recipe is not. Two refinements separate a
container that merely runs today from one that will
rebuild to the same image next year. Nüst and
colleagues collect the full set of practices in their
ten simple rules for writing Dockerfiles for
reproducible data science [@nust2020]; the two that do
the most work are the following.

The first refinement pins the base image by its
**digest** rather than by a **tag**. A tag such as
`rocker/tidyverse:4.4.0` is a mutable pointer. It names
an image today, but the maintainers may rebuild and
repoint it tomorrow, for instance to patch a security
flaw in the operating system. A recipe that named
the tag would then build on different bytes, without any
change to the recipe itself. A digest is an immutable
hash of the exact image content, so that

```docker
FROM rocker/tidyverse@sha256:1c2f...e8a9
```

names not 'whatever `4.4.0` points to' but one specific,
byte-identical image that the digest will denote
forever. Pinning the digest is the difference between a
`FROM` line that is a promise and one that is a hope.

@fig-tag-digest follows one unchanged recipe across a
year to show what the difference costs. The recipe is not
edited at any point in either row; the only thing that
varies is whether its `FROM` line names the tag or the
digest. Note that nothing in the upper row is an error or
a misconfiguration on the maintainer's part. Repointing a
tag to patch a security flaw is the correct thing for
them to do, which is exactly why a tag cannot carry a
reproducibility claim.

```{r}
#| label: fig-tag-digest
#| fig-cap: 'One unchanged recipe built repeatedly over a year, under a tag and under a digest. The upper row builds on whatever the tag currently points at, so a maintainer rebuild in April and another in September silently change the ground the analysis stands on, giving three different environments from identical text. The lower row names a digest and receives the same bytes at every build. The maintainer rebuilds are ordinary maintenance, not mistakes: the point is that a mutable pointer cannot carry an immutable claim.'
#| fig-alt: 'Two horizontal timelines running from January to December. The upper timeline, labeled FROM rocker/r-ver colon 4.4.0, is divided into three differently colored segments broken at April and September, marked as maintainer rebuilds, and annotated three environments from one recipe. The lower timeline, labeled FROM rocker/r-ver at sha256, is a single unbroken segment of one color across the whole year, annotated one environment, always.'
#| fig-width: 7
#| fig-height: 2.6

library(ggplot2)
library(tibble)

spans <- tibble(
  row = c('FROM rocker/r-ver:4.4.0',
          'FROM rocker/r-ver:4.4.0',
          'FROM rocker/r-ver:4.4.0',
          'FROM rocker/r-ver@sha256:6f05...'),
  xmin = c(0, 3.5, 8.5, 0),
  xmax = c(3.5, 8.5, 12, 12),
  build = c('build 1', 'build 2', 'build 3', 'build 1')
)
spans$row <- factor(spans$row,
                    levels = c('FROM rocker/r-ver@sha256:6f05...',
                               'FROM rocker/r-ver:4.4.0'))

ggplot(spans) +
  geom_rect(aes(xmin = xmin, xmax = xmax,
                ymin = as.numeric(row) - 0.26,
                ymax = as.numeric(row) + 0.26,
                fill = build),
            color = 'white', linewidth = 1.4) +
  scale_fill_manual(values = c('build 1' = '#14746f',
                               'build 2' = '#7fb1ad',
                               'build 3' = '#c9e2df'),
                    guide = 'none') +
  scale_y_continuous(breaks = 1:2,
                     labels = levels(spans$row),
                     limits = c(0.45, 2.95)) +
  scale_x_continuous(breaks = c(0, 3.5, 8.5, 12),
                     labels = c('Jan', 'Apr', 'Sep', 'Dec')) +
  annotate('text', x = 3.5, y = 2.42, label = 'maintainer\nrebuild',
           size = 2.6, lineheight = 0.95, color = 'grey25') +
  annotate('text', x = 8.5, y = 2.42, label = 'maintainer\nrebuild',
           size = 2.6, lineheight = 0.95, color = 'grey25') +
  annotate('text', x = 12.15, y = 2, hjust = 0,
           label = 'three environments\nfrom one recipe',
           size = 2.7, lineheight = 0.95, color = '#a33') +
  annotate('text', x = 12.15, y = 1, hjust = 0,
           label = 'one environment,\nalways',
           size = 2.7, lineheight = 0.95, color = 'grey25') +
  coord_cartesian(xlim = c(0, 16.6), clip = 'off') +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 9.5) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(family = 'mono', size = 8, hjust = 1)
  )
```

The second refinement concerns what happens inside the
image when it installs packages. An installer pointed at
a live package repository resolves 'the current version'
to whatever is current on the day of the build, so that
an image built in March and an image built in June from
the same recipe carry different packages. The remedy is
to point the in-image installer at a **dated snapshot**
of the repository, a frozen view of the repository as it
stood on a chosen day. With the snapshot fixed, 'the
current version' resolves to the same version on every
build. The two refinements together close the loop: the
digest fixes the ground the image is built on, and the
dated snapshot fixes what is installed onto it, so that
the recipe becomes a deterministic function from text to
image.

A small, faithful Dockerfile with both refinements in
place, and with the environment variables of the next
section, looks like this. It is shown as static text;
we do not execute it here.

```docker
# Pin the base image by immutable digest, not a tag.
FROM rocker/r-ver:4.4.0@sha256:6f05...8eac

LABEL org.opencontainers.image.licenses="GPL-3.0-or-later" \
      zzcollab.base.image="rocker/r-ver:4.4.0" \
      zzcollab.ppm.snapshot="2026-06-27"

# Determinism: fix locale, time zone, and threads.
ENV LANG=en_US.UTF-8 \
    LC_ALL=en_US.UTF-8 \
    TZ=UTC \
    OMP_NUM_THREADS=1

# Point the installer at a dated repository snapshot,
# so 'the current version' is the same on every build.
RUN echo 'options(repos = c(CRAN = "https://packagemanager.posit.co/cran/__linux__/noble/2026-06-27"))' \
        >> /usr/local/lib/R/etc/Rprofile.site

# Bootstrap renv, then restore the exact package set.
COPY renv.lock renv.lock
RUN R -e "install.packages('renv')"
RUN R -e "renv::init(bare = TRUE); renv::restore()"
```

The reader will note how the container and the lockfile
divide the labor. The `FROM` line and the snapshot fix
the environment, the operating system, the R, and what
may be installed; the `COPY renv.lock` and
`renv::restore()` fix the exact package versions inside
that environment. Neither alone is enough. A container
without a lockfile fixes the environment but not which
package versions live in it; a lockfile without a
container fixes the packages but not the platform. A
genuinely L2 analysis uses both, and the honest reading
of an image is that it reaches L2 only when both
refinements are present and a lockfile is restored
inside it.

::: {.callout-important}
## The honest level

An image built on a mutable tag, installing from a live
repository, is not L2, however much it looks like a
container. It will rebuild to a different image next
year, and a result it produced cannot be regenerated
from the recipe alone. Reaching L2 requires the digest
pin, the dated snapshot, and a restored lockfile
together. Match the claim to what the recipe actually
fixes. A useful discipline is to build the image twice
and confirm the two builds carry the same content, that
is, the same operating system and the same R and package
versions, for example by comparing `renv::status()` and
the installed apt and R package versions across the two
builds; if the content differs, the recipe is not yet
pinned, and the honest level is still L1. Do not test
this by comparing the two image digests: Docker digests
are not reproducible across rebuilds by default, because
per-layer timestamps and file modification times differ,
so two builds of a correctly pinned recipe carry
different digests. Bit-identical digests require
reproducible-build tooling, such as `SOURCE_DATE_EPOCH`
and the BuildKit rewrite-timestamp option, which is a
separate concern from whether the content is pinned.
:::

## Environment-variable determinism

A fixed operating system and a fixed R are still not
quite enough, because the same R on the same operating
system can behave differently according to a handful of
environment variables that the container inherits. Two
of these deserve concrete illustration, because the
failures they cause are silent and the fixes are one
line each.

The first is the **locale**, carried in `LANG` and
`LC_ALL`, which governs, among other things, how text is
sorted and how numbers are formatted and parsed. A
locale that uses a comma as the decimal mark, common
across much of Europe, writes the number one and a half
as `1,5`, whereas a locale that uses a period writes
`1.5`. Code that reads a data file written under one
convention on a machine configured for the other will
not error. It will misread the numbers. To make the
failure concrete, consider the same numeric vector
written under each convention, and what a parser that
assumes a period decimal mark reads back from the
comma-convention strings.

```{r}
#| label: tbl-decimal-mark
#| tbl-cap: 'The same numbers written under a period decimal mark and a comma decimal mark, and what a parser assuming a period reads back from each. The comma-convention values are silently misread, with no error raised.'
library(tibble)

x <- c(1.5, 2200.75, 0.333)

period <- formatC(x, format = 'f', digits = 3, big.mark = ',')

comma <- gsub(',', '#', period)
comma <- gsub('\\.', ',', comma)
comma <- gsub('#', '.', comma)

read_back <- as.numeric(gsub(',', '', comma))

decimal_demo <- tibble(
  original      = x,
  period_locale = period,
  comma_locale  = comma,
  parsed_value  = read_back
)

knitr::kable(decimal_demo)
```

The parsed values are not approximately wrong; they are
wildly wrong, and nothing warns the analyst. The number
`1.5` returns as `1500`, the number `2200.75` as
roughly `2.2`, and the number `0.333` as `333`, because
the period-assuming parser treats the European comma as
noise and strips it. Fixing `LANG` and `LC_ALL` in the
image removes the ambiguity: every container runs under
the one declared locale, so the parse is the same
everywhere the image runs. This is exactly why the
generated Dockerfile above sets both to `en_US.UTF-8`.

The second is the **time zone**, carried in `TZ`. Date
and time arithmetic depends on it, and an analysis that
buckets events by calendar day will assign a
late-evening event to different days in different time
zones. Fixing `TZ=UTC` in the image makes the
calculation independent of where the container runs.

The third is the **thread count**, carried in variables
such as `OMP_NUM_THREADS`. Numerical libraries that
implement linear algebra may split a computation across
several threads. And because floating-point addition is
not exactly associative, the order in which partial sums
are combined can change the last bits of a result. Two
runs of the same bootstrap or the same mixed model, one
on four threads and one on eight, may therefore differ
in the final digits. Which variable actually governs the
summation order depends on the BLAS implementation the
image carries: `OPENBLAS_NUM_THREADS` for OpenBLAS,
`MKL_NUM_THREADS` for the Intel MKL, and `OMP_NUM_THREADS`
for the OpenMP-parallel sections. The rocker default
reference BLAS is already single-threaded, though, and
needs no such variable to be deterministic. Setting the OpenMP and
the BLAS thread counts to one together, rather than any
one of them alone, is what forces a single, deterministic
order of summation, trading a little speed for bitwise
reproducibility. Whether that
trade is worth making is a judgment, and it is the
analyst's; the point is that a container lets the
judgment be recorded and enforced rather than left to
whatever the host defaults to.

::: {.callout-note icon=false collapse=true}
## Check your understanding: which variable

A collaborator in Berlin and a collaborator in San
Diego run the same containerised script on the same
image and obtain identical package versions, yet their
daily case-count tables disagree by one case on several
days, always at the boundaries between days. Which
environment variable is the likely culprit, and why
does pinning it in the image resolve the disagreement?

::: {.callout-note icon=false collapse=true}
## Answer

The likely culprit is `TZ`, the time zone. If the image
does not fix it, each container may inherit the host's
time zone, and an event recorded near midnight is
assigned to different calendar days in Berlin and San
Diego. The daily buckets then differ at the day
boundaries, which is exactly the pattern described.
Setting `TZ` in the image, conventionally to `UTC`,
makes the day assignment independent of the host, so
both collaborators bucket the events identically. Note
that identical package versions, that is, a shared L1,
did nothing to prevent this: the failure is an
environment failure, which is precisely what the
container layer exists to fix.
:::
:::

## Profiles as base images

The choice of base image is, in practice, a choice from
a short menu, and the framework this book uses names the
entries **profiles**. A profile is a base-image bundle,
and selecting one sets the `FROM` line and little else.
Four profiles cover most needs.

The **minimal** profile builds on `rocker/r-ver`: R on
Ubuntu with the development essentials and nothing more.
It is the smallest base, and the right one when the
analysis pulls in whatever it needs through the
lockfile. The **tidyverse** profile builds on
`rocker/tidyverse`, adding the tidyverse packages and
`pandoc`, and is the natural default for data analysis.
The **rstudio** profile builds on `rocker/rstudio`,
adding the RStudio Server environment for those who
prefer to work in the browser-based IDE. The
**publishing** profile builds on `rocker/verse`, which
carries a LaTeX toolchain, and is the base to choose
when the deliverable is a rendered PDF manuscript.

The design intent is that the profiles differ in as few
lines as possible. Moving a project from one profile to
another changes the `FROM` line and, at most, a single
system-installation line, while the
reproducibility-relevant configuration, the locale and
time-zone variables, the dated snapshot, and the
lockfile restore, stays constant. The profile decides
what the base carries; the lockfile decides what is
installed on top; and the two remain cleanly separated.

The archetypes of @sec-compendium map onto the profiles
naturally, and the mapping is the concrete form of the
first analyst judgment above. The same machinery, a
digest-pinned base, a dated snapshot, a restored
lockfile, and the determinism variables, serves every
archetype; only the base changes.

- A **data-analysis** archetype is well served by the
  tidyverse or minimal base, according to how much of
  its stack it draws from the base rather than the
  lockfile.
- A **manuscript** archetype needs the publishing base,
  because rendering to PDF requires the LaTeX toolchain
  that `rocker/verse` carries and the lighter bases do
  not.
- An **R-package** archetype needs very little: R and
  the check tools suffice, so the minimal base is the
  honest choice, and provisioning a LaTeX toolchain it
  never uses would be waste.
- A **simulation** archetype often wants more cores than
  software, so the base may be minimal while the
  container is run with a larger core allocation. Here
  the analyst must weigh the determinism point above:
  more cores may speed the simulation while changing the
  last bits of a result, so `OMP_NUM_THREADS` and the
  random-number discipline of @sec-dependencies deserve
  particular care.
- A **blog** archetype renders many posts rather than one
  report, and is well served by the publishing or
  tidyverse base according to whether any post needs
  PDF output.

The point to carry forward is that a research compendium
is not only a manuscript, and the container does not
assume it is. One set of machinery, varied by a single
`FROM` line, pins the environment for all of them.

## A first look at the two-layer team model

A container fixes an environment for one analyst. A team
needs every member to share the same environment, and
this introduces a wrinkle we shall only sketch here and
develop fully in @sec-archiving. Building an image from
a Rocker base and a full package set takes many minutes,
and it would be wasteful for every member of a lab, and
every one of a lab's projects, to rebuild the same base
from scratch.

The remedy is a two-layer model. A team lead builds a
comprehensive **team image** once, carrying the lab's
preferred packages, and pushes it to a shared registry.
Each project then builds a thin second layer on top of
that team image, rather than on the raw Rocker base. The
first layer, the team image, is built once and reused
everywhere; the second layer, the per-project image,
adds only what the project needs and builds in seconds
rather than minutes. Because the team image is itself
pinned by digest and recorded in a committed file, every
member pulls byte-identical bytes, so 'works on my
machine' cannot arise: there is, in the relevant sense,
only one machine. The mechanics of building, pushing,
digest-recording, and updating the team image are the
business of @sec-archiving. For now it is enough to see
that the image is not only a solo instrument but the
natural unit of a shared, reproducible environment.

## Worked example

Let us follow a concrete case through, in the manner of
the framework this book uses, which generates the
Dockerfile rather than asking the analyst to write one
by hand. A biostatistician is analyzing a registry
cohort. The deliverable is a set of tables and figures,
not a typeset manuscript, so the appropriate archetype
is data analysis and the appropriate profile is minimal:
R and the essentials, with the analysis stack drawn from
the lockfile. She scaffolds the project and builds the
image. The shell session, shown as static text and never
executed here, is short.

```bash
$ mkdir cohort-study && cd cohort-study
$ zzc minimal
$ make docker-build
$ make r
```

The `zzc minimal` step generates the project, including
a Dockerfile pinned to a digest and a dated snapshot;
`make docker-build` builds the image; and `make r` starts
a container and drops her into an R session inside it.
The generated Dockerfile is, in essence, the one shown
earlier: a digest-pinned `FROM rocker/r-ver`, the locale
and time-zone and thread variables, the dated snapshot
wired into the repository options, and a
`renv::restore()` that installs the locked packages
inside the fixed environment. The image records its own
provenance in labels, the base image, its digest, and
the snapshot date, so a later reader can see exactly what
was pinned without leaving the image.

Now suppose a reviewer asks for a typeset manuscript with
the analysis embedded. The archetype has shifted, and so
should the profile. She regenerates against the
publishing base:

```bash
$ zzc publishing
$ make docker-build
```

The `FROM` line now names `rocker/verse`, the LaTeX
toolchain comes with it, and the manuscript renders to
PDF inside the container. Almost nothing else in the
recipe changed: the locale and thread variables, the
dated snapshot, and the lockfile restore are identical.
This is the design intent made concrete. The same
reproducibility machinery served both archetypes, and
the move between them was a single changed line.

What has she gained, in the vocabulary of the ladder?
Before the container, at L1, her collaborator's restore
could fail for a missing system library, or succeed and
compute a different number under a different R. After
it, at L2, the operating system, the system libraries,
and the R are fixed and travel with the analysis, and
the locale and time-zone and thread variables are
declared rather than inherited. What she has not yet
gained is verification, L3: an image that builds and a
report that renders are not the same as a report whose
numbers have been regenerated and shown to match a
recorded baseline. That is the subject of
@sec-verification, and it is the honest boundary of what
this chapter delivers.

## Collaborating with an LLM

A language model is a capable assistant for container
work and a confident source of subtly wrong Dockerfiles,
because the most common Dockerfile on the public internet
is not a reproducible one. The analyst who delegates must
verify, and the following three exchanges show what to
watch for.

**Prompt.** 'Write me a Dockerfile for my R analysis.'

*Watch for.* The default answer will very likely open
with `FROM rocker/tidyverse:latest` or a bare version
tag, install packages from the live repository, and set
none of the determinism variables. Every one of these
is a reproducibility defect: the mutable tag, the
undated installer, the absent locale and time zone. The
result runs today and is not L2.

*Verification.* Confirm that the `FROM` line carries an
`@sha256:` digest, not merely a tag; that the in-image
installer points at a dated snapshot rather than a live
repository; and that `LANG`, `LC_ALL`, and `TZ` are set.
Then build the image twice and confirm the two builds
carry the same content, comparing the installed operating
system and the R and package versions, for instance
through `renv::status()` and the apt and R package
version lists, rather than comparing image digests, which
differ across rebuilds even for a correctly pinned recipe
because per-layer timestamps vary. If the content
differs, the recipe is not pinned, whatever the model
asserted.

**Prompt.** 'My container gives different numbers than my
laptop. Why?'

*Watch for.* The model is likely to reach first for the
code, suggesting a bug in the analysis, and to overlook
the environment. The environment is the more probable
cause precisely because the container was supposed to
fix it and may have fixed it incompletely: a missing
locale variable, a different thread count, an unpinned
package version inside a pinned environment.

*Verification.* Compare the environment inside the
container with the one outside it: inspect `LANG`,
`LC_ALL`, `TZ`, and the thread-count variables in both,
and compare the installed package versions against the
lockfile. Do not accept a diagnosis of the code until
the environment has been ruled out, because a difference
the container failed to pin is the more likely
explanation.

**Prompt.** 'Which Rocker base should I use for my
project?'

*Watch for.* The model tends to recommend a large,
capable base by default, often the verse or a
data-science image, regardless of need. Over-provisioning
inflates the image and obscures the real dependencies;
under-provisioning, recommending a minimal base for a
project that must render a PDF, produces a build that
fails only when the manuscript is rendered.

*Verification.* Match the base to the archetype, not to
the maximum. If the deliverable is a PDF manuscript,
confirm the base carries a LaTeX toolchain, for instance
by checking that the LaTeX tools are present in the built
image. If it is an R package, confirm that R and the
check tools suffice and resist any heavier base. The
judgment is the first of this chapter's analyst
contributions, and it is not one to delegate.

## Exercises

1. Take any Rocker image on your machine or on Docker
   Hub and find both its tag and its content digest.
   Explain, in two or three sentences, what could change
   about the image named by the tag while the image named
   by the digest could not, and why a reproducible recipe
   pins the latter.

2. You are handed a Dockerfile that begins
   `FROM rocker/tidyverse:latest` and installs packages
   with a plain `install.packages()` from the live
   repository. Identify every reason this recipe fails to
   reach L2, and rewrite the two lines that most need
   fixing.

3. For each of the five archetypes, data analysis,
   manuscript, R package, simulation, and blog, name the
   profile you would choose and justify the choice in one
   sentence. For the two archetypes where the choice is
   genuinely arguable, state the trade-off on both sides.

4. Extend the executable table in this chapter to a date
   example. Construct two synthetic character vectors
   representing the same three dates under a
   month-first and a day-first convention, parse each
   under a single assumed convention, and show which
   dates are silently swapped. Relate the result to the
   role of `TZ` and locale in the image.

5. Explain, to a colleague who has never met the idea,
   why setting `OMP_NUM_THREADS=1` can change the last
   digits of a bootstrap confidence interval or a
   mixed-model estimate. State clearly what is traded
   away by fixing it, and name one analysis where you
   would judge the trade worth making.

6. A collaborator reports that after pulling your image
   and running `renv::restore()`, she still obtains
   different package versions from yours. The environment
   is pinned; the packages are not matching. Diagnose the
   most likely cause, and explain why a container alone,
   without the lockfile restored inside it, does not
   guarantee identical packages.

## Solutions

Attempt an exercise before expanding its solution. The first
asks the reader to inspect an image on their own machine, so
the values will vary; the reasoning is what the box records.

::: {.callout-note icon=false collapse=true}
## Exercise 1

The tag and the digest are found together:

```bash
docker pull rocker/r-ver:4.4.0
docker inspect --format='{{index .RepoDigests 0}}' rocker/r-ver:4.4.0
```

The tag is `rocker/r-ver:4.4.0`; the digest is the
`sha256:...` string, and it will differ from any value
printed here.

**What can change under the tag.** A tag is a mutable
label, a name someone chose to point at an image. The
maintainer can rebuild `4.4.0` to patch a system library,
pick up a security update to the base Ubuntu layer, or
correct a packaging mistake, and republish it under the
same tag. Everything about the image can change: its
system libraries, its preinstalled packages, its locale
data, even its R patch level. What cannot change is the
image named by the digest, because the digest is a content
hash: an altered image hashes differently and is therefore
a different image by construction, not by convention.

A reproducible recipe pins the digest because it wants the
second guarantee rather than the first. Pinning the tag
records an *intention* about which image to use; pinning
the digest records the image. The cost is that a
digest-pinned recipe does not receive security updates
silently, which is usually the right trade for an
analysis and the wrong one for a production service.
:::

::: {.callout-note icon=false collapse=true}
## Exercise 2

**Every reason the recipe fails to reach L2.**

1. `:latest` is the most mutable tag there is. The image
   the recipe resolves to changes without notice, so two
   builds of the identical Dockerfile can produce
   different environments.
2. A plain `install.packages()` draws from the live
   repository, so it installs whatever is current on the
   day of the build. The package versions are not pinned
   at all, which means the recipe does not even reach L1.
3. No lockfile is restored, so there is no record of what
   was installed even after the fact.
4. The determinism variables are absent: no `LANG`, no
   `TZ`, no thread count. The locale and time zone are
   inherited from the base and can shift with it.

The first two are the ones to rewrite:

```docker
FROM rocker/tidyverse:4.4.0@sha256:6f05...8eac

RUN Rscript -e "install.packages('renv'); renv::restore()"
```

The first line pins the image by digest, so the base
cannot move. The second restores from the committed
lockfile rather than resolving against the live
repository, so the package set is the recorded one. A
fuller repair would add the locale, time-zone, and
thread-count variables, but the exercise asks for the two
lines that most need fixing, and these are the two that
carry the recipe from unpinned to pinned.
:::

::: {.callout-note icon=false collapse=true}
## Exercise 3

**Data analysis:** tidyverse, because the analysis draws
most of its stack from the base rather than the lockfile
and the base saves the install. **Manuscript:**
publishing, because rendering to PDF needs the LaTeX
toolchain that `rocker/verse` carries and the lighter
bases do not. **R package:** minimal, because a package
must be checked against a clean environment carrying only
its declared dependencies, and a rich base can mask a
missing declaration. **Simulation:** minimal, because a
simulation typically needs a small numeric stack and image
size matters when the image is pulled onto many cluster
nodes. **Blog:** tidyverse or publishing, depending on
whether posts render to HTML alone.

**The two genuinely arguable cases.**

*Data analysis.* Tidyverse buys a fast build and a base
whose contents are already tested together. Minimal buys a
smaller image and, more importantly, honesty: everything
the analysis uses comes through the lockfile, so the
lockfile is a complete record rather than a supplement to
an opaque base. The trade is build convenience against
completeness of the record.

*Blog.* Publishing carries LaTeX, which a blog rendering
only to HTML never uses, and it is a large image to pull
and store for that. Tidyverse is lighter but leaves the
blog unable to render a PDF version of a post without a
rebuild. The trade is image size against format
optionality, and it turns on whether any post will ever
need PDF.

The general shape: a heavier base is faster to start with
and less honest about what the analysis actually depends
on.
:::

::: {.callout-note icon=false collapse=true}
## Exercise 4

```r
library(tibble)
dates <- c('01/02/2026', '03/04/2026', '05/06/2026')
tibble(
  written     = dates,
  as_us       = format(as.Date(dates, '%m/%d/%Y'), '%d %B %Y'),
  as_european = format(as.Date(dates, '%d/%m/%Y'), '%d %B %Y')
)
```

Every one of the three parses successfully under both
conventions and means something different under each:
`01/02/2026` is 2 January under the day-first reading and
1 February under the month-first reading. No error is
raised, because both readings are valid dates. This is the
same failure mode as the decimal-mark table earlier in the
chapter, and it is worse in one respect: the decimal-mark
case produces visibly implausible magnitudes, whereas a
swapped date is simply a different plausible date.

The dates that are *silently* swapped are exactly those
where both components are twelve or under, which is
roughly a third of a calendar year. A date such as
`25/12/2026` fails to parse under the month-first
convention and so announces itself; `01/02/2026` does not.
The dangerous cases are the ones that do not error.

**The relation to `TZ` and locale.** Neither variable
changes how an explicit format string parses, so a script
that writes `%d/%m/%Y` is safe under any locale. They bite
where the format is *inferred*: a locale-dependent parser,
a spreadsheet export, or a `read_csv()` guessing a column
type will resolve the ambiguity differently under
different locale settings. Fixing `LANG` and `TZ` in the
image removes that variability, which is why the chapter
sets them. The stronger discipline is to state the format
explicitly and never rely on inference, and to prefer ISO
8601 in any file the analysis writes, since `2026-02-01`
is unambiguous under every convention.
:::

::: {.callout-note icon=false collapse=true}
## Exercise 5

**Why the digits move.** Floating-point addition is not
associative: adding a long column of numbers left to right
can give a slightly different answer from adding it in
pairs and combining, because each intermediate result is
rounded. A multithreaded numerical library splits the sum
across threads and combines the partial results, and the
order in which the partials arrive depends on the thread
count and on scheduling. So the same data can produce
answers differing in the last few digits between a
one-thread and an eight-thread run, and sometimes between
two eight-thread runs. A bootstrap interval or a
mixed-model estimate is built on many such reductions, so
the difference propagates to the reported bounds.

Setting `OMP_NUM_THREADS=1` removes the variability by
removing the parallelism: there is one summation order and
it is the same every time.

**What is traded away.** Speed, and potentially a great
deal of it. A mixed model on a large dataset, or a
bootstrap with many replicates, can run several times
faster across eight threads. Fixing the thread count to
one converts a determinism problem into a wall-clock
problem.

**Where the trade is worth making.** Any analysis whose
outputs will be verified against recorded values, which is
to say anything reaching for L3, since the verification
compares numbers and the numbers must be stable. The
concrete case is a regulatory submission: a fixed thread
count for the final analysis makes the reported estimates
reproducible bit for bit, and the extra hours are
irrelevant against the cost of a discrepancy an auditor
cannot explain. For exploratory work during development,
the trade is usually not worth making.

A refinement worth mentioning: one can develop with
threads and fix the count only for the final verified run,
which recovers most of the speed and all of the
determinism where it matters.
:::

::: {.callout-note icon=false collapse=true}
## Exercise 6

**The most likely cause** is that `renv::restore()` did
not restore into the library the session is actually
using, or did not run against the lockfile you think it
did. In practice this resolves to one of a few things: the
project was opened from a directory where the lockfile is
absent, so `restore()` had nothing to read; the container
mounts a host `renv` cache or library that shadows the
project library; `RENV_PATHS_LIBRARY` points somewhere
unexpected inside the image; or the lockfile itself was
not committed, so the collaborator restored an older
version of it.

The diagnostic sequence is `renv::status()` to see whether
the library and the lockfile agree, then `.libPaths()` to
see which library is in force, then `renv::project()` to
confirm which project `renv` believes it is in. Between
them these locate all four causes.

**Why the container alone does not guarantee identical
packages.** A container pins the environment: the
operating system, the system libraries, and the R runtime.
The packages installed into the image at build time are
pinned by the image, but any package the analysis installs
or restores *afterward* is not, because it is resolved at
run time against whatever the lockfile or the repository
then says. The image and the lockfile do different jobs,
and the chapter's framing is exactly this: the image is
the L1-to-L2 step and the lockfile is the L0-to-L1 step.
Skipping the lower rung does not become safe because you
climbed the higher one.

The compact statement: the container guarantees the same
*floor*, the lockfile guarantees the same *packages*, and
an analysis needs both.
:::

## Further reading

The foundational case for containers in reproducible
research is @boettiger2015, which introduces Docker to a
scientific-computing audience and is the natural first
read for anyone new to the tool. The Rocker images
themselves are documented by @boettiger2017, and the
wider ecosystem of R containerization tooling is
surveyed thoroughly by @nust2020rockerverse; both are
worth consulting before building a base image of one's
own.

On making a container genuinely reproducible rather than
merely convenient, the essential reference is @nust2020,
whose ten simple rules for writing Dockerfiles codify
the digest pin, the dated snapshot, and much else this
chapter has drawn from. Read it as a checklist against
any Dockerfile one is asked to trust.

For the division of labor between the container and the
package layer, return to @ushey2020 on `renv`, keeping in
mind that its remit stops exactly where this chapter's
begins. The compendium structure into which both fit is
the subject of @marwick2018, whose `rrtools` first
combined a lockfile and a Dockerfile in a packaged
analysis. The alternatives to Docker as a runtime, from
the high-performance-computing runtime Apptainer
(formerly Singularity) [@kurtzer2017] to the
browser-based Binder
[@jupyter2018] and the container-free Nix, are taken up
in @sec-backends, and the team-image model sketched here
is developed in full in @sec-archiving.
