# Testing and Code Quality {#sec-testing}

::: {.bookepigraph}
```{=latex}
\begin{flushright}\small
```
*Program testing can be used to show the presence of bugs,
but never to show their absence!*

[Edsger W. Dijkstra]{.smallcaps}, *Notes on Structured
Programming* (1972)
```{=latex}
\end{flushright}
```
:::

## Learning objectives

By the end of this chapter the reader should be
able to:

- Explain the reproducibility paradox, and say
  why testing is a reproducibility layer rather
  than a software nicety.
- Distinguish capturing reproducibility from
  validating it, and place testing squarely in
  the validation register.
- Write the four kinds of test a compendium
  wants, using `tinytest` bare expectations:
  unit, integration, data-validation, and
  reproducibility tests.
- Test behavior rather than implementation, so
  that a test survives an honest refactor.
- Apply `lintr`, `styler`, and a pre-commit hook
  to keep code legible and consistent, and say
  why legibility is itself a reproducibility aid.
- State how the testing effort shifts across the
  compendium archetypes, from an R package to a
  data analysis to a simulation to a blog.

## Orientation

Consider a registry study in which an analyst
has written a small function to compute a mean
biomarker value for each treatment arm. The
function carries a quiet bug: it sums the values
but forgets to divide by their count. The
analysis runs without error. It produces a
table, a figure, a p-value. Sent to a colleague,
cloned onto a second machine, restored from a
lockfile inside a pinned container, it runs again
and produces the identical table, the identical
figure, the identical p-value. Every rung of the
ladder of @sec-levels is satisfied: the source is
locatable, the packages are pinned, the
environment is pinned. The result is perfectly
reproducible. It is also perfectly wrong.

This is the reproducibility paradox, and it is
the pivot of this chapter. A buggy function that
returns the wrong answer returns the *same* wrong
answer everywhere, on every machine, at every
later date. Reproducibility machinery, faithfully
applied, does not detect the error; it propagates
it with mechanical fidelity. The lesson, pressed
by Leek and Peng in their aptly titled note that
reproducible research can still be wrong
[@leek2015], is that reproducibility is a floor
and not a ceiling. It guarantees that a result can
be regenerated. It says nothing whatever about
whether the result is correct.

The remedy is testing, and the reason testing
belongs in a book about reproducibility, rather
than in a separate book about software
engineering, is precisely the paradox. Pinning a
package version, building a container, restoring a
lockfile: each of these *captures* a determinant
of the computation, and none of them examines
whether the computation is right. Testing captures
nothing. A test adds no package to the lockfile
and no library to the image. What a test does is
*confirm* that a function, a pipeline, or a dataset
behaves as the analyst intended. It belongs, in the
vocabulary this book has used since @sec-intro, to
the validation register rather than the capture
register, alongside the continuous integration and
output verification of @sec-verification. Capture
fixes the inputs; validation checks that the
fixed inputs still produce the answer they are
supposed to. A reproducible analysis without tests
is trusted; a reproducible analysis with tests is
checked, and the difference between trusting and
checking is the difference this chapter develops.

There is a further and more constructive link to
the ladder. The highest rung, L3 verified, is
reached not by capturing more but by *confirming*
that a recorded computational output regenerates.
That confirmation is itself a test, of a
particular kind we shall meet below as the
reproducibility test. Testing is therefore not
merely adjacent to the levels; the topmost level
is defined by a test. We begin with the human
judgments that no testing tool can make, then
develop the four kinds of test in turn, then take
up code quality as a second validation feature,
and close with the way the emphasis shifts across
archetypes.

## The analyst's contribution

Testing tooling runs the checks, tallies the
passes and failures, and reports them. It does not
originate a single one of the judgments that give
a test its value. Three of these judgments are
worth naming at the outset, because they are the
human core of everything that follows.

1. **The analyst decides what 'correct' means.**
   A test compares a computed value against an
   expected one, and the expected one has to come
   from somewhere outside the code under test. This
   source of independent truth is called an oracle:
   a mean of one, two, and three that the analyst
   knows to be two; a count of rows that a data
   dictionary fixes at some known number; a
   coefficient that a textbook formula supplies. No
   tool knows the right answer. The analyst supplies
   it, and a test is only as trustworthy as the
   oracle behind it.

2. **The analyst chooses which behaviors to pin.**
   A function has an infinity of properties, most of
   them incidental to its purpose. The analyst
   decides which few constitute the function's
   contract, the behaviors a future change must not
   break, and writes tests for those. This is the
   distinction, developed below, between testing
   behavior and testing implementation, and it is a
   design judgment, not a mechanical one.

3. **The analyst encodes the domain constraints.**
   The rule that an age lies between zero and one
   hundred and twenty, that a treatment arm is one
   of a fixed set of labels, that a probability sits
   in the unit interval: these are not facts the
   computer can infer. They are clinical and
   scientific knowledge, and a data-validation test
   is the place where that knowledge is written down
   in executable form, so that a violation halts the
   analysis rather than passing silently into a
   result.

::: {.callout-note}
## The vocabulary of this chapter

- **Unit test.** A focused check that one small
  piece of code, typically a single function,
  behaves as intended on known inputs.
- **Integration test.** A check that several
  components behave correctly when composed into a
  pipeline, from raw input to final result.
- **Data-validation test.** A check on the data
  itself rather than the code: that a value lies in
  a plausible range, that no impossible ages appear,
  that the expected columns are present.
- **Reproducibility test.** A check that a recorded
  computational result regenerates, either across
  repeated runs with a fixed seed or against a
  stored baseline.
- **Oracle.** The source of independent truth a
  test compares against; the known correct answer.
- **`tinytest`.** A lightweight, zero-dependency R
  testing framework in which a test file is a plain
  script of bare, top-level expectation calls
  [@vanderloo2020].
- **Expectation.** A single assertion, such as
  `expect_equal(actual, expected)`, that passes or
  fails.
- **Linting.** Static analysis of source code for
  stylistic and likely-error patterns, performed by
  `lintr`, without running the code.
- **Styling.** Automatic reformatting of source code
  to a consistent layout, performed by `styler`.
- **Pre-commit hook.** A script that git runs
  automatically before a commit is recorded, used
  here to run the linter and the styler as a gate.
:::

## The reproducibility paradox, stated precisely

It is worth stating the paradox with a little more
care, because its precise form dictates the remedy.
A computational result depends on the code, the
data, and the stack of determinants beneath them
that @sec-levels enumerated. The capture tools of
Part II fix those determinants, so that the
mapping from input to output is held constant
across machines and across time. But holding a
mapping constant says nothing about whether the
mapping is the intended one. If the code encodes
the function 'sum' where the analyst meant the
function 'mean', then capture faithfully preserves
'sum', and every reproduction returns the sum. The
error is not in the reproduction; the reproduction
is flawless. The error is in the mapping, and the
only instrument that examines the mapping is a
test.

Two consequences follow, and they cut in opposite
directions. On the one hand, a result that fails to
reproduce is thereby shown to be untrustworthy,
which is a genuine service; irreproducibility is a
detectable defect. On the other hand, a result that
does reproduce has been shown only to be stable, not
to be right, and stability is easily mistaken for
correctness by an analyst who has watched the same
number appear three times. The paradox is dangerous
exactly because reproduction is so persuasive.
Testing exists to break the false inference from
'it came out the same' to 'it came out correct'.

::: {.callout-important}
## The honest level

Placing an analysis at L2 pinned environment, or
even at L3 verified, is a claim about
reproducibility, not about correctness. A result
can sit at the top of the ladder and still be
wrong, if the code faithfully regenerates a
mistaken number. When reporting a level, the honest
formulation is that the analysis reliably
reproduces a result, and that the result has, to
the extent the test suite reaches, been checked for
correctness. The two claims are distinct, and
conflating them is the precise error the paradox
warns against.
:::

## tinytest: bare expectations

The framework this book uses adopts `tinytest`
[@vanderloo2020] as its testing tool, and the
choice rewards a moment's explanation because it
shapes how the tests look. `tinytest` is
lightweight and carries no dependencies of its own,
which matters in a reproducible compendium: every
additional test dependency enlarges the lockfile
and the container image, and a testing framework
that dragged in a dozen packages would tax the very
machinery it is meant to protect. The more visible
consequence is stylistic. A `tinytest` test file is
an ordinary R script containing bare, top-level
expectation calls, with no wrapping function around
them.

A reader who has met `testthat` will notice the
difference at once. Where `testthat` nests each
assertion inside a `test_that('description', {
... })` block, `tinytest` writes the assertion
directly:

```r
# inst/tinytest/test-calculate_mean.R
expect_equal(calculate_mean(c(1, 2, 3)), 2)
expect_equal(calculate_mean(5), 5)
expect_true(is.nan(calculate_mean(numeric(0))))
```

Each line is a complete test. The file is also a
runnable script, so during development an analyst
can source it directly and watch the expectations
evaluate. The available expectations cover the
common cases, `expect_equal()`, `expect_true()`,
`expect_false()`, `expect_identical()`,
`expect_error()`, `expect_inherits()`, and a few
others, and a compendium's tests are collected in
`inst/tinytest/` and driven by a small
`tests/tinytest.R` runner that `R CMD check`
invokes. We follow the convention throughout: bare
top-level expectations, one concern per file, named
`test-<thing>.R`.

## The four kinds of test

A research compendium wants four kinds of test, and
they form a progression from the smallest unit of
code to the reproduction of a whole result. We take
them in turn, illustrating each on synthetic data
of the sort a health study might produce.

### Unit tests of functions

A unit test checks that one function behaves as
intended on inputs whose correct outputs are known
independently. The `calculate_mean` function of the
orientation is the canonical case. Its contract is
that it returns the arithmetic mean, and the oracle
is arithmetic done by hand:

```r
# inst/tinytest/test-calculate_mean.R
expect_equal(calculate_mean(c(2, 4, 6)), 4)
expect_equal(calculate_mean(c(-1, 0, 1)), 0)
expect_true(is.nan(calculate_mean(numeric(0))))
expect_equal(calculate_mean(c(1, NA, 3),
                            na.rm = TRUE), 2)
```

The first two lines pin the ordinary behavior
against hand-computed answers. the third and fourth
pin the edge cases, the empty vector and the
missing value, which are exactly the inputs on which
an untested function tends to misbehave. A unit test
suite is, in effect, a written record of what the
analyst believes each function promises.

### Integration tests of the pipeline

A unit test exercises a function in isolation, but a
health analysis is rarely one function. It is a
pipeline, in which raw data is loaded, cleaned,
modeled, and summarized, and the components can be
individually correct yet compose incorrectly. An
integration test runs the pipeline end to end and
checks the result:

```r
# inst/tinytest/test-cohort.R
raw <- load_cohort('analysis/data/raw_data/cohort_raw.csv')
cohort <- clean_cohort(raw)
fit <- fit_model(cohort)
summary_tbl <- summarise_fit(fit)

expect_inherits(fit, 'lm')
expect_true(all(c('term', 'estimate') %in%
                names(summary_tbl)))
expect_equal(nrow(summary_tbl), 3L)
```

The first expectation looks like a mistake and is not.
`fit_model()` returns a `glm`, because the outcome is
binary, and a `glm` object inherits from `lm`, so the
assertion holds. It is written that way deliberately:
`expect_inherits(fit, 'lm')` pins the contract that the
function returns something the modeling generics will
accept, whereas `expect_equal(class(fit), 'glm')` would
pin the implementation and break the first time the
function was changed to fit a Cox model instead.

The integration test does not re-check what the unit
tests already cover; it checks the seams, that the
output of `clean_cohort` is a valid input to
`fit_model`, that the model summary has the expected
shape. A pipeline that passes its unit tests can
still fail here, if a column is renamed in one step
and read under its old name in the next.

### Data-validation tests

The third kind of test turns its attention from the
code to the data. Much of what goes wrong in a
health analysis is not a coding error but a data
defect: an age of two hundred, a negative
laboratory value, a treatment label misspelled, a
column silently dropped by an upstream export. A
data-validation test encodes the analyst's domain
knowledge as executable constraints and fails when
the data violates them.

Before writing constraints it is worth looking at
what one is constraining. @tbl-diabetic-arms
tabulates the retinopathy trial introduced in the
conventions, and it is included here for a reason
beyond illustration: every cell in it can be counted
by hand from the data, which is precisely the
property an oracle needs. A test asserting that the
untreated arm contains 197 eyes is checkable by
someone who does not trust the function that
produced it. A test asserting that it contains
`nrow(subset(d, trt == 0))` eyes is not a test at
all; it is the function agreeing with itself.

```{r}
#| label: tbl-diabetic-arms
#| tbl-cap: 'Eyes and events by treatment arm in the diabetic retinopathy trial (`survival::diabetic`), the real dataset this book computes on. Each count is verifiable by hand from the 394 records, which is what makes the table usable as a test oracle rather than merely as a description.'

library(tibble)
library(knitr)

d <- survival::diabetic
d$arm <- factor(d$trt, levels = c(0, 1),
                labels = c('Untreated', 'Laser'))

arms <- tibble(
  Arm = levels(d$arm),
  Eyes = as.integer(table(d$arm)),
  Events = as.integer(tapply(d$status, d$arm, sum)),
  `Event rate` = sprintf(
    '%.3f', as.numeric(tapply(d$status, d$arm, mean))
  )
)

kable(arms, align = 'lrrr')
```

The difference between the arms is large and in the
expected direction, which is a reason to be careful
rather than pleased: a validation test should assert
what the protocol requires, not what this particular
extract happens to show. An assertion that the event
rate in the laser arm is below one third would pass
today and would fail on a different trial, or on a
corrected export, for reasons having nothing to do
with a defect. Constraints belong to the protocol,
not to the data in front of you.

```r
# inst/tinytest/test-cohort.R
raw <- load_cohort('analysis/data/raw_data/cohort_raw.csv')
cohort <- clean_cohort(raw)

# the expected columns are present
expect_true(all(c('mrn', 'age_at_enroll', 'sex', 'incident_dm')
                %in% names(raw)))

# ages lie within the enrollment criteria, not merely
# within the observed range
expect_true(all(cohort$age_at_enroll >= 18 &
                cohort$age_at_enroll <= 105, na.rm = TRUE))

# sex takes only its two coded levels once cleaned
expect_true(all(cohort$sex %in% c('F', 'M')))

# outcome missingness stays within what the analysis
# can tolerate
expect_true(mean(is.na(cohort$incident_dm)) < 0.15)
```

These are the tests that most directly repay a
health scientist, because they catch the errors that
no amount of correct code can prevent, the errors
that live in the data before the analysis begins. In
a data-analysis or simulation archetype, where there
may be few reusable functions to unit-test, the
data-validation tests are often the heart of the
suite.

### Reproducibility tests

The fourth kind closes the loop back to the ladder.
A reproducibility test checks that a result
regenerates, and it comes in two forms. The first
fixes a seed and confirms that a stochastic
procedure returns the same draw twice, which guards
against the accidental omission of `set.seed` and
against the silent RNG-default changes discussed in
@sec-intro:

```r
# inst/tinytest/test-cohort.R
cohort <- clean_cohort(
  load_cohort('analysis/data/raw_data/cohort_raw.csv')
)

expect_equal(bootstrap_ci(cohort, n = 50),
             bootstrap_ci(cohort, n = 50))
```

Note where the seed lives. `bootstrap_ci()` takes it as
an argument and sets it internally, so the test does not
set one, and two calls agree because the function is
deterministic given its arguments rather than because
the caller remembered. A procedure that depends on the
caller setting a seed first is one `set.seed` away from
irreproducible, and moving the seed into the signature
makes forgetting it impossible.

The second form is stronger and is the single-result
analogue of the L3 check: it compares a freshly
computed result against a baseline recorded at an
earlier and trusted moment, typically stored as a
small file alongside the tests.

```r
# inst/tinytest/test-baseline.R
baseline <- readRDS('inst/tinytest/baseline-estimate.rds')
current <- run_primary_analysis()
expect_equal(current$estimate, baseline,
             tolerance = 1e-6)
```

The tolerance is not decoration. An exact comparison
would fail whenever a different BLAS or a different
thread count reorders the floating-point summations
inside the fit, as @sec-containers describes, and a
check that fails for reasons unrelated to the analysis
trains its reader to ignore it.

A suite that includes this last test is doing, at
the scale of a single result, what `zzc verify --full`
does at the scale of the whole compendium in
@sec-verification: regenerating a recorded output
and confirming that it matches. L3 proper is that
full-compendium check, which rebuilds the environment
and re-runs the whole analysis; the baseline test is
its single-result counterpart, not a substitute that
places a project at the verified rung on its own.

The four kinds are not four ways of doing one thing.
Each catches defects the others do not see, and
@fig-test-coverage sets four representative defects
against them. Read the bottom row first. Re-execution
alone, with no test of any kind, catches none of the
four, because every one of these defects is
deterministic and will reproduce perfectly. That row is
the reproducibility paradox in tabular form, and it is
the reason this chapter exists.

```{r}
#| label: fig-test-coverage
#| fig-cap: 'Four representative defects against the four kinds of test, with a fifth row for re-execution alone. A filled cell marks the kind of test most likely to catch that defect. The matrix is close to diagonal, which is the point: the kinds are complements rather than alternatives, and a suite missing one kind is blind to a whole class of defect. The bottom row is empty because every defect shown is deterministic, so re-running the analysis reproduces it faithfully.'
#| fig-alt: 'A grid with four defect columns and five test-kind rows. The columns are: dropped rows change the sample size; ages encoded in months; a function returns variance instead of standard deviation; a seeded bootstrap changes after an R upgrade. The rows are unit test, integration test, data-validation test, reproducibility test, and re-execution alone. Filled cells run roughly diagonally: the unit test catches the variance error; the integration test and the data-validation test both catch the dropped rows; the data-validation test catches the ages in months; the reproducibility test catches the changed bootstrap. The re-execution row is entirely empty.'
#| fig-width: 7
#| fig-height: 3.4

library(ggplot2)
library(tibble)

defects <- c('dropped rows change\nthe sample size',
             'ages encoded\nin months',
             'variance returned\nfor a standard deviation',
             'seeded bootstrap moves\nafter an R upgrade')
kinds <- c('Re-execution alone',
           'Reproducibility test',
           'Data-validation test',
           'Integration test',
           'Unit test')

cover <- expand.grid(defect = defects, kind = kinds,
                     stringsAsFactors = FALSE)
cover$defect <- factor(cover$defect, levels = defects)
cover$kind <- factor(cover$kind, levels = kinds)

hits <- c('Unit test|variance returned\nfor a standard deviation',
          'Integration test|dropped rows change\nthe sample size',
          'Data-validation test|dropped rows change\nthe sample size',
          'Data-validation test|ages encoded\nin months',
          'Reproducibility test|seeded bootstrap moves\nafter an R upgrade')
cover$catches <- paste(cover$kind, cover$defect, sep = '|') %in% hits

ggplot(cover, aes(x = defect, y = kind)) +
  geom_tile(aes(fill = catches), color = 'white', linewidth = 2) +
  scale_fill_manual(values = c('TRUE' = '#14746f',
                               'FALSE' = '#eef2f2'),
                    guide = 'none') +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 9) +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(size = 7.6, lineheight = 0.95),
    axis.text.y = element_text(face = 'bold', size = 8.4)
  )
```

::: {.callout-note icon=false collapse=true}
## Check your understanding: which test catches it?

For each defect, name the kind of test most likely
to catch it. (a) A colleague refactors
`clean_cohort` and accidentally drops the rows with
a missing outcome, changing the sample size. (b) An
upstream export encodes age in months rather than
years, so ages read as up to fourteen hundred. (c)
A function meant to return a standard deviation
returns a variance. (d) After an R upgrade, a
seeded bootstrap returns different draws than the
manuscript reported.

::: {.callout-note icon=false collapse=true}
## Answer

(a) An integration test on the pipeline,
which would see the changed row count, or a
data-validation test asserting the expected sample
size. (b) A data-validation test on the age range.
(c) A unit test against a hand-computed standard
deviation. (d) A reproducibility test comparing the
seeded result against a recorded baseline, which is
the single-result form of the L3 check. Note that
the unit and
data-validation tests catch defects that
reproduction alone would happily propagate: this is
the paradox in operation.
:::
:::

## Test behavior, not implementation

A test is worth writing only if it survives changes
that ought not to break it and fails on changes that
ought to. The way to secure this is to test what a
function does, its observable behavior, rather than
how it does it, its internal implementation. The
distinction is easiest to see in a pair of tests for
the same function:

```r
# fragile: pins an implementation detail
expect_true('filter' %in% all.names(body(my_clean)))

# robust: pins the behavior
result <- my_clean(data.frame(x = c(1, NA, 3)))
expect_false(anyNA(result))
```

The first test asserts that `my_clean` uses a
function called `filter` internally. If the analyst
later rewrites `my_clean` in base R, removing the
`filter` call while leaving its behavior
unchanged, the first test fails, and it fails for no
good reason: the function still does exactly what it
promised. Such a test punishes honest refactoring
and trains the analyst to distrust the suite. The
second test asserts only that the cleaned data
carries no missing values, which is the behavior
that actually matters, and it passes under any
correct implementation. As a general rule, a test
should fail when, and only when, the function's
contract is broken. Tests that fail for other
reasons are a tax on maintenance, and Wilson and
colleagues, in their best-practices and
good-enough-practices guidance, make behavioral
testing a recurring recommendation [@wilson2014;
@wilson2017].

A corollary concerns how much to test. A test that
merely confirms a result is a data frame, without
checking any value in it, tests too little and gives
false comfort. A test that pins every incidental
property tests too much and breaks on every
revision. The judgment of what to pin, named above
as the second of the analyst's contributions, is the
skill the tooling cannot supply.

## Code quality as a validation feature

Testing confirms that code is correct. A second
family of tools confirms that code is *legible and
consistent*, and although these tools do not check
correctness directly, they belong to the same
validation register and serve the same
reproducibility end. The argument is one of
mechanism. Code that a reader can follow is code
whose errors a reader can find; consistent
formatting removes the visual noise that hides a
transposed argument or a misplaced parenthesis; and
a single house style across a compendium means that
a difference in the code is always a difference in
meaning, never merely a difference in whitespace.
Wilson and colleagues, in both of their practice
papers, treat readable, conventional code as a
component of reproducible work rather than an
aesthetic preference [@wilson2014; @wilson2017];
Sandve and colleagues, in their ten simple rules,
make the companion point that the provenance and the
recorded results of an analysis must themselves be
preserved [@sandve2013].

Two tools mechanize this, and a third enforces them.

**Linting** is static analysis: `lintr` reads the
source without running it and flags stylistic
deviations and likely errors, an assignment with
`=` where the house style is `<-`, a line past the
width limit, a variable assigned and never used, a
call to a function that does not exist. Because it
does not execute the code, it catches a class of
mistake that tests, which only exercise the paths
they call, may miss.

**Styling** is automatic reformatting: `styler`
rewrites the source into a consistent layout,
fixing indentation, spacing, and alignment, so that
formatting is never a matter for human attention or
for a code review. The two are complementary;
`styler` fixes the formatting that `lintr` would
otherwise complain about.

In the framework, both run inside the project
container through Makefile targets, so that no host
R installation is required and every collaborator
runs the identical checks:

```bash
$ make style   # reformat R code with styler
$ make lint    # analyze R code with lintr
```

**A pre-commit hook** enforces them. A hook is a
script that git runs automatically at a defined
moment; a pre-commit hook runs before a commit is
recorded, and can refuse the commit if a check
fails. Configuring the linter and styler as a
pre-commit hook means that unstyled or unlinted code
cannot enter the history in the first place, which
is a far stronger guarantee than a convention that
relies on the analyst's memory. The configuration
lives in a `.pre-commit-config.yaml` file at the
project root:

```yaml
# .pre-commit-config.yaml
repos:
  - repo: https://github.com/lorenzwalthert/precommit
    rev: v0.4.3
    hooks:
      - id: style-files
      - id: lintr
      - id: readme-rmd-rendered
      - id: no-browser-statement
```

With this in place, an attempt to commit code that
`styler` would reformat, or that `lintr` flags, or
that contains a stray `browser()` left over from
debugging, is stopped before the commit completes,
and the analyst is prompted to fix it. Continuous
integration, in @sec-verification, runs the same
checks again on the server, so that a collaborator
who has not installed the hook is caught
nonetheless. The hook is the fast local gate; CI is
the authoritative remote one.

::: {.callout-note icon=false collapse=true}
## Check your understanding: capture or validation?

Classify each of the following as belonging to the
capture register or the validation register, and
say what would be lost by removing it. (a) The
`renv.lock` lockfile. (b) A unit test of
`calculate_mean`. (c) The `.pre-commit-config.yaml`
hook. (d) The Dockerfile.

::: {.callout-note icon=false collapse=true}
## Answer

(a) Capture: it fixes the package
versions; removing it drops the analysis from L1 to
L0. (b) Validation: it checks that a captured
function is correct; removing it leaves the analysis
at the same level but now trusted rather than
checked. (c) Validation: it checks that code is
consistently formatted; removing it captures nothing
less but permits legibility to decay. (d) Capture:
it fixes the environment; removing it drops the
analysis from L2. The rule of thumb is that
removing a capture feature lowers the level, while
removing a validation feature leaves the level
unchanged and only stops the checking.
:::
:::

## Testing across the archetypes

The four kinds of test are the same in every
compendium, but their relative weight shifts with
the archetype, and it is worth making the shift
explicit, because a student who has met testing only
in the context of R packages may wrongly conclude
that a data analysis has nothing to test.

Testing is native to the **R-package** archetype.
The package is the setting in which the machinery
grew up: tests live in a conventional directory,
`R CMD check` runs them as a matter of course, and
the whole apparatus of unit testing was designed
around functions with clean inputs and outputs. A
package's suite is dominated by unit tests, because
a package is mostly a collection of functions, and
the universal gate that @sec-compendium described,
`R CMD check`, exercises them on every build.

The **data-analysis** and **simulation** archetypes
are where testing is less conventional and, for that
reason, most in need of emphasis. Such a compendium
may contain few reusable functions and no exported
interface, so the unit-test surface is small. Its
risk lives elsewhere: in the data, which may be
malformed, and in the result, which must
regenerate. Here the data-validation tests and the
reproducibility tests carry the weight. A simulation
archetype in particular leans on the reproducibility
test, because its entire output is a function of a
seed, and a test that fixes the seed and checks the
draw is the guarantee that the reported numbers can
be recovered. The point developed at length in
@sec-verification, that seeding is the load-bearing
determinant of a simulation, has its testing
counterpart exactly here.

The **manuscript** archetype adds one further
concern, that the rendered report itself continues
to build, which sits between testing and the
literate-programming machinery of @sec-literate; a
test that the report renders, and that its headline
numbers match a baseline, guards the paradox at the
level of the whole document.

The **blog** archetype tests the code that its posts
depend on. A teaching blog whose posts contain
runnable examples has a reproducibility obligation
to its readers no smaller than a manuscript's to its
reviewers: a code snippet that no longer runs, or
that now returns a different number than the prose
around it claims, is a published error. Testing the
functions and the example computations that the
posts rely on is how a blog keeps that promise
across the many posts it renders. In every
archetype, then, the tests are drawn from the same
four kinds; what changes is which kinds bear the
load.

## Worked example

We now make the paradox concrete and executable. The
following chunk defines the `calculate_mean` function
in two forms, one correct and one carrying the bug of
the orientation, and checks each against three inputs
whose means are known by hand. Rather than call the
`tinytest` package, which may be absent from a bare R
session, we hand-roll a one-line expectation, which
is all a bare check really is: a comparison of an
actual value against an expected one.

```{r}
#| label: tbl-paradox
#| tbl-cap: 'A correct function and a buggy variant checked against three inputs with hand-computed means. The buggy variant returns the same wrong value on every input and would return it on every machine and every rerun: it is perfectly reproducible and still wrong.'

# the function under test, in a correct and a
# buggy form
calculate_mean <- function(x, na.rm = FALSE) {
  if (na.rm) x <- x[!is.na(x)]
  sum(x) / length(x)
}

calculate_mean_buggy <- function(x) {
  sum(x)                 # bug: division dropped
}

# a minimal hand-rolled expectation: no testing
# package required
expect <- function(actual, expected, tol = 1e-8) {
  isTRUE(abs(actual - expected) < tol)
}

# three inputs whose means are known independently
cases <- list(c(1, 2, 3),
              c(10, 20, 30, 40),
              c(5, 5, 5, 5, 5))
oracle <- c(2, 25, 5)

results <- data.frame(
  input = vapply(cases,
                 function(x) paste(x, collapse = ', '),
                 character(1)),
  expected = oracle,
  correct = vapply(cases, calculate_mean, numeric(1)),
  buggy = vapply(cases, calculate_mean_buggy,
                 numeric(1))
)
results$correct_passes <-
  mapply(expect, results$correct, oracle)
results$buggy_passes <-
  mapply(expect, results$buggy, oracle)

knitr::kable(results)
```

The table records the whole of the lesson. The
correct function passes every check; the buggy
function fails every check, and it fails them
identically. Were we to rerun this chunk a thousand
times, on a thousand machines, the `buggy` column
would hold the same wrong values every time. The bug
is not a source of variability, which reproducibility
tooling might eventually surface; it is a source of
constant, faithful, reproducible error, which only a
comparison against an independent oracle can catch.
That comparison is what a test is.

In a real compendium, the hand-rolled `expect` gives
way to the `tinytest` expectations, and the checks
live in a file under `inst/tinytest/`. The following
is that file, shown as a static listing rather than
run here, and it is exactly the unit test that would
have caught the bug before it reached a result:

```r
# inst/tinytest/test-calculate_mean.R
expect_equal(calculate_mean(c(1, 2, 3)), 2)
expect_equal(calculate_mean(c(10, 20, 30, 40)), 25)
expect_equal(calculate_mean(rep(5, 5)), 5)
expect_true(is.nan(calculate_mean(numeric(0))))
expect_equal(calculate_mean(c(1, NA, 3),
                            na.rm = TRUE), 2)
```

The runner `tests/tinytest.R` that `R CMD check`
invokes is a three-line file:

```r
# tests/tinytest.R
if (requireNamespace('tinytest', quietly = TRUE)) {
  tinytest::test_package('mycompendium')
}
```

and the pre-commit configuration shown earlier
completes the picture: the linter and styler guard
the code's legibility on every commit, the tinytest
suite guards its correctness, and continuous
integration re-runs both on a clean machine so that
neither guarantee depends on the author's local
setup.

## Collaborating with an LLM

A language model is a capable assistant for writing
tests, and a dangerous one for a specific reason: it
will readily produce tests that pass, which is not
the same as tests that are correct. The analyst's
verification duty is heaviest exactly where the
model is most fluent.

**Prompt.** 'Here is my `calculate_mean` function.
Write `tinytest` unit tests for it, including edge
cases.'

*Watch for.* A model tends to derive the expected
values by running the function it was given, rather
than by independent computation. If the function is
buggy, the model may generate an oracle that
matches the bug, producing tests that pass against
wrong behavior and thereby certifying the error.
It may also test only the paths the code obviously
takes and miss the empty vector or the missing
value.

*Verification.* Check every expected value by hand
or from an independent formula, not by running the
function. The oracle must come from outside the code
under test; that is the whole point of an oracle.
Confirm that the edge cases you care about, the
empty input, the single value, the missing value,
each have a test.

**Prompt.** 'Write data-validation tests for this
cohort data frame.'

*Watch for.* The model will propose generic checks,
non-missing identifiers, numeric outcomes, that are
reasonable but do not encode *your* domain. It does
not know that your outcome is a probability bounded
in the unit interval, or that your ages were
recorded in years, or which treatment labels are
admissible.

*Verification.* Supply the domain constraints
yourself, or check that the generated ranges match
your data dictionary. A validation test asserting a
plausible-looking but wrong range is worse than no
test, because it lends false authority.

**Prompt.** 'Explain why this test is failing.'

*Watch for.* The model may recommend changing the
test to match the current output, which makes the
failure disappear without establishing whether the
code or the test was wrong. A failing test is
sometimes correct: it may be reporting a real bug.

*Verification.* Decide first, from the oracle,
whether the code or the expectation is at fault, and
only then change the one that is wrong. Never
silence a failing test by editing its expected value
to whatever the code currently returns; that
converts a test into a tautology.

## Exercises

1. Take a function you have written for a current
   analysis and write three `tinytest` unit tests
   for it: one ordinary case with a hand-computed
   oracle, and two edge cases. State, for each, where
   the expected value came from.

2. Introduce the `calculate_mean` bug of this chapter
   into a copy of one of your own summary functions,
   confirm that the analysis still runs and produces
   a stable result, and then write the single unit
   test that would have caught it. Reflect in a
   sentence on why reproduction did not catch it.

3. Write a data-validation test file for a dataset
   you use, encoding at least four domain constraints:
   an expected set of columns, a plausible range for a
   continuous variable, an admissible set of labels
   for a categorical variable, and a bound on the
   missing-data rate. Justify each range from a data
   dictionary or from clinical knowledge.

4. Classify each test you wrote in exercises 1 and 3
   as belonging to the capture register or the
   validation register, and state what removing it
   would and would not change about the analysis's
   position on the L0 to L3 ladder.

5. Write a reproducibility test of the baseline form:
   run a small analysis, save its result with
   `saveRDS`, and write a test that recomputes the
   result and compares it against the saved baseline
   with an appropriate tolerance. Then change one line
   of the analysis and observe the test fail.

6. Install the `precommit` package, add the
   `.pre-commit-config.yaml` of this chapter to a
   project, and attempt to commit deliberately
   misformatted R code. Report what the hook does, and
   contrast it with what continuous integration would
   do in @sec-verification if the hook were absent.

## Solutions

Attempt an exercise before expanding its solution. These
exercises are performed on the reader's own functions and
data, so the code will vary; the boxes record the form the
answer should take and the reasoning behind it.

::: {.callout-note icon=false collapse=true}
## Exercise 1

The form of the answer matters more than the function
chosen. For a hypothetical `summarize_visits()`:

```r
# inst/tinytest/test-summarize_visits.R
expect_equal(summarize_visits(c(2, 4, 6)), 4)
expect_true(is.nan(summarize_visits(numeric(0))))
expect_error(summarize_visits('a'))
```

**Where each expected value came from** is the substance of
the exercise, and the three answers should be different.

The ordinary case comes from **arithmetic done by hand**,
independently of the function. This is what makes it an
oracle: if the expected value were computed by running the
function, the test would assert only that the function
agrees with itself, which is the commonest way a test
suite becomes worthless.

The empty-input case comes from **a decision about the
contract**. There is no arithmetic to do; the analyst must
decide what the function should do with no data and then
encode the decision. The test is as much a specification
as a check.

The wrong-type case comes from **the same contract**,
asserting that the function refuses input it cannot
handle rather than returning something misleading.

A reader who finds they cannot state where an expected
value came from has usually written a change-detector
rather than a test.
:::

::: {.callout-note icon=false collapse=true}
## Exercise 2

The bug is the dropped division, so the function returns a
sum where a mean is intended:

```r
summarize_visits <- function(x) sum(x)   # bug
```

The analysis still runs, produces a number, and produces
the *same* number on every machine and every rerun. If the
project is containerized and its packages pinned, that
number is perfectly reproducible.

The catching test is the ordinary case from exercise 1:

```r
expect_equal(summarize_visits(c(2, 4, 6)), 4)
```

It fails immediately, returning 12 against an expected 4.

**Why reproduction did not catch it.** Because reproduction
asks whether the same inputs give the same outputs, and
they do. The buggy function is deterministic; nothing about
it varies between machines or over time. Reproducibility
machinery captures determinants and propagates whatever
the code computes with mechanical fidelity, including the
wrong thing. The only way to detect the error is to compare
the output against a value obtained *independently of the
code*, which is what the hand-computed oracle supplies and
what no amount of environment capture can.

This is the reproducibility paradox in one line: capture
fixes the inputs, and validation checks that the fixed
inputs produce the intended answer.
:::

::: {.callout-note icon=false collapse=true}
## Exercise 3

The four constraints, in the form the chapter uses:

```r
# inst/tinytest/test-data-cohort.R
cohort <- readRDS('analysis/data/derived_data/cohort.rds')

expect_true(
  all(c('id', 'age', 'sex', 'outcome') %in% names(cohort)),
  info = 'the expected columns are present'
)
expect_true(
  all(cohort$age >= 18 & cohort$age <= 105, na.rm = TRUE),
  info = 'age lies within the enrolled range'
)
expect_true(
  all(cohort$sex %in% c('F', 'M', 'Other')),
  info = 'sex takes only coded levels'
)
expect_true(
  mean(is.na(cohort$outcome)) < 0.05,
  info = 'outcome missingness stays under five percent'
)
```

**The justifications** are the point, and each should come
from outside the data.

The **column set** comes from the data dictionary: these
are the variables the analysis consumes, and a missing one
means an upstream export changed. The **age range** comes
from the protocol's eligibility criteria, not from the
observed minimum and maximum; deriving the bound from the
data would make the test unable to detect the very
excursion it exists to catch. The **label set** comes from
the coding sheet. The **missingness bound** comes from
what the analysis can tolerate before its conclusions
become unsafe, which is a judgment the analyst must make
and record.

The recurring discipline: a range read off the current
data is a description, and a range read off the protocol
is a test.
:::

::: {.callout-note icon=false collapse=true}
## Exercise 4

**All of them belong to the validation register.** Unit
tests and data-validation tests alike pin no determinant:
they add no package to the lockfile and no library to the
image. They confirm that what was captured behaves as
intended.

**What removing them would change.** Nothing about the
position on the L0 to L3 ladder. An analysis at L2 with a
full test suite and the same analysis at L2 with no tests
sit on the same rung, because the rung grades how far down
the determinant stack the result is locked, and tests lock
nothing.

**What removing them would change instead** is whether
anyone has reason to believe the result is right. The
analysis becomes trusted rather than checked, and the
class of error the tests were catching, a wrong function
and an out-of-range dataset, returns to being invisible.

The one qualification worth making, because it is a real
subtlety rather than a hedge: removing tests does bear on
L3 indirectly. L3 requires evidence that recorded outputs
regenerate, and a test suite is one of the mechanisms that
produces such evidence. A project relying on its tests as
its L3 evidence would fall to L2 on their removal. It is
the *verification role* that matters there, not the
testing role, and a reader who makes this distinction has
understood the chapter's central division.
:::

::: {.callout-note icon=false collapse=true}
## Exercise 5

```r
# once, to record the baseline
set.seed(42)
result <- fit_model(cohort)
saveRDS(coef(result), 'inst/tinytest/baseline-coefs.rds')
```

```r
# inst/tinytest/test-reproducibility.R
baseline <- readRDS('baseline-coefs.rds')
set.seed(42)
current <- coef(fit_model(cohort))
expect_equal(current, baseline, tolerance = 1e-8,
             info = 'model coefficients match the baseline')
```

Changing one line of the analysis makes the test fail,
reporting the coefficients that moved.

**On the tolerance**, which is the part of this exercise
that repays thought. An exact comparison is the strongest
check and will produce failures that mean nothing:
floating-point summation reorders under a different BLAS
or a different thread count, as @sec-containers discusses,
and the last digits move without the analysis changing. A
tolerance of `1e-8` accepts that noise and still catches
any change large enough to alter a reported figure. A
tolerance set too loosely stops detecting real changes,
and there is no universally correct value; it should be
chosen against the precision at which results are
reported.

**What this test is and is not.** It is the single-result
form of the L3 check, and it is a *change detector*: it
asserts that today's answer matches the recorded one, not
that either is correct. It is worth being clear about
this, because the baseline was produced by the same code
it now guards. If the code was wrong when the baseline was
recorded, this test will faithfully defend the wrong
answer, which is the paradox reappearing one level up.
The unit tests of exercise 1 are what establish
correctness; this test establishes stability.
:::

::: {.callout-note icon=false collapse=true}
## Exercise 6

**What the hook does.** The commit is refused. The
formatting hook rewrites the offending files in place and
aborts, so the commit does not complete and the working
tree now contains the reformatted code, which must be
staged and committed again. The effect is that
misformatted code cannot enter the history at all.

**The contrast with continuous integration.** CI would let
the commit through. The code enters the history, the push
succeeds, and the check runs afterward on the server,
reporting a failure against a commit that already exists.
Fixing it requires a second commit, so the history now
records both the violation and its repair.

The differences worth naming are three. **Timing:** the
hook acts before the record is written, CI after.
**Locality:** the hook runs on the analyst's machine and
can therefore be skipped, with `--no-verify` or by a
collaborator who never installed it, whereas CI runs on
the server and cannot. **Scope:** the hook must be fast
enough to sit inside a commit, so it does cheap things
like formatting and linting, whereas CI can rebuild the
container and run the full suite.

The two are complements rather than alternatives, and the
chapter's framing explains why: the hook is a convenience
that keeps trivia out of the history, and CI is the gate
that actually guarantees anything, because it is the one
nobody can skip.
:::

## Further reading

The single most important reference for this chapter
is the short note by Leek and Peng, that reproducible
research can still be wrong [@leek2015], which states
the paradox that motivates every test in a compendium
and argues for a prevention-oriented rather than a
correction-oriented stance toward analytical error.

For the testing tool itself, van der Loo's
documentation of `tinytest` [@vanderloo2020] explains
the design rationale for a zero-dependency,
bare-expectation framework, and is worth reading for
the argument that a testing framework should not
itself be a heavy dependency.

For the practice of testing and code quality as
components of scientific computing rather than as
software-engineering appendages, the two papers by
Wilson and colleagues are the standard references:
the best-practices paper [@wilson2014] and the
good-enough-practices paper [@wilson2017], the latter
pitched deliberately at researchers without a
software background. Sandve and colleagues' ten
simple rules [@sandve2013] cover the same ground in a
compact, rule-by-rule form, and situate readable code
and recorded results within a general reproducibility
discipline.

For the connection between testing and the highest
rung of the ladder, the reader is referred forward to
@sec-verification, where the reproducibility test of
this chapter reappears at the scale of the whole
compendium as output verification, and to @sec-levels
for the capture-and-validation distinction that places
testing in the validation register. Peng's early
account of reproducibility as a spectrum [@peng2011]
remains the clearest short statement of why confirming
a result, and not merely producing it, is a minimum
scientific standard.
