# Version Control with Git and GitHub {#sec-version-control}

::: {.bookepigraph}
```{=latex}
\begin{flushright}\small
```
*Maintaining a high degree of transparency in scientific
reporting is essential not just for gaining trust and
credibility within the scientific community but also for
facilitating the development of new ideas.*

[Karthik Ram]{.smallcaps}, *Git Can Facilitate Greater
Reproducibility and Increased Transparency in Science*
(2013)
```{=latex}
\end{flushright}
```
:::

## Learning objectives

By the end of this chapter the reader should be able
to:

- Describe the failure mode that version control
  cures, the proliferation of files named
  `analysis_final_v2_FINAL.R` and the loss of any
  record of what changed and why.
- Explain the core git model: the repository, the
  commit as an atomic, message-bearing unit of change,
  and the commit history as an auditable record of how
  a result came to be.
- Perform the everyday git operations at the shell:
  staging and committing, reading the log and the
  diff, and using a `.gitignore` to keep files out of
  the history.
- Exercise the health-specific judgment of what must
  never be committed, above all protected health
  information and identifiable data, and understand why
  deleting such a file in a later commit does not
  remove it from the history.
- Write an archetype-appropriate `.gitignore`, and say
  why a data-analysis compendium and an R-package
  compendium ignore different files.
- Use a remote (GitHub or GitLab) for backup,
  distribution, and collaboration, including the
  fork-and-pull-request model for multi-site teams and
  tags and releases for marking the exact state of an
  analysis at submission or publication.
- Place version control on the reproducibility ladder,
  as the artifact that first makes an analysis L0
  locatable and the substrate on which continuous
  integration and collaboration are later built, and
  state plainly the layers of the determinant stack
  that git does not pin.

## Orientation

Open the folder of almost any analysis that has run
for more than a month without version control and the
same artifact greets you. There is `analysis.R`, and
beside it `analysis_v2.R`, `analysis_final.R`,
`analysis_final_v2.R`, and, inevitably,
`analysis_final_v2_FINAL.R`. A dated copy or two,
`analysis_2025-03-14.R`, sits among them, and a
`analysis_reviewer.R` from the week the referee
reports came back. Every file is a manual snapshot,
taken because the analyst feared to overwrite a
version that worked. The scheme fails in three ways at
once. It does not say which file produced the figure
in the submitted manuscript. It does not say what
differs between any two of them, so recovering a change
means reading both side by side. And it does not say
*why* any change was made, because the reason lived in
the analyst's head and is now gone.

This is the failure that version control exists to
cure, and the cure is not a tidier naming scheme but a
different object. A version-control system keeps a
single working copy of each file and records, as a
separate and complete history, every state that copy
has passed through, together with a note on each change
saying what it was and why it was made. The five
`FINAL` files collapse back into one `analysis.R`, and
the four snapshots become four entries in a history
that can be read, searched, and, if need be, restored.
The tool that has won this role, across the sciences
and across software of every kind, is git, and the
service that most often hosts a git history on the
network is GitHub, with GitLab its principal
institutional sibling. This chapter teaches both from
first principles, because version control is the single
most universally taught element of a modern
reproducibility course and the foundation on which the
rest of this book stands.

It is also the foundation in a precise, technical
sense. @sec-compendium made the compendium the unit of
reproducible work, and @sec-levels placed the bare
compendium at the bottom rung of the ladder, L0
locatable. The artifact that marks L0 is exactly a
version-controlled repository: an analysis is locatable
when, and only when, it is under version control and
can therefore be found, read, and referred to by a
second person. Everything the later chapters add, the
lockfile of @sec-dependencies, the container of
@sec-containers, the continuous integration of
@sec-verification, is installed on top of a
version-controlled history. Version control is thus not
one tool among many but the substrate the others
require.

## The analyst's contribution

Version control automates the mechanics of history
perfectly and automates none of the judgment that
makes a history useful and safe. Three decisions belong
to the analyst and are the subject of this chapter as
much as the commands are.

1. **Deciding what must never be committed.** A commit
   is, by design, permanent and copied to every clone
   of the repository. The judgment of what may safely
   be made permanent and public, and what must be kept
   out of the history entirely, is the analyst's, and
   in health research it is a judgment with legal and
   ethical force. Protected health information,
   identifiable data, and credentials must never enter
   the history, because, as we shall see, deleting them
   in a later commit does not remove them from it. This
   is the first and gravest of the three judgments.

2. **Choosing the granularity and the message of a
   commit.** A commit is a unit of change, and how much
   change it should contain, and what its message
   should say, are matters of craft that no tool
   decides. A history of one enormous commit reading
   'did the analysis' is almost as useless as no
   history at all. A history of small, coherent commits
   whose messages say why each change was made is an
   audit trail. The difference is entirely the
   analyst's discipline.

3. **Deciding when to branch and when to tag.** Whether
   a piece of work warrants a branch of its own, and
   whether a particular state of the analysis warrants a
   permanent tag, are design decisions about the shape
   of the history. They are cheap to make and expensive
   to omit: the state of the code at the moment a
   manuscript was submitted is worth a tag, and no tool
   will place it for you.

::: {.callout-note}
## The vocabulary of this chapter

New terms are defined again on first use; they are
collected here for reference.

- **Repository (repo).** A directory whose contents
  and full history git tracks, held in a hidden
  `.git` subdirectory. A *clone* is a complete copy of
  a repository, history and all.
- **Working tree.** The ordinary files in the
  repository directory as they currently exist on
  disk, as opposed to the versions recorded in the
  history.
- **Commit.** A recorded snapshot of the tracked files
  at one moment, carrying an author, a timestamp, a
  message, and an immutable identifier (a *hash*). The
  atomic unit of history.
- **Staging area (index).** The set of changes marked,
  with `git add`, to go into the next commit. Staging
  lets a commit hold a chosen subset of the current
  changes.
- **`.gitignore`.** A file listing path patterns that
  git should not track, used to keep derived outputs,
  large data, and, critically, sensitive files out of
  the history.
- **Branch.** A movable name for a line of
  development. The default branch is conventionally
  called `main`.
- **Remote.** A copy of the repository hosted
  elsewhere, typically on GitHub or GitLab, used for
  backup, distribution, and collaboration. `origin` is
  the conventional name of the primary remote.
- **Tag.** A permanent, human-readable name for one
  specific commit, used to mark a milestone such as a
  manuscript submission.
- **Pull request (merge request).** A proposal, made on
  the remote, to merge one branch or fork into another,
  reviewed before it is accepted.
:::

## What version control records, and what it cures

The value of a version-control system is easiest to
state as the negation of the `FINAL`-file failure. That
failure loses three things, and version control
restores each.

It loses the *identity of the working version*. With a
drawer of dated copies, no file is authoritative; with
git, there is one `analysis.R`, and its history holds
every prior state, so the question 'which version is
current' cannot arise. It loses the *content of each
change*. Two snapshot files differ in ways that can be
recovered only by comparing them line by line by hand;
a git history presents each change as a *diff*, a precise
statement of the lines added and removed, that git
computes on demand from the snapshots it stores, since
git keeps full snapshots, not diffs. And it loses the *reason for each
change*, the single most valuable and most perishable
piece of information about an analysis. Why was the
outlier-exclusion threshold changed from three standard
deviations to two? A dated file cannot say. A commit
whose message reads 'Tighten outlier threshold to 2 SD
per reviewer 2's comment' says exactly, and says it
forever.

This last point is worth dwelling on, because it is the
one that turns version control from a convenience into
an instrument of scientific accountability. A
computational result is not only a number; it is the
end of a chain of decisions, each of which could have
gone otherwise. In a regulated analysis, a clinical
trial subject to audit, a study destined for a
guideline, the chain of decisions is itself an object
of scrutiny, and 'why did the estimate change between
the interim and the final analysis' is a question an
auditor is entitled to ask and the analyst is obliged
to answer. A commit history in which each change carries
its reason is the answer, already written, timestamped,
and attributable. The history is the audit trail.

## The git model: repository, commit, history

Git is often taught as a list of commands to memorize,
which is why it is so often found bewildering. It is far
better understood as a small and coherent data model,
after which the commands are obvious. There are three
ideas.

The first is the **repository**. A git repository is an
ordinary directory, your project, that git has been
told to track, whereupon it maintains, in a hidden
`.git` subdirectory, the complete history of everything
under that directory. The ordinary files you edit are
the *working tree*; the `.git` directory is the history.
Creating a repository is a single command run once in
the project directory.

```bash
cd diabetes-cohort
git init
```

The second idea, and the central one, is the **commit**.
A commit is a snapshot of the tracked files at one
moment, recorded together with four things: who made it,
when, a message describing it, and a hash that names it.
That hash, a forty-character SHA-1 string usually shown
in an abbreviated form such as `a1b2c3d`, is computed not
from the snapshot alone but over the whole commit object,
which references the snapshot, folds in the metadata,
and, crucially, records the hash of the commit before it.
(Git also supports repositories that use the longer
SHA-256 hash of sixty-four characters, though SHA-1
remains the default.) Because each commit's hash
incorporates its parent's, the history is a chain in
which no past entry can be altered without changing its
hash and the hash of every commit after it. This is what
makes a git history *tamper-evident*, a property of some
consequence for a regulated analysis: the record cannot
be quietly rewritten. The hashes are, for practical
purposes, unique, and git guards against the known SHA-1
collision attacks with built-in collision detection.

The third idea is the **history** itself, the chain of
commits, each pointing back to its parent, running from
the first commit to the most recent. Reading that chain
is reading the story of how the analysis came to its
current state. A branch, which we treat below, is
nothing more than a movable name pointing at one commit
in this structure.

The everyday cycle of work has just two steps against
this model, and they exist because git deliberately
separates *choosing* what goes into a commit from
*making* the commit. First one *stages* the changes to
be recorded, with `git add`; then one *commits* the
staged changes, with `git commit`, attaching the
message. The staging step is not ceremony: it lets a
single commit hold a coherent subset of the work in
progress, so that an afternoon's editing across several
concerns can be recorded as several focused commits
rather than one indiscriminate lump.

```bash
# edit analysis.R, then see what has changed
git status

# stage the file for the next commit
git add analysis.R

# record the staged change with a message
git commit -m "Add Cox model for 30-day readmission"
```

A note on the message, since it is the part that carries
the scientific value. The convention, and it is worth
adopting from the first day, is a short imperative
summary line of about fifty characters, optionally
followed by a blank line and a paragraph giving the
*why*. 'Fix bug' is a wasted message. 'Correct
person-time denominator to exclude pre-enrollment
follow-up' is a message that will still be intelligible,
and still exculpatory, at an audit three years later.

::: {.callout-note icon=false collapse=true}
## Check your understanding: why stage at all?

A colleague argues that the staging area is needless
friction: since one almost always wants to commit
everything one has changed, git should simply commit the
whole working tree and spare the analyst a `git add`.
What does the staging step buy that a commit-everything
model would lose?

::: {.callout-note icon=false collapse=true}
## Answer

It buys the ability to make a commit an *atomic unit of
one idea* rather than a snapshot of whatever happened to
be on disk. In an afternoon an analyst may fix a bug in
the cleaning script, add an unrelated figure, and jot a
half-finished note in a third file. Committing all three
together produces a history entry that cannot be
described in one honest message and cannot be reverted
without disturbing unrelated work. Staging lets the
analyst record the bug fix as one commit, with its own
message and reason, and the figure as another, leaving
the unfinished note uncommitted. The audit trail is only
as good as the coherence of its individual commits, and
staging is the mechanism that coherence requires. (For
the genuine everyday case where one *does* want
everything, `git commit -a` stages all tracked changes
in one step, so the friction is opt-out, not mandatory.)
:::
:::

## Reading the history: the log and the diff

A history that cannot be read is of little use, and git
provides two complementary views of it. The **log** is
the list of commits, most recent first, each with its
hash, author, date, and message. It is the table of
contents of the analysis's development.

```bash
# the full log
git log

# a compact one-line-per-commit view
git log --oneline

# which commits touched analysis/scripts/
git log --oneline -- analysis/scripts/
```

The **diff** is the complementary view: not the list of
changes but the content of one. `git diff` shows the
changes in the working tree that have not yet been
staged; `git show` displays the full diff of any single
commit named by its hash; and `git diff` between two
commit hashes shows everything that changed between them.

```bash
# what have I changed but not yet staged?
git diff

# show one commit's changes in full
git show a1b2c3d

# what changed between two points in the history?
git diff a1b2c3d f4e5d6c
```

Together the log and the diff answer the two questions
the `FINAL`-file drawer could not: *when and why* did a
thing change (from the log) and *what exactly* did the
change consist of (from the diff). To make this concrete,
consider the commit history of the first weeks of an
incident-diabetes cohort study, rendered below as it
might appear to an auditor reviewing how the reported
hazard ratio came to be. The table is generated from
synthetic data, so that we may display a realistic audit
trail without any real project's contents.

```{r}
#| label: tbl-audit-trail
#| tbl-cap: 'A synthetic commit history for the first weeks of a cohort analysis, read as an audit trail. Each row is one atomic, message-bearing commit; together they record not only what changed but, in the message, why. Note the reviewer-driven change to the outlier threshold, the kind of decision an auditor is entitled to trace.'

library(tibble)
library(dplyr)
library(knitr)

commit_history <- tibble::tibble(
  hash = c('a1b2c3d', 'e4f5a6b', 'c7d8e9f',
           'b0a1c2d', 'd3e4f5a', 'f6a7b8c',
           'a9b0c1d'),
  date = as.Date(c('2026-02-03', '2026-02-05',
                   '2026-02-09', '2026-02-12',
                   '2026-02-18', '2026-02-24',
                   '2026-03-02')),
  author = c('r.thomas', 'r.thomas', 'r.thomas',
             'a.okafor', 'r.thomas', 'a.okafor',
             'r.thomas'),
  message = c(
    'Initialise compendium and .gitignore',
    'Add cleaning script for registry extract',
    'Build analysis-ready cohort table',
    'Fit Cox model for 30-day readmission',
    'Tighten outlier threshold to 2 SD per reviewer 2',
    'Add Kaplan-Meier figure to report',
    'Correct person-time denominator'),
  files_changed = c(2L, 1L, 3L, 1L, 1L, 2L, 1L)
)

commit_history |>
  dplyr::mutate(date = format(date, '%Y-%m-%d')) |>
  knitr::kable(
    col.names = c('Hash', 'Date', 'Author',
                  'Message', 'Files'),
    align = c('l', 'l', 'l', 'l', 'r'))
```

The reader will note two features that a bare list of
files could never carry. The fifth commit records not
only that the outlier threshold changed but that it
changed *in response to a reviewer*, and the final
commit records a correction to the person-time
denominator, the sort of methodological fix whose
provenance an auditor will want to establish. The commit
history has turned a sequence of edits into a
defensible account of how the result was reached.

## What must never be committed: `.gitignore` and PHI

A commit records the files git is tracking, and by
default git offers to track every file in the working
tree. Two categories of file must be kept out of that
net, and the mechanism for keeping them out is the
`.gitignore` file, a plain-text list of path patterns
that git will refuse to track.

The first category is the merely unwanted: files that
*can* be committed but *should not* be, because they are
regenerable or bulky and would only clutter the history.
Rendered outputs (a compiled PDF, an `.html` report, the
figures a script regenerates on every run) are the
clearest case, because committing them couples the
history to products that the code already determines.
Large raw data files bloat every clone. Editor and
operating-system cruft (`.Rhistory`, `.DS_Store`,
`.Rproj.user/`) is pure noise. None of these is
dangerous; all of them are better ignored.

The second category is the category that makes this
section a matter of ethics and law rather than tidiness:
files that must *never* enter the history because making
them permanent and shareable would be a breach. In
health research three kinds dominate.

- **Protected health information and identifiable
  data.** A file containing patient identifiers, dates
  of birth, medical-record numbers, free-text clinical
  notes, or any data from which an individual could be
  re-identified must never be committed. This is the
  first rule of version control in a health setting, and
  it admits no exception.
- **Credentials.** Database passwords, API tokens, the
  contents of a `.Renviron` holding a secret, private
  keys: a credential committed to a repository, and
  worse pushed to a remote, is a credential to be
  treated as compromised.
- **Restricted or embargoed data** more generally, under
  a data-use agreement that forbids redistribution, even
  when not directly identifying.

::: {.callout-important}
## Deleting a file does not remove it from the history

The gravest and least intuitive fact about git is that
its history is *append-only by design*. If a file
containing protected health information is committed,
and the mistake is noticed and the file deleted in a
later commit, the file is gone from the working tree but
*remains in full in the history*, recoverable by anyone
with a clone from the commit that introduced it. The
deletion is itself just another commit layered on top;
it conceals nothing. Removing the file from the history
requires rewriting the history, with tools such as `git
filter-repo`, which changes every subsequent commit
hash, must be forced onto every clone and the remote,
and is error-prone precisely when it matters most.
Worse, if the offending commit was ever pushed to a
remote, one must assume it has been cloned and cached
beyond recall, and the only sound response is to treat
the exposed data as breached and follow the applicable
incident procedure. The remedy is therefore prevention:
a correct `.gitignore` written *before* the first
commit, and the habit of reading `git status` before
every commit to confirm that nothing sensitive has been
staged. Never commit PHI, because git is built so that
you cannot cleanly take it back.
:::

A `.gitignore` is a small file, and its content depends
on the archetype of the compendium, which is the
concrete point at which the archetypes of @sec-compendium
touch this chapter. The reproducibility machinery is
identical across archetypes; the list of files to ignore
is not. A **data-analysis compendium** ignores its
*derived outputs and its raw data*: the regenerable
tables and figures, and the large or sensitive registry
extract that must stay out of the history entirely. An
**R-package compendium** has no `analysis/` outputs to
ignore but must ignore its *build artifacts*: the
`.Rcheck` directory, the built `.tar.gz` and binary
archives, and the `src/` object files a compiled package
leaves behind. The two `.gitignore` files barely
overlap, and writing the right one is part of naming the
archetype.

```bash
# .gitignore for a data-analysis compendium
# --- sensitive: never commit ---
analysis/data/raw_data/*
.Renviron
*.pem

# --- regenerable outputs ---
analysis/figures/*
analysis/tables/*
analysis/report/*.pdf
analysis/report/*.html

# --- editor and OS cruft ---
.Rhistory
.RData
.Rproj.user/
.DS_Store
```

```bash
# .gitignore for an R-package compendium
# --- build artefacts ---
*.tar.gz
*.Rcheck/
src/*.o
src/*.so
docs/            # generated pkgdown site

# --- editor and OS cruft ---
.Rhistory
.RData
.DS_Store
```

Two refinements deserve mention. A leading `!` re-includes
a pattern otherwise ignored, which is how one ignores a
whole data directory but keeps a small, non-sensitive
example file or a data dictionary under version control
(`analysis/data/raw_data/*` followed by
`!analysis/data/raw_data/data_dictionary.csv`). And a
committed-by-accident file, once added to `.gitignore`,
is still tracked until explicitly untracked with `git rm
--cached`; ignoring a file does not retroactively remove
an already-committed copy, which returns us to the
warning above.

::: {.callout-note icon=false collapse=true}
## Check your understanding: the same file, two archetypes

A `report.pdf` sits at the top of one analyst's ignore
list and is deliberately committed by another. A
`data_dictionary.csv` describing a restricted registry
is committed, while the registry extract it describes is
ignored. Explain both choices.

::: {.callout-note icon=false collapse=true}
## Answer

The rendered `report.pdf` is a regenerable output: in a
data-analysis or manuscript compendium it is a product
the code reproduces on render, so committing it couples
the history to a derivative and is usually avoided. An
analyst who nonetheless commits it, perhaps to preserve
the exact submitted PDF as a fixed reference, is making a
defensible different judgment about what the artifact is
*for*. The point is that it is a judgment, not a rule.
The `data_dictionary.csv` and the registry extract sit on
opposite sides of the one line that is not a matter of
taste. The dictionary carries variable names and codings,
which are metadata and safe to share, and belongs in the
history so a reader can interpret the analysis. The
extract carries the protected health information itself
and must never be committed, whatever its convenience.
The `!` re-inclusion pattern exists precisely to keep the
safe metadata while ignoring the unsafe data beside it.
:::
:::

## Branches, lightly

For a solo analyst working linearly, a single line of
commits on the `main` branch is often the whole story,
and it is honest to say so rather than to impose ceremony.
But two situations make branches worth the small
investment of learning them. The first is *speculative
work*: trying an alternative model specification, or a
reviewer's proposed reanalysis, without disturbing the
`main` line that produced the current results. The second
is *collaboration*, treated in the next section, where
each contributor works on a branch of their own and the
branches are merged only when ready.

A branch is created and switched to in one command, and
merged back with another.

```bash
# create and switch to a branch for a reanalysis
git switch -c sensitivity-analysis

# ... make commits on the branch ...

# return to main and merge the branch in
git switch main
git merge sensitivity-analysis
```

The mental model is the one given above: a branch is a
movable name for a line of development, and `main` is
merely the branch git creates by default. Nothing about
branching changes the commit model; it only lets two
lines of commits coexist and later rejoin. We keep the
treatment light here and return to the collaborative use
of branches, which is where they earn their keep, when we
reach pull requests.

@fig-branch draws the shape. The point worth taking from
it is that the branch does not copy the history: it names
a new commit whose parent is a commit on `main`, and the
merge names a commit with two parents. Everything before
the branch point is shared, and no hash anywhere in the
diagram changes when the branch is created or merged.

```{r}
#| label: fig-branch
#| fig-cap: 'A branch and its merge, drawn as the commit graph git actually stores. The lower line is `main`; the upper line is a sensitivity-analysis branch created at commit B and merged back at M. Commits A through C exist once and are shared by both lines: the branch adds commits rather than copying them. The merge commit M is the only commit in the diagram with two parents, which is what records that the two lines rejoined and when.'
#| fig-alt: 'A commit graph with two horizontal rows of circles joined by lines, with time running left to right. The lower row, labeled main, runs A, B, C, then the merge commit M, then D. The upper row, labeled a sensitivity-analysis branch, holds two commits S1 and S2. A line rises from B to S1, marking the branch point, and a line descends from S2 to M, the merge. M is annotated as the only commit with two parents. Commits A, B, and C lie on the shared portion of the graph rather than being copied onto the branch.'
#| fig-width: 7
#| fig-height: 2.5

library(ggplot2)
library(tibble)

nodes <- tibble(
  id    = c('A', 'B', 'C', 'M', 'D', 'S1', 'S2'),
  x     = c(1, 2, 3, 4.6, 5.7, 3.1, 4.0),
  y     = c(0, 0, 0, 0, 0, 0.8, 0.8),
  shade = c(rep('main', 5), 'branch', 'branch')
)

edges <- tibble(
  x    = c(1, 2, 3, 4.6, 2, 3.1, 4.0),
  xend = c(2, 3, 4.6, 5.7, 3.1, 4.0, 4.6),
  y    = c(0, 0, 0, 0, 0, 0.8, 0.8),
  yend = c(0, 0, 0, 0, 0.8, 0.8, 0)
)

ggplot() +
  geom_segment(data = edges,
               aes(x = x, y = y, xend = xend, yend = yend),
               color = 'grey45', linewidth = 0.45) +
  geom_point(data = nodes, aes(x = x, y = y, fill = shade),
             shape = 21, size = 9, stroke = 0.5,
             color = 'grey30', show.legend = FALSE) +
  geom_text(data = nodes, aes(x = x, y = y, label = id),
            size = 3, fontface = 'bold') +
  scale_fill_manual(values = c(main = '#c3ddda',
                               branch = '#f0e2c4')) +
  annotate('text', x = 0.55, y = 0, label = 'main',
           hjust = 1, fontface = 'bold', size = 3.2) +
  annotate('text', x = 2.55, y = 0.8, label = 'branch',
           hjust = 1, fontface = 'bold', size = 3.2) +
  annotate('text', x = 4.6, y = -0.52,
           label = 'M: two parents', size = 2.9,
           color = 'grey25') +
  annotate('text', x = 2.0, y = -0.52,
           label = 'branch point', size = 2.9,
           color = 'grey25') +
  coord_cartesian(xlim = c(-0.1, 6.1), ylim = c(-0.8, 1.1)) +
  theme_void()
```

## The remote: GitHub, GitLab, and collaboration

Everything so far is *local*: the repository, its
history, its branches all live in the `.git` directory on
one machine, and if that machine is lost, so is the work.
A **remote** is a copy of the repository hosted elsewhere,
on GitHub, on GitLab, or on an institution's own server,
and connecting a local repository to a remote serves three
distinct purposes that are worth separating.

The first is **backup**. A remote holds a complete copy of
the history, so that a lost laptop costs at most the
commits made since the last push. The second is
**distribution**. A public remote is the mechanism by
which a compendium becomes locatable to the world: a URL
a reviewer can be given, a repository a reader can clone,
the concrete form of the L0 rung of the ladder. The third
is **collaboration**, the purpose that shapes the rest of
this section.

The commands that connect to and synchronize with a
remote are few. One names the remote once, then pushes
local commits to it and pulls others' commits from it.

```bash
# connect the local repo to a remote named origin
git remote add origin https://github.com/lab/diabetes-cohort.git

# push the main branch and remember the link
git push -u origin main

# later: send new local commits up
git push

# and bring collaborators' commits down
git pull
```

For a team working across sites, a clinical trial with
several participating centers, a multi-institution
consortium, the dominant model on GitHub and GitLab is
**fork-and-pull-request** (GitLab calls the latter a
*merge request*, the same thing under another name). A
contributor *forks* the canonical repository, making a
personal copy on the remote, works on a branch of that
fork, and then opens a **pull request** proposing that
their branch be merged into the canonical repository. The
pull request is a unit of review: the maintainers see
exactly the diff being proposed, discuss it inline,
request changes, and merge it only when satisfied. For a
regulated multi-site analysis this is more than
convenience. The pull request is a documented,
attributable, reviewable gate through which every change
to the shared analysis passes, and its record, who
proposed what, who reviewed it, what was said, and when it
was merged, is itself part of the audit trail, complementing
the commit history with a record of the *review* of each
change.

The last piece of the remote's vocabulary is the pair
**tags and releases**, and they matter to reproducibility
specifically. A tag is a permanent, human-readable name
for one commit: `git tag -a v1.0-submission -m "State at
journal submission"` marks, forever and unambiguously, the
exact state of the analysis at the moment a manuscript was
submitted. A *release* on GitHub or GitLab is a tag dressed
with notes and downloadable archives. The reproducibility
payoff is that a reviewer, or the analyst a year later, can
recover the precise state that produced the submitted
numbers by naming the tag, rather than guessing which
commit was current on the submission date. In @sec-archiving
we shall see the same tagged state deposited to a public
archive with a citable identifier; the tag is the local
anchor to which that archival identity attaches.

::: {.callout-note icon=false collapse=true}
## Check your understanding: what does the pull request add?

A single analyst, working alone, keeps a clean commit
history with careful messages. A reviewer asks why, if the
history is already an audit trail, a multi-site trial
should bother with the additional machinery of forks and
pull requests. What does the pull request record that the
commit history does not?

::: {.callout-note icon=false collapse=true}
## Answer

The commit history records what changed and, in its
messages, why; it does not record who *reviewed* the change
before it entered the shared analysis, or what they said.
In a solo project there is no reviewer, so nothing is lost.
In a multi-site trial, where a change to the shared
analysis by one center affects every center's results, the
review is itself a controlled step, and the pull request is
the artifact that records it: the proposed diff, the
discussion, the requested changes, the approval, and the
merge, each attributable and timestamped. The commit
history answers 'what was changed and why'; the pull
request answers 'who checked it, and did they agree'. For a
regulated analysis both questions have auditors, and the
two records together are stronger than either alone.
:::
:::

## What git does not do

Having made version control the foundation, we must be
exact about the boundary of what it provides, because the
whole discipline of this book is matching a claim to what
has actually been secured. Git versions *text and tracks
its history*. It records, faithfully and tamper-evidently,
every state of the files under it and the reason for each
change. That is the entirety of what it does, and it is
the L0 rung of the ladder: an analysis under version
control is locatable, findable and readable by a second
person, and no more.

In particular, git does not pin the *package versions* an
analysis uses. A repository cloned a year later contains
the same `analysis.R`, but running it installs whatever
package versions are current, which may not be the versions
the analyst used, so the code that ran may no longer run,
or may run and return a different number. Git also does not
pin the *computational environment*: the R version, the
operating system, the system libraries a package links
against are nowhere in the history, and a clone is at the
mercy of whatever the next machine provides. These are
exactly the layers of the determinant stack that
@sec-levels enumerated as unpinned at L0, and closing them
is the work of the two chapters that follow: the lockfile
of @sec-dependencies pins the package layer to reach L1,
and the container of @sec-containers pins the environment
layer to reach L2. Version control is necessary for all of
this, since the lockfile and the Dockerfile are themselves
files whose history git tracks, and sufficient for none of
it. It puts the analysis on the ladder; it does not climb
it.

::: {.callout-important}
## The honest level

Placing an analysis under version control and pushing it
to GitHub achieves exactly L0, locatable, and a claim
should say no more. A repository is not 'reproducible'
because it is on GitHub; it is *findable*, which is the
necessary first rung and not the summit. A methods
section that writes 'code is available at github.com/...'
has stated an L0 fact accurately, and should not dress it
as an L1 or L2 claim. The honest sentence names the rung:
'The analysis code is version-controlled and available at
[URL]; package versions are pinned with renv (L1) and the
environment with a container (L2)' claims each rung only
if the corresponding artifact is actually present. Version
control is where the honesty of a reproducibility claim
begins, because it is the rung every higher claim is built
upon.
:::

## Worked example

To make the chapter concrete, we take the
incident-diabetes cohort of the running example from an
untracked folder to a tagged, remotely hosted repository,
staying at the level of the commands and the judgment
behind them. The shell sessions below are illustrative and
are not executed.

The analyst begins with a folder of scripts and a registry
extract, and the very first act, before a single commit,
is to write the `.gitignore`, because the extract contains
protected health information that must never enter the
history.

```bash
cd diabetes-cohort
git init
# write .gitignore FIRST, ignoring raw_data and credentials
git status   # confirm the extract is NOT listed as trackable
```

Only once `git status` confirms that the sensitive extract
is ignored does the analyst make the first commit, the one
that lifts the folder to L0, locatable.

```bash
git add .gitignore analysis/scripts/ DESCRIPTION
git commit -m "Initialise compendium and .gitignore"
```

Work proceeds as a sequence of small, message-bearing
commits: the cleaning script, the derived cohort table, the
Cox model, each its own commit with its own reason, exactly
the history tabulated in @tbl-audit-trail. When a reviewer
later asks why the outlier threshold changed, the answer is
one `git log` away.

To try a reviewer's proposed reanalysis without disturbing
the results already reported, the analyst branches.

```bash
git switch -c reviewer-sensitivity
# ... reanalysis commits ...
git switch main            # main is untouched and still runs
```

The work is backed up and made locatable by connecting a
remote and pushing.

```bash
git remote add origin https://github.com/lab/diabetes-cohort.git
git push -u origin main
```

A collaborator at a second site forks the repository, adds
a site-specific adjustment on a branch of their fork, and
opens a pull request. The analyst reviews the diff inline,
requests one change, and merges it, leaving a reviewable
record of the exchange. Finally, at journal submission, the
analyst marks the exact state.

```bash
git tag -a v1.0-submission -m "State at journal submission"
git push origin v1.0-submission
```

The folder is now a version-controlled, remotely hosted,
tagged repository: firmly at L0, with a clean audit trail,
a reviewed collaboration record, and a permanent anchor to
the submitted state. It is *not* yet L1 or L2, because
nothing has pinned the packages or the environment, and the
honest reproducibility statement says exactly that. The
lockfile of @sec-dependencies is the next commit.

## Collaborating with an LLM

A language model is a fluent assistant for git, which is a
domain rich in memorized command recipes, and precisely
that fluency is the hazard: the model will confidently
produce a command that does what was asked and more,
including irreversible operations, and it cannot see the
sensitivity of your files. The three triples below mark the
division of labor.

**Prompt.** 'I accidentally committed a file of patient
data three commits ago. Give me the git commands to remove
it.'

*Watch for.* The model will supply a history-rewriting
recipe (`git filter-repo`, or the older `filter-branch`, or
a `reset`) and present it as a clean fix. It cannot know
whether the offending commit was ever pushed to a remote,
which is the fact that determines whether removal is even
possible, and it will not, on its own, raise the breach
question.

*Verification.* Treat the exposure, not the git command, as
the primary problem. Establish first whether the commit was
pushed; if it was, assume the data has been cloned and
cached and follow your institution's incident procedure,
because no git command can recall it. Only then, and for a
purely local history, apply a rewriting tool, verifying on a
copy of the repository first. The model's recipe addresses
the mechanics and is blind to the obligation.

**Prompt.** 'Write a `.gitignore` for my R health-data
analysis project.'

*Watch for.* A generic R `.gitignore` (the standard
`.Rhistory`, `.RData`, `.Rproj.user/` list) that omits the
one category that matters most, the raw data and credentials
specific to *your* project, because the model cannot see
your directory and does not know where your PHI lives.

*Verification.* Take the generic list as a starting point,
then add, by hand, the explicit paths of every raw-data
directory, every `.Renviron` or credential file, and every
restricted dataset, and confirm with `git status` that each
is untracked before the first commit. The generic patterns
are safe to accept; the project-specific sensitive paths are
yours to supply.

**Prompt.** 'Write good commit messages for these changes.'

*Watch for.* Messages that fluently describe *what* changed,
which the model can read from the diff, but invent or omit
the *why*, which is the scientifically load-bearing part and
which the model cannot know. A plausible-sounding 'Update
outlier threshold' hides the reviewer request that a real
audit trail must record.

*Verification.* Supply the reason yourself. Let the model
draft the imperative summary line from the diff, then edit
in the motivation, the reviewer comment, the corrected
denominator, the protocol amendment, that turns a
description into an accountable record. The model can
summarize the change; only the analyst knows why it was
made.

## Exercises

1. Take a folder of your own currently versioned by the
   `FINAL`-file method, count the manual snapshot files it
   contains, and for two of them state what changed between
   them and why. Note which of those two questions you can
   still answer, and use the difficulty as a measure of what
   a commit history would have preserved.

2. Initialize a git repository in a scratch directory, make
   three commits that each change one file, and then use
   `git log --oneline` and `git show` to recover, for the
   middle commit, both what changed and the message you
   wrote. Confirm that the commit hashes are unchanged by a
   later commit.

3. Write a `.gitignore` for a data-analysis compendium and a
   second for an R-package compendium, and list the patterns
   that appear in one but not the other. Explain each
   difference in terms of what the archetype produces.

4. Explain, to a colleague who has just deleted a
   PHI-containing file in a new commit and believes the
   problem solved, why the file is still in the history, what
   determines whether it can be removed, and what to do if
   the offending commit was already pushed to GitHub.

5. Construct the sequence of commands by which a collaborator
   at a second trial site would contribute a change through a
   fork and a pull request, and state, for each step, what
   part of the audit trail it records that a direct push to
   `main` would not.

6. An analyst pushes a version-controlled compendium to
   GitHub and writes in the methods section that the analysis
   is 'fully reproducible'. State the rung the analysis has
   actually reached, name the two layers of the determinant
   stack that remain unpinned, and rewrite the sentence so
   that it claims exactly what version control secures and no
   more.

## Solutions

Attempt an exercise before expanding its solution. The first
two ask the reader to work in their own directories, so what
they produce will vary; the boxes say what a good answer
establishes.

::: {.callout-note icon=false collapse=true}
## Exercise 1

No single correct answer. What a good answer establishes is
the asymmetry between the two questions, because the
asymmetry is the argument for version control.

Almost every reader can answer *what changed* between two
snapshot files, because the files are still there and a
diff tool will show it. Almost no reader can answer *why*,
because the reason was never written down anywhere. It
lived in the analyst's head, or in an email, or in a
conversation after a meeting, and it is gone. Readers
commonly find they can reconstruct the what for files a
year old and cannot reconstruct the why for files a month
old.

That gap is precisely what a commit history preserves.
The diff is recoverable from the files themselves, so it
is the cheaper half; the message is not recoverable from
anything, so it is the half that decays. This is why the
chapter insists that a commit is an atomic unit of one
idea with a message attached: the message is the only part
of the record that cannot be reconstructed later.

A reader who finds that they *can* answer why, because
they named the files carefully, has demonstrated the
`FINAL` method working at its best, and should notice how
much of the reasoning fits in a filename.
:::

::: {.callout-note icon=false collapse=true}
## Exercise 2

The mechanics are straightforward, so the substance is the
last clause.

```bash
git init
# edit file_a.txt
git add file_a.txt && git commit -m 'Add the cohort filter'
# edit file_b.txt
git add file_b.txt && git commit -m 'Raise the outlier threshold'
# edit file_c.txt
git add file_c.txt && git commit -m 'Add the summary table'
git log --oneline
git show <middle-hash>
```

`git log --oneline` lists the three commits newest first,
each with an abbreviated hash and its subject line.
`git show` on the middle hash prints both halves of the
record: the message you wrote, and the diff of what
changed.

On the hashes: recording the middle commit's hash before
making the third commit and comparing afterward shows it
unchanged. This is worth dwelling on. A commit hash is
computed over the commit's content *and* its parent's
hash, so it fixes the entire history leading to it. Adding
a commit on top cannot alter it, because the new commit
points backward and nothing points forward. That is the
property that makes the history an audit trail rather than
a mutable log.

The honest qualification, which a careful reader may
raise: hashes are stable under *adding* history, not under
*rewriting* it. An `amend` or a `rebase` produces new
commits with new hashes, and the old ones become
unreferenced. The audit-trail property holds for a history
that is appended to, which is why rewriting shared history
is treated as a serious act.
:::

::: {.callout-note icon=false collapse=true}
## Exercise 3

The two files share most of their content and differ in a
way that follows from what each archetype produces.

Common to both: `.Rproj.user/`, `.Rhistory`, `.RData`,
`.DS_Store`, and `renv/library/`, which is a local install
tree reconstructable from the lockfile and must never be
committed.

In the **data-analysis** file but not the package one:
patterns for rendered outputs, `*.html` and `*.pdf`, since
the report is a product the code regenerates; the derived
data directory, or selected paths within it; and, most
importantly, the raw-data extract if it carries protected
information.

In the **R-package** file but not the analysis one:
`inst/doc/`, which holds built vignettes, and `src/*.o`
and `src/*.so`, the compiled object files, which exist only
if the package carries compiled code and are rebuilt on
every install.

The explanation to give for each difference is the same
shape: an archetype ignores what it *generates* and commits
what it *is*. A data-analysis compendium generates a
report and so ignores it. A package generates compiled
objects and built vignettes and so ignores those. Neither
ignores its own source.

The one line that is not a matter of archetype is the data
carrying protected health information, which is ignored in
every archetype, always, and is the subject of the next
exercise.
:::

::: {.callout-note icon=false collapse=true}
## Exercise 4

The colleague's mental model is that a repository holds the
current state of the files. It holds their whole history,
and that is the point of it.

**Why the file is still there.** Deleting a file in a new
commit records the deletion as a change; the commit that
added the file is untouched, and the file's contents are
retrievable from it by anyone who clones the repository.
The deletion is a new fact appended to the record, not an
erasure of the old one.

**What determines whether it can be removed.** Whether the
history has been shared. Removing content means rewriting
every commit from the offending one forward, with
`git filter-repo` or an equivalent, which produces new
hashes for all of them. If the offending commit exists only
in the local repository, this is a local operation and
costs nothing but care. If it has been pushed, every clone
and fork holds the old objects, and rewriting the local
copy does not touch them.

**What to do if it was already pushed to GitHub.** Treat
the data as disclosed and act in that order. Notify whoever
is responsible for the data, an IRB or a privacy office,
because a disclosure has occurred and the notification
obligation is not a technical question. Rotate anything
rotatable. Then rewrite the history and force-push, ask
GitHub Support to purge cached views and any dangling
objects, and require every collaborator to re-clone rather
than pull, since a pull onto a rewritten history
reintroduces the old objects. Note the ordering: the
cleanup is the last step and the least important one,
because it does not undo the exposure.

The lesson the chapter draws is that this is why the
`.gitignore` for protected data is written before the first
commit rather than after the first mistake.
:::

::: {.callout-note icon=false collapse=true}
## Exercise 5

A serviceable sequence, with the audit contribution of each
step.

```bash
# 1. Fork the canonical repository on the forge.
git clone https://github.com/<site2>/<study>.git
cd <study>
git remote add upstream https://github.com/<coord>/<study>.git

# 2. Branch from an up-to-date main.
git fetch upstream && git checkout -b fix/outlier-threshold upstream/main

# 3. Commit the change.
git add R/clean_cohort.R
git commit -m 'Raise outlier threshold to 4 SD per DSMB request'

# 4. Push to the fork and open a pull request.
git push origin fix/outlier-threshold
```

What each step records that a direct push would not. The
**fork** establishes that the contributor did not hold
write access to the canonical analysis, so the change could
not enter without someone else's action. The **branch**
separates the proposal from the accepted state, so the
history of `main` contains only what was agreed. The
**commit** is the same in either workflow and records what
and why. The **pull request** is the artifact that carries
everything a direct push destroys: the proposed diff before
it was accepted, the discussion, any requested changes and
the revisions answering them, the identity of the approver,
and the timestamp of the approval. The **merge** records
that the approval preceded the entry into `main`.

The compressed version of the answer: the commit history
answers 'what changed and why'; the pull request answers
'who checked it, and did they agree'. A direct push to
`main` answers only the first, and in a multi-site trial
the second question has auditors too.
:::

::: {.callout-note icon=false collapse=true}
## Exercise 6

**The rung reached is L0.** Version control makes the
analysis locatable and gives it a history. It pins the
analysis code, which is a real achievement and the
precondition for everything above it, but the ladder grades
how far *down* the determinant stack a result is locked,
and the code sits at the top of that stack.

**Two layers that remain unpinned.** The package versions,
since nothing in a git repository records which version of
`dplyr` produced the numbers; and the environment beneath
them, the R runtime, the operating system, and the system
libraries. A reader may equally name the numerical
libraries or the session configuration and be right; the
point is to name layers the repository does not touch. The
common wrong answer is to name the data, which version
control *can* pin for a small committed dataset, though
usually does not for a large or protected one.

**A rewritten sentence.** 'The analysis code is version
controlled and deposited at [location], so the exact code
that produced these results can be retrieved. The software
environment was not recorded; re-execution may therefore
require package versions differing from those used here,
and we have not verified that the results regenerate.'

The three clauses map onto the three claims: what is
secured, what is not, and what has not been checked. The
last clause is the one authors most often omit, and it is
the difference between claiming L0 honestly and claiming
L3 by implication.
:::

## Further reading

- @ram2013 is the direct argument for git as an instrument
  of reproducibility and transparency in science, and is the
  natural first reading for a researcher who has met git only
  as a software tool; it makes the case for the commit
  history as a record of scientific provenance.
- @bryan2018 is the standard orientation to version control
  written for data scientists rather than software engineers,
  and it motivates GitHub as the collaboration substrate with
  the working analyst's concerns foremost; her longer online
  work *Happy Git and GitHub for the useR* is the companion
  practical manual.
- @wilson2017 and @sandve2013 place version control among the
  good practices of reproducible computational research, the
  former as one of a set of habits for scientific computing
  and the latter among its rules for reproducible analysis;
  read them to see version control situated alongside the
  other pillars this book develops.
- The layers that version control does not pin are the
  subject of @sec-dependencies and @sec-containers, and the
  continuous integration and collaboration built upon a git
  history are developed in @sec-verification and
  @sec-archiving; this chapter is best read as the foundation
  those four presuppose.
